---
title: "deseq2_main_analysis"
author: "Kira Goff"
date: "2026-05-25"
output: html_document
---

# 1. Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

```{r load-libraries}
library(DESeq2)

library(dplyr)
library(tibble)
library(readr)
library(here)
library(ggplot2)
library(apeglm)

library(pheatmap)
library(ComplexUpset)
library(EnhancedVolcano)
library(pcaExplorer)
```


```{r define-directories}
base_dir <- "./"
import_dir <- "./kallisto"
results_dir <- "./results"
```



# 2. Import Data

NOTE: Ideally, use tximport to import your transcript-level abundances.
The system I'm working on has long-standing libcairo issues that impact ragg and R packages that use it as a downstream dependency.

```{r import-kallisto}

colData <- read.csv(file.path(base_dir,"meta.csv"), header = TRUE, sep = ",")

# Get list of subfolders
sample_folders <- list.dirs(import_dir, full.names = TRUE, recursive = FALSE)
count_list <- list()

# Loop over data
for (folder in sample_folders) {
  sample_name <- basename(folder)
  tsv_file <- file.path(folder, "abundance.tsv")
  df <- read.delim(tsv_file, stringsAsFactors = FALSE)
  df <- df %>% select(target_id, est_counts)

  # est_counts to integers so DESeq2 doesn't throw a hissy
  df$est_counts <- round(df$est_counts)

  colnames(df)[2] <- sample_name
  count_list[[sample_name]] <- df
}

# Merge all 24 dataframes by target_id
countData <- Reduce(function(x, y) merge(x, y, by = "target_id"), count_list)
countData <- countData %>% select(target_id, everything())

write.table(countData, file.path(results_dir,"countDataAll.tsv"), sep = "\t", row.names = FALSE, quote = FALSE)

# Keep only rows where target_id starts with "P"
countData <- countData[grepl("^P", countData$target_id), ]

write.table(countData, file.path(results_dir,"countDataPeri.tsv"), sep = "\t", row.names = FALSE, quote = FALSE)

# Set target_id as row names
rownames(countData) <- countData$target_id
countData$target_id <- NULL

head(countData)
```


```{r import-annotations}
# Combination of annotations was done using my annotation-integration.Rmd script

peri_anno <- read_tsv(file.path(base_dir,"peri-anno-definitions.tsv"))

# trim to what you want
anno_operon <- peri_anno[, c(1, 2, 9, 10, 11, 14, 15, 36, 12, 13, 27, 28, 25, 26, 29, 30, 20)]
```



# 3. Global Modeling

## 3.1 Process Data

```{r make-peri-object}
dds <- DESeqDataSetFromMatrix(
  countData = countData,
  colData = colData,
  design = ~ time + exposure + time:exposure
)
```


```{r global-models}
dds_full <- dds
design(dds_full) <- ~ time + exposure + time:exposure
dds_full <- DESeq(dds_full)

dds_full_lrt <- DESeq(
  dds,
  test = "LRT",
  reduced = ~ time + exposure
)
```


```{r variance-stabilization-full}
vsd_full <- vst(dds_full, blind = FALSE)
```


## 3.2 QC and Visualization

```{r pca-explorer-full}
pcaExplorer(dds_full)
```

If you'd rather not use pcaExplorer:
```{r pca-full}
plotPCA(vsd_full)
```

```{r disp-est-full}
plotDispEsts(dds_full)
```

```{r disp-est-vst-full}
plotDispEsts(vsd_full)
```

```{r}
sample_dists <- dist(t(assay(vsd_full)))

pheatmap(
  as.matrix(sample_dists),
  annotation_col = as.data.frame(
    colData(dds_full)[, c("time", "exposure")]
  )
)
```


```{r heatmap}
# Significant genes from LRT
res_full_lrt <- results(dds_full_lrt)

sig_genes <- rownames(
  subset(res_full_lrt, padj < 0.05)
)

vsd_sig <- vsd_full[sig_genes, ]

mat <- assay(vsd_sig)

# Row scaling
mat <- t(scale(t(mat)))

pheatmap(
  mat,
  show_rownames = FALSE,
  cluster_cols = TRUE,
  annotation_col = as.data.frame(
    colData(dds_full)[, c("time", "exposure")]
  )
)
```

# 4. Targeted Analysis of 06h and 24h

In this case, the gene expression at 72 hours is starkly different than at 06h or 24h.

At 72h:
- much of the CHCA has been degraded
- the Peribacillus is making endospores
- the % of the reads coming from Peribacillus is ~20%


## 4.1 Process Data

```{r make-object-0624}

dds_06_24 <- dds[, dds$time != "72h"]
colData(dds_06_24) <- droplevels(colData(dds_06_24))
design(dds_06_24) <- ~ time + exposure + time:exposure
dds_06_24 <- DESeq(dds_06_24)
```


```{r variance-stabilization-0624}
vsd_06_24 <- vst(dds_06_24, blind = FALSE)
```


## 4.2 QC and Visualization

```{r pca-explorer-0624}
pcaExplorer(dds_06_24)
```

If you'd rather not use pcaExplorer:
```{r pca-Exp-vsd-0624}
plotPCA(vsd_06_24)
```

```{r disp-est-0624}
plotDispEsts(dds_06_24)
```


## 4.3 Extract Results

```{r get-contrasts}
# Time
res_time <- lfcShrink(
  dds_06_24,
  coef = "time_24h_vs_06h",
  type = "apeglm"
)
# Exposure
res_exposure <- lfcShrink(
  dds_06_24,
  coef = "exposure_Exposed_vs_Control",
  type = "apeglm"
)

# Time: Exposure Interaction
res_interaction <- lfcShrink(
  dds_06_24,
  coef = "time24h.exposureExposed",
  type = "apeglm"
)

```

```{r sig-results-0624}
sig_time <- rownames(res_time)[which(res_time$padj < 0.05)]

sig_exposure <- rownames(res_exposure)[which(res_exposure$padj < 0.05)]

sig_interaction <- rownames(res_interaction)[which(res_interaction$padj < 0.05)]
```


```{r summarize-contrasts}
get_sig_count <- function(res, alpha = 0.05) {
  sum(res$padj < alpha, na.rm = TRUE)
}

summary_tbl <- tibble(
  model_term = c("time", "exposure", "interaction"),
  significant_genes = c(
    get_sig_count(res_time),
    get_sig_count(res_exposure),
    get_sig_count(res_interaction)
  )
)
```


```{r export-results-0624}
write_csv(
  as.data.frame(res_time) %>%
    rownames_to_column("gene_id"),
  file.path(results_dir, "res_time.csv")
)

write_csv(
  as.data.frame(res_exposure) %>%
    rownames_to_column("gene_id"),
  file.path(results_dir, "res_exposure.csv")
)

write_csv(
  as.data.frame(res_interaction) %>%
    rownames_to_column("gene_id"),
  file.path(results_dir, "res_interaction.csv")
)
```


## 4.4 Visualize Results

```{r volcano-plots-0624}
alpha <- 0.05
lfc_cutoff <- 1
contrast_set <- res_interaction

contrast_name <- deparse(substitute(contrast_set))
save <- sprintf("volcano-%s-qval_%g-lfc_%g.png", contrast_name, alpha, lfc_cutoff)

ev <- EnhancedVolcano(
  contrast_set,
  lab = rownames(contrast_set),
  x = "log2FoldChange",
  y = "padj",
  pCutoff = alpha,
  FCcutoff = lfc_cutoff,
  selectLab = NULL
)
ggsave (file.path(results_dir,save))


```


```{r volcano-0624}
# variables
alpha <- 0.05
lfc_cutoff <- 1
contrast_set <- res_exposure
contrast_name <- "res_exposure"
headline <- "06h and 24h Model"
subhead <- "Contrast - Exposure"
#

save <- sprintf("volcano-%s-qval_%g-lfc_%g.png", contrast_name, alpha, lfc_cutoff)

ev <- EnhancedVolcano(
  contrast_df,
  lab = rep("", nrow(contrast_df)),
  x = "log2FoldChange",
  y = "padj",
  pCutoff = alpha,
  FCcutoff = lfc_cutoff,
  # for whatever reason, Enhanced Volcano is using p unless you do a manual overwrite of legend labels
  legendLabels = c(
    "NS",
    "Log2 FC",
    "q-value",
    "q-value & Log2FC"
  ),
  ylab = "-log10 q",
  xlab = "Log2FC",
  title = headline,
  subtitle = subhead
)

ggsave(filename = file.path(results_dir, save))
```




```{r ma-0624}
plotMA(
  res_time,
  ylim = c(-5, 5),
  main = "Time Effect"
)

plotMA(
  res_exposure,
  ylim = c(-5, 5),
  main = "Exposure Effect"
)

plotMA(
  res_interaction,
  ylim = c(-5, 5),
  main = "Interaction Effect"
)
```

## 4.5 Visualize Contrast Effects

```{r summarize-contrasts}
all_genes <- rownames(res_time)
padj_df <- data.frame(gene = rownames(res_time)) %>%
  mutate(
    padj_time          = res_time$padj,
    log2FC_time        = res_time$log2FoldChange,
    padj_exposure      = res_exposure$padj,
    log2FC_exposure    = res_exposure$log2FoldChange,
    padj_interaction   = res_interaction$padj,
    log2FC_interaction = res_interaction$log2FoldChange
  )

padj_df <- padj_df %>%
  rowwise() %>%
  mutate(
    best_term = {
      vals <- c(time = padj_time, exposure = padj_exposure, interaction = padj_interaction)
      vals <- vals[!is.na(vals)]
      if (length(vals) == 0) NA_character_ else names(vals)[which.min(vals)]
    }
  ) %>%
  ungroup()

write_tsv(padj_df, file.path(results_dir,"fullmodel_06h24h_byterm.tsv"))
```

```{r best-model-term}
term_counts <- padj_df %>%
  filter(!is.na(best_term)) %>% # ignore remaining NAs
  group_by(best_term) %>%
  summarise(count = n())

# Plot
ggplot(term_counts, aes(x = best_term, y = count, fill = best_term)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(
    title = "Distribution of Genes by Most Significant Term",
    x = "Best Term",
    y = "Number of Genes"
  ) +
  scale_fill_brewer(palette = "Set2")

ggsave(file.path(results_dir,"bargraph-terms.png"))
```

```{r upset-sig-terms}
gene_lists <- list(
  Time = sig_time,
  Exposure = sig_exposure,
  Interaction = sig_interaction
)

# Create a binary matrix
upset_df <- tibble(gene = all_genes) %>%
  mutate(
    Time = gene %in% sig_time,
    Exposure = gene %in% sig_exposure,
    Interaction = gene %in% sig_interaction
  )


upset(upset_df, intersect = c("Time", "Exposure", "Interaction"))

ggsave(file.path(results_dir,"upset-terms.png"))
```

# 5. 06h Control/Exposure Comparison

```{r make-06h}
dds_06h <- dds[, dds$time == "06h"]
# Drop unused factor levels
colData(dds_06h) <- droplevels(
  colData(dds_06h)
)
```


```{r model-06h}
design(dds_06h) <- ~ exposure

dds_06h <- DESeq(dds_06h)

res_06h_exposure <- lfcShrink(
  dds_06h,
  coef = "exposure_Exposed_vs_Control",
  type = "apeglm"
)

write_csv(
  as.data.frame(res_06h_exposure) %>%
    rownames_to_column("gene_id"),
  file.path(results_dir, "res_06h_exposure.csv")
)

```



```{r variance-stabilization-0624}
vsd_06h <- vst(dds_06h, blind = FALSE)
```


```{r pca-explorer-06}
pcaExplorer(dds_06h)
```

If you'd rather not use pcaExplorer:
```{r pca-Exp-vsd-06}
plotPCA(vsd_06h)
```


```{r volcano-06}
# variables
alpha <- 0.05
lfc_cutoff <- 1
contrast_set <- res_06h_exposure
contrast_name <- "res_06h_exposure"
headline <- "06h Control:Exposure"
#

save <- sprintf("volcano-%s-qval_%g-lfc_%g.png", contrast_name, alpha, lfc_cutoff)

ev <- EnhancedVolcano(
  contrast_df,
  lab = rep("", nrow(contrast_df)),
  x = "log2FoldChange",
  y = "padj",
  pCutoff = alpha,
  FCcutoff = lfc_cutoff,
  # for whatever reason, Enhanced Volcano is using p unless you do a manual overwrite of legend labels
  legendLabels = c(
    "NS",
    "Log2 FC",
    "q-value",
    "q-value & Log2FC"
  ),
  ylab = "-log10 q",
  xlab = "Log2FC",
  title = headline,
  subtitle=""
)

ev

ggsave(filename = file.path(results_dir, save))
```


# 6. 24h Control/Exposure Comparison

```{r make-24h}
dds_24h <- dds[, dds$time == "24h"]
# Drop unused factor levels
colData(dds_24h) <- droplevels(
  colData(dds_24h)
)
```


```{r model-24h}
design(dds_24h) <- ~ exposure

dds_24h <- DESeq(dds_24h)

res_24h_exposure <- lfcShrink(
  dds_24h,
  coef = "exposure_Exposed_vs_Control",
  type = "apeglm"
)

write_csv(
  as.data.frame(res_24h_exposure) %>%
    rownames_to_column("gene_id"),
  file.path(results_dir, "res_24h_exposure.csv")
)

```


```{r variance-stabilization-2424}
vsd_24h <- vst(dds_24h, blind = FALSE)
```


```{r pca-explorer-24}
pcaExplorer(dds_24h)
```

If you'd rather not use pcaExplorer:
```{r pca-Exp-vsd-24}
plotPCA(vsd_24h)
```




```{r volcano-24}
# variables
alpha <- 0.05
lfc_cutoff <- 1
contrast_set <- res_24h_exposure
contrast_name <- "res_24h_exposure"
headline <- "24h Control:Exposure"
#

save <- sprintf("volcano-%s-qval_%g-lfc_%g.png", contrast_name, alpha, lfc_cutoff)

ev <- EnhancedVolcano(
  contrast_df,
  lab = rep("", nrow(contrast_df)),
  x = "log2FoldChange",
  y = "padj",
  pCutoff = alpha,
  FCcutoff = lfc_cutoff,
  # for whatever reason, Enhanced Volcano is using p unless you do a manual overwrite of legend labels
  legendLabels = c(
    "NS",
    "Log2 FC",
    "q-value",
    "q-value & Log2FC"
  ),
  ylab = "-log10 q",
  xlab = "Log2FC",
  title = headline,
  subtitle=""
)

ev

ggsave(filename = file.path(results_dir, save))
```



```{r}
sessionInfo()
```
