---
title: "2 - cluster analysis and over representation"
author: "Kira Goff"
date: "2026-05-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(DESeq2)
library(ggplot2)
```

use the contrasts to create a three dimensional vector

```{r}
df <- tibble(
  gene = rownames(res_time),

  time_effect        = res_time$log2FoldChange,
  exposure_effect    = res_exposure$log2FoldChange,
  interaction_effect = res_interaction$log2FoldChange
)

```

Model space for clustering
```{r}

mat <- df %>%
  select(time_effect, exposure_effect, interaction_effect) %>%
  as.matrix()

rownames(mat) <- df$gene
mat_scaled <- scale(mat)
```

OR: at 24h

```{r}
df24 <- df %>%
  mutate(
    exposure_06h = exposure_effect,
    exposure_24h = exposure_effect + interaction_effect
  )
mat24 <- df24 %>%
  select(time_effect, exposure_06h, exposure_24h) %>%
  as.matrix()

rownames(mat24) <- df24$gene
mat_scaled_24 <- scale(mat24)
```

```{r}
ggplot(df24, aes(exposure_06h, exposure_24h)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, linetype = "dashed") +
  theme_minimal() +
  labs(
    x = "Exposure effect (06h)",
    y = "Exposure effect (24h)",
    title = "Time modulation of exposure response"
  )
```

```{r}
ggplot(df24, aes(exposure_06h, exposure_24h)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, linetype = "dashed") +
  theme_minimal() +
  labs(
    x = "Exposure effect (06h)",
    y = "Exposure effect (24h)",
    title = "Time modulation of exposure response"
  )
```


```{r}
mat <- mat_scaled_24

# convert to matrix (safe)
mat <- as.matrix(mat)

# remove non-finite rows
keep <- apply(mat, 1, function(x) all(is.finite(x)))

mat_clean <- mat[keep, ]

mat_scaled_24_clean <- scale(mat_clean)

library(cluster)

sil_width <- numeric(10)

for (k in 2:10) {
  km <- kmeans(mat_scaled_24_clean, centers = k, nstart = 25)

  ss <- silhouette(km$cluster, dist(mat_scaled_24_clean))

  sil_width[k] <- mean(ss[, 3])
}

plot(2:10, sil_width[2:10], type = "b")
```

```{r}

library(cluster)

sil_width <- c()

for (k in 2:10) {
  km <- kmeans(mat_scaled_24_clean, centers = k, nstart = 25)
  ss <- silhouette(km$cluster, dist(mat_scaled_24_clean))
  sil_width[k] <- mean(ss[, 3])
}

plot(2:10, sil_width[2:10], type = "b")
```

```{r}
results <- list()

for (k in 2:10) {

  model <- cmeans(mat_scaled_24_clean, centers = k, m = 1.25)

  cluster <- max.col(model$membership)

  sil <- silhouette(cluster, dist(mat_scaled_24_clean))

  results[[k]] <- list(
    silhouette = mean(sil[, 3]),
    entropy = mean(-rowSums(model$membership * log(model$membership)))
  )
}
```

```{r}
sil_vals <- sapply(results, `[[`, "silhouette")
ent_vals <- sapply(results, `[[`, "entropy")

plot(2:10, sil_vals[2:10], type="b", col="blue",
     ylab="Score", xlab="k")
lines(2:10, ent_vals[2:10], col="red")
legend("bottomright", c("silhouette", "entropy"),
       col=c("blue", "red"), lty=1)
```

# using log2fc at 06 and 24 as our space, instead of the vectors


```{r}
df_exp <- tibble(
  gene = rownames(res_06h_exposure),

  log2FC_06h = res_06h_exposure$log2FoldChange,
  log2FC_24h = res_24h_exposure$log2FoldChange
)
```

```{r}
ggplot(df_exp, aes(log2FC_06h, log2FC_24h)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, linetype = "dashed") +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  theme_minimal() +
  labs(
    x = "Exposure effect (06h)",
    y = "Exposure effect (24h)",
    title = "Exposure response modulation over time"
  )
```

```{r}
library(e1071)

mat_exp <- df_exp %>%
  select(log2FC_06h, log2FC_24h) %>%
  as.matrix()

rownames(mat_exp) <- df_exp$gene

mat_exp <- scale(mat_exp)
```

```{r}
library(e1071)
library(cluster)

k_vals <- 2:10

sil_vals <- numeric(length(k_vals))
fuzz_vals <- numeric(length(k_vals))  # fuzziness / entropy
```

```{r}
for (i in seq_along(k_vals)) {

  k <- k_vals[i]

  model <- cmeans(mat_scaled_24_clean, centers = k, m = 1.25)

  # hard assignment (for silhouette)
  cl <- model$cluster

  ss <- silhouette(cl, dist(mat_scaled_24_clean))
  sil_vals[i] <- mean(ss[, 3], na.rm = TRUE)

  # fuzziness (entropy of membership)
  fuzz_vals[i] <- mean(
    -rowSums(model$membership * log(model$membership)),
    na.rm = TRUE
  )
}
```

```{r}
plot(k_vals, sil_vals,
     type = "b",
     col = "blue",
     ylab = "Score",
     xlab = "k",
     main = "Cluster validity")

lines(k_vals, fuzz_vals, type = "b", col = "red")

legend("topright",
       legend = c("Silhouette", "Fuzziness (entropy)"),
       col = c("blue", "red"),
       lty = 1)
```
non-vsd data 

1. create three dimensional matrix

```{r build-and-clean-matrix}
mat <- cbind(
  time       = res_time$log2FoldChange,
  exposure   = res_exposure$log2FoldChange,
  interaction = res_interaction$log2FoldChange
)

rownames(mat) <- rownames(res_time)

# Remove rows with NA
mat_clean <- na.omit(mat)

# Scale by row (gene-wise z-score)
mat_scaled <- t(scale(t(mat_clean)))
```

2. silhouette widths

```{r}
library(cluster)
library(factoextra)

# Compute average silhouette width for k = 2:10
fviz_nbclust(
  mat_scaled,
  kmeans,
  method = "silhouette",
  k.max = 10
)
```

4 or 6 is probably correct

```{r make-clusters}
set.seed(123)

km <- kmeans(mat_scaled, centers = 6)

clusters <- km$cluster
```

```{r}
clustered_mat <- data.frame(
  mat_scaled,
  cluster = factor(clusters)
)
```

```{r}
library(plotly)

pca <- prcomp(mat_scaled)

pca_df <- data.frame(
  PC1 = pca$x[,1],
  PC2 = pca$x[,2],
  PC3 = pca$x[,3],
  cluster = factor(clusters)
)

plot_ly(
  pca_df,
  x = ~PC1,
  y = ~PC2,
  z = ~PC3,
  color = ~cluster,
  type = "scatter3d",
  mode = "markers"
)
```


2. figure out clustering

3. assign clusters

4. project cluster scatterplot into three dimensional space



```{r}
library(pheatmap)

pheatmap(
  mat_scaled,
  cluster_rows = TRUE,
  show_rownames = FALSE
)
```
```{r}
matplot(
  t(mat_scaled[1:100,]),
  type = "l",
  lty = 1,
  xlab = "Condition",
  ylab = "Scaled log2FC"
)
```

## okay let's restart here
project the clusters onto the log2fc 06h vs log2fc 24

```{r}
plot_df <- merge(
  data.frame(
    gene = rownames(res_06h_exposure),
    log2FC_06h = res_06h_exposure$log2FoldChange
  ),
  
  data.frame(
    gene = rownames(res_24h_exposure),
    log2FC_24h = res_24h_exposure$log2FoldChange
  ),
  
  by = "gene"
)
```

```{r}
cluster_df <- data.frame(
  gene = rownames(mat_scaled),
  cluster = factor(clusters)
)

plot_df <- merge(plot_df, cluster_df, by = "gene")
```

```{r}
library(ggplot2)


ggplot(plot_df,
       aes(log2FC_06h,
           log2FC_24h,
           color = cluster)) +
  geom_point(alpha = 0.7) +
  theme_classic() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed")
```

##OR, using mfuzz for clusters

```{r}
library(Mfuzz)
library(Biobase)

# Convert to ExpressionSet
eset <- ExpressionSet(assayData = as.matrix(mat_scaled))

# Estimate fuzzifier
m <- mestimate(eset)

# Try different cluster numbers
Dmin(eset, m, crange = 2:10, repeats = 3)
```
```{r}
cl <- mfuzz(eset, c = 4, m = m)
```


```{r}
mfuzz.plot(
  eset,
  cl = cl,
  mfrow = c(2,2),
  time.labels = colnames(mat_scaled)
)
```

```{r}
cl4 <- mfuzz(eset, c = 4, m = m)

mfuzz.plot(eset,
           cl = cl4,
           mfrow = c(2,2),
           time.labels = c("06h", "24h", "interaction"))
```

```{r}
dev.new()

mfuzz.plot(
  eset,
  cl = cl4,
  mfrow = c(2,2),
  time.labels = c("06h", "24h", "interaction")
)
```

```{r}
library(tidyverse)

plot_long <- plot_df %>%
  pivot_longer(
    cols = c(log2FC_06h, log2FC_24h),
    names_to = "time",
    values_to = "log2FC"
  )

tt <-ggplot(plot_long,
       aes(time,
           log2FC,
           group = gene,
           color = cluster)) +
  geom_line(alpha = 0.1) +
  stat_summary(
    aes(group = cluster),
    fun = mean,
    geom = "line",
    linewidth = 2
  ) +
  facet_wrap(~cluster) +
  theme_classic()
tt
```


```{r}
library(Mfuzz)
library(tidyverse)

m <- mestimate(eset)

plot_list <- list()

for (k in 3:7) {

  set.seed(123)

  # fuzzy clustering
  cl <- mfuzz(eset, c = k, m = m)

  # extract hard cluster assignment (max membership)
  cluster_assign <- apply(cl$membership, 1, which.max)

  plot_df_k <- plot_df %>%
    mutate(cluster = factor(cluster_assign))

  plot_long <- plot_df_k %>%
    pivot_longer(
      cols = c(log2FC_06h, log2FC_24h),
      names_to = "time",
      values_to = "log2FC"
    )

  p <- ggplot(plot_long,
              aes(time,
                  log2FC,
                  group = gene,
                  color = cluster)) +
    geom_line(alpha = 0.1) +
    stat_summary(
      aes(group = cluster),
      fun = mean,
      geom = "line",
      linewidth = 2
    ) +
    facet_wrap(~cluster) +
    theme_classic() +
    ggtitle(paste("Mfuzz clustering k =", k))

  plot_list[[paste0("k_", k)]] <- p
}
```

```{r}
plot_list[["k_3"]]
plot_list[["k_4"]]
plot_list[["k_5"]]
```
```{r}
plot_list[["k_6"]]
plot_list[["k_7"]]
```
```{r}

calc_separation <- function(membership_matrix) {
  apply(membership_matrix, 1, function(x) {
    sort_x <- sort(x, decreasing = TRUE)
    sort_x[1] - sort_x[2]
  })
}
```

```{r}
library(Mfuzz)

m <- mestimate(eset)

sep_summary <- data.frame()

for (k in 2:10) {

  set.seed(123)
  cl <- mfuzz(eset, c = k, m = m)

  sep <- calc_separation(cl$membership)

  sep_summary <- rbind(
    sep_summary,
    data.frame(
      k = k,
      mean_separation = mean(sep, na.rm = TRUE),
      median_separation = median(sep, na.rm = TRUE),
      pct_low_sep = mean(sep < 0.2)
    )
  )
}
```

```{r}
sep_summary
```

```{r}
library(ggplot2)

ggplot(sep_summary,
       aes(x = k, y = mean_separation)) +
  geom_line() +
  geom_point(size = 3) +
  theme_classic() +
  ylab("Mean cluster separation") +
  xlab("Number of clusters (k)")
```


# on VSD
```{r build-and-clean-matrix}
mat <- cbind(
  time       = res_time$log2FoldChange,
  exposure   = res_exposure$log2FoldChange,
  interaction = res_interaction$log2FoldChange
)

rownames(mat) <- rownames(res_time)

# Remove rows with NA
mat_clean <- na.omit(mat)

# Scale by row (gene-wise z-score)
mat_scaled <- t(scale(t(mat_clean)))
```





##############

```{r vst-matrix}
mat_vst <- assay(vst(dds, blind = TRUE))

# optional: remove low-expression genes first
mat_vst <- mat_vst[rowMeans(mat_vst) > 5, ]

# remove zero-variance genes
vst_var <- apply(mat_vst, 1, var)
mat_vst <- mat_vst[vst_var > 0, ]

# final safety check
mat_vst <- mat_vst[complete.cases(mat_vst), ]
```


```{r lfc-matrix}
mat_lfc <- cbind(
  time = res_time$log2FoldChange,
  exposure = res_exposure$log2FoldChange,
  interaction = res_interaction$log2FoldChange
)

rownames(mat_lfc) <- rownames(res_time)

# remove rows with NA or infinite values
mat_lfc <- mat_lfc[is.finite(rowSums(mat_lfc)), ]
mat_lfc <- mat_lfc[complete.cases(mat_lfc), ]

# remove zero-variance genes
lfc_var <- apply(mat_lfc, 1, var)
mat_lfc <- mat_lfc[lfc_var > 0, ]
```

```{r shared-geneset}
common_genes <- intersect(rownames(mat_vst), rownames(mat_lfc))

mat_vst <- mat_vst[common_genes, ]
mat_lfc <- mat_lfc[common_genes, ]
```

```{r}
pca_vst <- prcomp(t(mat_vst), center = TRUE, scale. = FALSE)

plot_ly(
  x = pca_vst$x[,1],
  y = pca_vst$x[,2],
  z = pca_vst$x[,3],
  type = "scatter3d",
  mode = "markers"
) %>%
  layout(title = "PCA - VST space")

```


```{r}
pca_lfc <- prcomp(t(mat_lfc), center = TRUE, scale. = TRUE)

plot_ly(
  x = pca_lfc$x[,1],
  y = pca_lfc$x[,2],
  z = pca_lfc$x[,3],
  type = "scatter3d",
  mode = "markers"
) %>%
  layout(title = "PCA - LFC space")
```


```{r check-pca-loadings}
summary(pca_vst)
```

```{r}
summary(pca_lfc)
```

```{r}
library(Mfuzz)

m <- mestimate(eset)

k_values <- 3:7

results <- data.frame()

for (k in k_values) {

  set.seed(123)
  cl <- mfuzz(eset, c = k, m = m)

  membership <- cl$membership
  hard <- apply(membership, 1, which.max)

  # separation score
  sep <- apply(membership, 1, function(x) {
    sort(x, decreasing = TRUE)[1] - sort(x, decreasing = TRUE)[2]
  })

  results <- rbind(results, data.frame(
    k = k,
    mean_separation = mean(sep),
    median_separation = median(sep),
    low_conf_pct = mean(sep < 0.2),
    n_clusters = k
  ))
}

results
```



# oh my fucking god

```{r}
vsd <- vst(dds, blind = TRUE)
mat_vsd <- assay(vsd)

# remove low-expression genes
mat_vsd <- mat_vsd[apply(mat_vsd, 1, var) > 0, ]

# remove NA / Inf rows (correct object!)
mat_vsd <- mat_vsd[complete.cases(mat_vsd), ]

# ensure numeric safety
mat_vsd <- as.matrix(mat_vsd)
storage.mode(mat_vsd) <- "numeric"

```

```{r define-function}
evaluate_k <- function(k, m) {

  eset <- ExpressionSet(assayData = mat_vsd)

  cl <- mfuzz(eset, c = k, m = m)

  membership <- cl$membership

  sep <- apply(membership, 1, function(x) {
    sort(x, decreasing = TRUE)[1] - sort(x, decreasing = TRUE)[2]
  })

  hard <- apply(membership, 1, which.max)

  pca <- prcomp(mat_vsd, center = TRUE, scale. = TRUE)
  dist_pca <- dist(pca$x[, 1:5])

  sil <- silhouette(hard, dist_pca)

  data.frame(
    k = k,
    mean_fuzzy_sep = mean(sep, na.rm = TRUE),
    silhouette = mean(sil[, 3], na.rm = TRUE),
    prop_ambiguous = mean(sep < 0.2, na.rm = TRUE)
  )
}
```

```{r}
m <- mestimate(ExpressionSet(assayData = mat_vsd))

results <- do.call(rbind, lapply(3:7, function(k) {
  evaluate_k(k, m)
}))
```




```{r}
eset <- ExpressionSet(assayData = mat_vsd)
m <- mestimate(eset)
```

```{r}
get_plot_df <- function(k, m) {

  cl <- mfuzz(eset, c = k, m = m)

  cluster <- apply(cl$membership, 1, which.max)

  data.frame(
    gene = rownames(mat_vsd),
    log2FC_06h = mat_lfc[, "time"],
    log2FC_24h = mat_lfc[, "exposure"],
    cluster = factor(cluster),
    k = k
  )
}
```

```{r}
plot_list <- lapply(3:7, function(k) get_plot_df(k, m))

plot_df_all <- do.call(rbind, plot_list)
```


#######

```{r}
mat_lfc <- cbind(
  time = res_time$log2FoldChange,
  exposure = res_exposure$log2FoldChange,
  interaction = res_interaction$log2FoldChange
)
```

##### for fucking fuck's sake

```{r}
common_genes <- intersect(rownames(mat_vsd), rownames(mat_lfc))

mat_vsd <- mat_vsd[common_genes, ]
mat_lfc <- mat_lfc[common_genes, ]
```

```{r}
mat_lfc <- as.matrix(mat_lfc)
storage.mode(mat_lfc) <- "numeric"

mat_lfc <- mat_lfc[complete.cases(mat_lfc), ]
mat_lfc <- mat_lfc[apply(mat_lfc, 1, var) > 0, ]
```

```{r}
library(Mfuzz)

eset <- ExpressionSet(assayData = mat_lfc)
m <- mestimate(eset)
```

```{r}
evaluate_k <- function(k, m) {

  cl <- mfuzz(eset, c = k, m = m)

  membership <- cl$membership

  sep <- apply(membership, 1, function(x) {
    sort(x, decreasing = TRUE)[1] - sort(x, decreasing = TRUE)[2]
  })

  hard <- apply(membership, 1, which.max)

  # PCA in LFC space (3D geometry preserved)
  pca <- prcomp(mat_lfc, center = TRUE, scale. = TRUE)

  d <- dist(pca$x[, 1:3])
  sil <- silhouette(hard, d)

  data.frame(
    k = k,
    silhouette = mean(sil[, 3]),
    fuzzy_sep = mean(sep),
    ambiguity = mean(sep < 0.2)
  )
}
```

```{r}
results <- do.call(rbind, lapply(3:7, function(k) evaluate_k(k, m)))
results
```














```{r}
mat_lfc <- as.matrix(mat_lfc)
storage.mode(mat_lfc) <- "numeric"
```

```{r}
library(Mfuzz)

eset <- ExpressionSet(assayData = mat_lfc)
m <- mestimate(eset)
```

```{r}
get_plot_df <- function(k, m) {

  set.seed(123)

  cl <- mfuzz(eset, c = k, m = m)

  cluster <- apply(cl$membership, 1, which.max)

  data.frame(
    gene = rownames(mat_lfc),
    log2FC_06h = mat_lfc[, "time"],
    log2FC_24h = mat_lfc[, "exposure"],
    interaction = mat_lfc[, "interaction"],
    cluster = factor(cluster),
    k = k
  )
}
```

```{r}
plot_df_all <- do.call(
  rbind,
  lapply(3:7, function(k) get_plot_df(k, m))
)
```

```{r}
library(ggplot2)

ggplot(plot_df_all,
       aes(x = log2FC_06h,
           y = log2FC_24h,
           color = cluster)) +
  geom_point(alpha = 0.6, size = 1) +
  facet_wrap(~k) +
  theme_classic() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed")
```
```{r}
library(cluster)
library(Mfuzz)

eset <- ExpressionSet(assayData = mat_lfc)
m <- mestimate(eset)
```

```{r}
sil_scores <- sapply(3:7, function(k) {

  set.seed(123)

  cl <- mfuzz(eset, c = k, m = m)

  hard <- apply(cl$membership, 1, which.max)

  pca <- prcomp(mat_lfc, center = TRUE, scale. = TRUE)

  d <- dist(pca$x[, 1:3])

  mean(silhouette(hard, d)[, 3])
})
```

```{r}
get_traj_df <- function(k, m) {

  set.seed(123)

  cl <- mfuzz(eset, c = k, m = m)

  membership <- cl$membership

  cluster <- apply(membership, 1, which.max)

  plot_long <- data.frame(
    gene = rownames(mat_lfc),
    time = rep(colnames(mat_lfc), each = nrow(mat_lfc)),
    value = as.vector(t(mat_lfc)),
    cluster = rep(cluster, times = ncol(mat_lfc)),
    k = k
  )

  plot_long
}
```

```{r}
traj_all <- do.call(
  rbind,
  lapply(3:7, function(k) get_traj_df(k, m))
)
```

```{r}
library(ggplot2)

ggplot(traj_all,
       aes(x = time,
           y = value,
           group = gene,
           color = factor(cluster))) +
  geom_line(alpha = 0.05) +
  stat_summary(aes(group = cluster),
               fun = mean,
               geom = "line",
               linewidth = 1.5) +
  facet_wrap(~k) +
  theme_classic()
```

```{r}
plot_df <- plot_df[match(rownames(eset), plot_df$gene), ]
```

```{r}
library(Mfuzz)
library(tidyverse)

m <- mestimate(eset)

plot_list <- list()
```


#######

```{r}
mat_lfc <- as.matrix(mat_lfc)
storage.mode(mat_lfc) <- "numeric"

mat_lfc <- mat_lfc[complete.cases(mat_lfc), ]
mat_lfc <- mat_lfc[apply(mat_lfc, 1, var) > 0, ]
```

```{r}
library(Biobase)

eset <- ExpressionSet(assayData = mat_lfc)
```

```{r}
dim(eset)
head(exprs(eset))
```

```{r}
all(rownames(mat_lfc) == rownames(exprs(eset)))
```

```{r}
m <- mestimate(eset)
```

## SWITCH IT UP

filter genes from the contrasts to build the background universe

```{r}
# Define thresholds
p_cutoff <- 0.05
lfc_cutoff <- 0.5  # Adjust this to 0 if you want to filter purely on padj

# Find significant genes per axis
sig_time <- rownames(subset(res_time, padj < p_cutoff & abs(log2FoldChange) >= lfc_cutoff))
sig_exposure <- rownames(subset(res_exposure, padj < p_cutoff & abs(log2FoldChange) >= lfc_cutoff))
sig_interaction <- rownames(subset(res_interaction, padj < p_cutoff & abs(log2FoldChange) >= lfc_cutoff))

# Combine to create your final clustering universe (Union)
clustering_universe <- unique(c(sig_time, sig_exposure, sig_interaction))

# Build the 3D Matrix with this specific universe
df_3d <- data.frame(
  LFC_time = res_time[clustering_universe, "log2FoldChange"],
  LFC_exposure = res_exposure[clustering_universe, "log2FoldChange"],
  LFC_interaction = res_interaction[clustering_universe, "log2FoldChange"],
  row.names = clustering_universe
)

cat("Number of genes in your clustering universe:", nrow(df_3d), "\n")
```
```{r}
# Load required packages
library(Mfuzz)
library(cluster)
library(ggplot2)

# 1. Convert our 3D dataframe to a Matrix and then an ExpressionSet
mat_3d <- as.matrix(df_3d)
eset_3d <- new("ExpressionSet", exprs = mat_3d)

# 2. Estimate the fuzzifier parameter (m)
# This controls how "fuzzy" the boundaries between clusters are.
m_est <- mestimate(eset_3d)
cat("Estimated fuzzifier (m):", m_est, "\n")

# 3. Calculate Silhouette Scores for k = 2 through 10
k_range <- 2:10
avg_sil_widths <- numeric(length(k_range))

# We use e1071::cmeans because Mfuzz is built on top of it
set.seed(42) # For reproducibility
for (i in seq_along(k_range)) {
  k_val <- k_range[i]
  
  # Run fuzzy c-means
  fcm_res <- e1071::cmeans(mat_3d, centers = k_val, m = m_est, method = "cmeans")
  
  # Calculate silhouette widths based on the hardest cluster assignment
  sil_res <- silhouette(fcm_res$cluster, dist(mat_3d))
  avg_sil_widths[i] <- mean(sil_res[, "sil_width"])
}

# 4. Plot the Silhouette Scores to visually evaluate the optimal K
df_sil <- data.frame(k = k_range, avg_silhouette = avg_sil_widths)

ggplot(df_sil, aes(x = k, y = avg_silhouette)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "darkblue", size = 3) +
  scale_x_continuous(breaks = k_range) +
  theme_minimal() +
  labs(
    title = "Silhouette Analysis for Optimal K",
    x = "Number of Clusters (k)",
    y = "Average Silhouette Width"
  )
```

Local peak at 7/8, major dropoff at 9

```{r}
# 1. Run the final Mfuzz clustering with k = 7
set.seed(42) # For reproducibility
mfuzz_res <- mfuzz(eset_3d, centers = 8, m = m_est)

# 2. Extract membership values and cluster assignments
# 'membership' is a matrix showing how strongly each gene belongs to each of the 7 clusters
membership_matrix <- mfuzz_res$membership

# Assign each gene to its single highest-scoring cluster (hard clustering assignment)
gene_cluster_assignments <- data.frame(
  gene_id = rownames(mat_3d),
  cluster = mfuzz_res$cluster
)

# 3. Quick check: how many genes are in each of the 7 clusters?
table(gene_cluster_assignments$cluster)
```

```{r}
# ==========================================
# 1. EXTRACT TRUE MODEL-DERIVED LFC VALUES
# ==========================================

# Axis X: True Log2FC at 06h (Main effect of exposure)
res_LFC_06h <- res_06h_exposure
# Axis Y: True Log2FC at 24h 
# We use the 'contrast' argument to cleanly extract: (Exposure at 24h) - (Control at 24h)
res_LFC_24h <- res_24h_exposure
# ==========================================
# 2. BUILD THE PLOTTING DATAFRAME
# ==========================================
df_model_lfc <- data.frame(
  gene_id = clustering_universe,
  model_LFC_06h = res_LFC_06h[clustering_universe, "log2FoldChange"],
  model_LFC_24h = res_LFC_24h[clustering_universe, "log2FoldChange"]
)

# Merge with the Mfuzz clusters (k = 8)
df_final_plot <- inner_join(df_model_lfc, gene_clusters, by = "gene_id")

# ==========================================
# 3. PLOT THE 2D LFC SCATTERPLOT
# ==========================================
ggplot(df_final_plot, aes(x = model_LFC_06h, y = model_LFC_24h, color = cluster)) +
  geom_point(alpha = 0.6, size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  # Add a diagonal line to easily identify time-dependent changes
  geom_abline(intercept = 0, slope = 1, linetype = "dotted", color = "gray40") +
  scale_color_brewer(palette = "Set1") + 
  theme_minimal(base_size = 14) +
  labs(
    title = "Mfuzz 3D Clusters Projecting onto 2D LFC Space",
    subtitle = "Model-derived Log2 Fold Changes (Exposed vs Control)",
    x = "Log2 Fold Change at 06h",
    y = "Log2 Fold Change at 24h",
    color = "Cluster"
  ) +
  coord_fixed()
```




```{r}
library(DESeq2)
library(Mfuzz)
library(tidyverse)

# ==========================================
# 1. RUN MFUZZ WITH K = 8
# ==========================================
set.seed(42)
k_opt <- 8
mfuzz_res <- mfuzz(eset_3d, centers = k_opt, m = m_est)

# Extract cluster assignments
gene_clusters <- data.frame(
  gene_id = rownames(mat_3d),
  cluster = as.factor(mfuzz_res$cluster)
)

print(table(gene_clusters$cluster))

# ==========================================
# 2. CALCULATE 2D FOLD CHANGES FROM VSD
# ==========================================
# Extract the VSD expression matrix (log2-like scale)
vsd_mat <- assay(vsd) 

# Get sample metadata to map columns correctly
meta <- as.data.frame(colData(vsd))

# Add sample IDs as a column for easy grouping
meta$sample_id <- rownames(meta)

# We will average the replicates for our 4 distinct experimental groups
group_means <- unique(interaction(meta$exposure, meta$time))

# Function to safely get mean expression for a specific condition
get_group_mean <- function(exp_val, time_val) {
  samples <- meta$sample_id[meta$exposure == exp_val & meta$time == time_val]
  if(length(samples) == 1) {
    return(vsd_mat[clustering_universe, samples])
  } else {
    return(rowMeans(vsd_mat[clustering_universe, samples]))
  }
}

# Calculate the mean VSD values for our 4 groups across our 4,900 genes
mean_ctrl_06 <- get_group_mean("control", "06h")
mean_exp_06  <- get_group_mean("exposed", "06h")
mean_ctrl_24 <- get_group_mean("control", "24h")
mean_exp_24  <- get_group_mean("exposed", "24h")

# Calculate the 2D Fold Changes
df_2d <- data.frame(
  gene_id = clustering_universe,
  vsd_LFC_06h = mean_exp_06 - mean_ctrl_06,
  vsd_LFC_24h = mean_exp_24 - mean_ctrl_24
)

# Merge with our Mfuzz cluster assignments
df_plot <- inner_join(df_2d, gene_clusters, by = "gene_id")

# ==========================================
# 3. PLOT THE 2D SCATTERPLOT
# ==========================================
ggplot(df_plot, aes(x = vsd_LFC_06h, y = vsd_LFC_24h, color = cluster)) +
  geom_point(alpha = 0.6, size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  scale_color_brewer(palette = "Set1") + # Distinct colors for 8 clusters
  theme_minimal(base_size = 14) +
  labs(
    title = "Projection of Mfuzz 3D Clusters onto 2D Space",
    subtitle = "VSD-derived Log2 Fold Changes (Exposed vs Control)",
    x = "Log2 Fold Change at 06h",
    y = "Log2 Fold Change at 24h",
    color = "Cluster"
  ) +
  theme(legend.position = "right") +
  coord_fixed() # Keeps the scales uniform on both axes
```

```{r}
# 1. Keep the exact same group means we calculated using the correct levels
mean_ctrl_06 <- rowMeans(vsd_mat[clustering_universe, group_list[["Control.06h"]]])
mean_exp_06  <- rowMeans(vsd_mat[clustering_universe, group_list[["Exposed.06h"]]])
mean_ctrl_24 <- rowMeans(vsd_mat[clustering_universe, group_list[["Control.24h"]]])
mean_exp_24  <- rowMeans(vsd_mat[clustering_universe, group_list[["Exposed.24h"]]])

# 2. Build the dataframe using the ABSOLUTE expression values 
# Let's look at the trajectories of the EXPOSED samples over time, 
# or you could swap these out for the Control samples.
df_absolute_vsd <- data.frame(
  gene_id = clustering_universe,
  vsd_expression_06h = mean_exp_06, # Absolute VSD value at 6h
  vsd_expression_24h = mean_exp_24  # Absolute VSD value at 24h
)

# 3. Merging with the Mfuzz clusters
df_plot_vsd <- inner_join(df_absolute_vsd, gene_clusters, by = "gene_id")

# 4. Plotting absolute expression trajectories
ggplot(df_plot_vsd, aes(x = vsd_expression_06h, y = vsd_expression_24h, color = cluster)) +
  geom_point(alpha = 0.5, size = 1.2) +
  # Add a diagonal line: genes on this line have identical expression at both timepoints
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50") +
  scale_color_brewer(palette = "Set1") + 
  theme_minimal(base_size = 14) +
  labs(
    title = "Projection of 3D Clusters onto Absolute VSD Expression",
    subtitle = "Exposed Group: 06h vs 24h",
    x = "Variance Stabilized Expression at 06h",
    y = "Variance Stabilized Expression at 24h",
    color = "Cluster"
  ) +
  coord_fixed()
```

```{r}
# ==========================================
# 0. CONFIGURATION: SET YOUR K VALUE HERE
# ==========================================
k_val <- 7  # <-- Change this single number to dynamically update the whole block

# ==========================================
# 1. RUN MFUZZ CLUSTERING DYNAMICALLY
# ==========================================
set.seed(42) # For reproducibility
mfuzz_out <- mfuzz(eset_3d, centers = k_val, m = m_est)

# Update the gene cluster assignments mapping
gene_clusters_dynamic <- data.frame(
  gene_id = rownames(exprs(eset_3d)),
  cluster = as.factor(mfuzz_out$cluster)
)

# Print the breakdown of cluster sizes to the console
message(paste("--- Cluster Sizes for K =", k_val, "---"))
print(table(gene_clusters_dynamic$cluster))

# ==========================================
# 2. BUILD THE PLOTTING DATAFRAME
# ==========================================
# Axis X: True Log2FC at 06h (Main effect of exposure)
res_LFC_06h <- res_06h_exposure

# Axis Y: True Log2FC at 24h 
res_LFC_24h <- res_24h_exposure

df_model_lfc <- data.frame(
  gene_id = clustering_universe,
  model_LFC_06h = res_LFC_06h[clustering_universe, "log2FoldChange"],
  model_LFC_24h = res_LFC_24h[clustering_universe, "log2FoldChange"]
)

# Merge with the dynamic Mfuzz clusters
df_final_plot <- inner_join(df_model_lfc, gene_clusters_dynamic, by = "gene_id")

# ==========================================
# 3. PLOT THE 2D LFC SCATTERPLOT
# ==========================================
ggplot(df_final_plot, aes(x = model_LFC_06h, y = model_LFC_24h, color = cluster)) +
  geom_point(alpha = 0.6, size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  # Add a diagonal line to easily identify time-dependent changes
  geom_abline(intercept = 0, slope = 1, linetype = "dotted", color = "gray40") +
  # Using scale_color_discrete() so it doesn't break if K exceeds ColorBrewer limits
  scale_color_discrete() + 
  theme_minimal(base_size = 14) +
  labs(
    title = paste("Mfuzz 3D Clusters Projecting onto 2D LFC Space (K =", k_val, ")"),
    subtitle = "Model-derived Log2 Fold Changes (Exposed vs Control)",
    x = "Log2 Fold Change at 06h",
    y = "Log2 Fold Change at 24h",
    color = "Cluster"
  ) +
  coord_fixed()
```

```{r}
# Make sure you have these packages installed for high-contrast colors
# install.packages("pals") 
# install.packages("plotly") # Optional, for interactive hovering

library(pals)
library(plotly)

# ==========================================
# 0. CONFIGURATION: SET K AND PALETTE HERE
# ==========================================
k_val <- 7  

# Choose your style: 
# "glasbey"  -> Maximum distinctness (Up to 32 completely different colors)
# "kelly"    -> High contrast classic palette
# "turbo"    -> Vibrant, continuous-looking rainbow with distinct saturation
palette_type <- "glasbey" 

# ==========================================
# 1. RUN MFUZZ CLUSTERING DYNAMICALLY
# ==========================================
set.seed(42) 
mfuzz_out <- mfuzz(eset_3d, centers = k_val, m = m_est)

gene_clusters_dynamic <- data.frame(
  gene_id = rownames(exprs(eset_3d)),
  cluster = as.factor(mfuzz_out$cluster)
)

# ==========================================
# 2. BUILD THE PLOTTING DATAFRAME
# ==========================================
res_LFC_06h <- res_06h_exposure
res_LFC_24h <- res_24h_exposure

df_model_lfc <- data.frame(
  gene_id = clustering_universe,
  model_LFC_06h = res_LFC_06h[clustering_universe, "log2FoldChange"],
  model_LFC_24h = res_LFC_24h[clustering_universe, "log2FoldChange"]
)

df_final_plot <- inner_join(df_model_lfc, gene_clusters_dynamic, by = "gene_id")

# ==========================================
# 3. SET UP THE COLOR PALETTE DYNAMICALLY
# ==========================================
color_vector <- switch(palette_type,
  "glasbey" = as.vector(pals::glasbey(k_val)),
  "kelly"   = as.vector(pals::kelly(k_val + 1)[-1]), # Drop white/black baseline
  "turbo"   = as.vector(pals::turbo(k_val))
)

# ==========================================
# 4. PLOT THE 2D LFC SCATTERPLOT
# ==========================================
p <- ggplot(df_final_plot, aes(x = model_LFC_06h, y = model_LFC_24h, color = cluster, text = gene_id)) +
  geom_point(alpha = 0.7, size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted", color = "gray40") +
  # Inject our high-contrast color vector
  scale_color_manual(values = color_vector) + 
  theme_minimal(base_size = 14) +
  labs(
    title = paste("Mfuzz 3D Clusters on 2D LFC Space (K =", k_val, ")"),
    subtitle = paste("Palette:", palette_type),
    x = "Log2 Fold Change at 06h",
    y = "Log2 Fold Change at 24h",
    color = "Cluster"
  ) +
  coord_fixed() +
  theme(legend.position = "right")

# Display the static plot
print(p)

# OPTIONAL: Uncomment the line below to make the plot fully interactive in RStudio!
# ggplotly(p, tooltip = "text")
```

```{r}
# Ensure you have plotly installed: install.packages("plotly")
library(paletteer)
library(plotly)

# ==========================================
# 0. CONFIGURATION: SET K AND PALETTE HERE
# ==========================================
k_val <- 6  

# Using paletteer to select a highly distinct 10-color palette
# Feel free to change this to "awtools::mpalette" or "colorblindr::OkabeIto"
chosen_palette <- "ggthemes::Classic_10"

# ==========================================
# 1. RUN MFUZZ CLUSTERING DYNAMICALLY
# ==========================================
set.seed(42) 
mfuzz_out <- mfuzz(eset_3d, centers = k_val, m = m_est)

gene_clusters_dynamic <- data.frame(
  gene_id = rownames(exprs(eset_3d)),
  cluster = as.factor(mfuzz_out$cluster)
)

# ==========================================
# 2. BUILD THE PLOTTING DATAFRAME
# ==========================================
res_LFC_06h <- res_06h_exposure
res_LFC_24h <- res_24h_exposure

df_model_lfc <- data.frame(
  gene_id = clustering_universe,
  model_LFC_06h = res_LFC_06h[clustering_universe, "log2FoldChange"],
  model_LFC_24h = res_LFC_24h[clustering_universe, "log2FoldChange"]
)

df_final_plot <- inner_join(df_model_lfc, gene_clusters_dynamic, by = "gene_id")

# ==========================================
# 3. BUILD THE STATIC GGPLOT
# ==========================================
# Crucial: We add a 'text' aesthetic mapping inside aes(). 
# This tells plotly exactly what to display in the hover tooltip box.
p_static <- ggplot(df_final_plot, aes(
    x = model_LFC_06h, 
    y = model_LFC_24h, 
    color = cluster,
    text = paste0(
      "<b>Gene:</b> ", gene_id, "<br>",
      "<b>Cluster:</b> ", cluster, "<br>",
      "<b>LFC 06h:</b> ", round(model_LFC_06h, 3), "<br>",
      "<b>LFC 24h:</b> ", round(model_LFC_24h, 3)
    )
  )) +
  geom_point(alpha = 0.6, size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted", color = "gray40") +
  scale_color_paletteer_d(chosen_palette) + 
  theme_minimal(base_size = 12) +
  labs(
    title = paste("Interactive Mfuzz 3D Clusters Projected onto 2D LFC Space (K =", k_val, ")"),
    x = "Log2 Fold Change at 06h",
    y = "Log2 Fold Change at 24h",
    color = "Cluster"
  )

# ==========================================
# 4. CONVERT TO INTERACTIVE PLOTLY
# ==========================================
# 'tooltip = "text"' tells plotly to ONLY show our custom formatted string on hover
p_interactive <- ggplotly(p_static, tooltip = "text") %>%
  # This option ensures the 1:1 aspect ratio layout carries over perfectly into plotly
  layout(yaxis = list(scaleanchor = "x", scaleratio = 1))

# Display the interactive plot in your RStudio Viewer or RMarkdown output
p_interactive
```

```{r}
# Make sure you have these installed: 
# install.packages("pals")
# install.packages("plotly")

library(pals)
library(plotly)

# ==========================================
# 0. CONFIGURATION: SET YOUR K VALUE HERE
# ==========================================
k_val <- 7  # <-- Change this single number to dynamically update everything

# Choose your pals palette: 
# "glasbey" -> Max distinctness (adjacent clusters get radically different colors)
# "kelly"   -> Classic maximum-contrast palette
chosen_pals_style <- "glasbey"

# ==========================================
# 1. RUN MFUZZ CLUSTERING DYNAMICALLY
# ==========================================
set.seed(42) # For reproducibility
mfuzz_out <- mfuzz(eset_3d, centers = k_val, m = m_est)

# Update the gene cluster assignments mapping
gene_clusters_dynamic <- data.frame(
  gene_id = rownames(exprs(eset_3d)),
  cluster = as.factor(mfuzz_out$cluster)
)

# Print the breakdown of cluster sizes to the console
message(paste("--- Cluster Sizes for K =", k_val, "---"))
print(table(gene_clusters_dynamic$cluster))

# ==========================================
# 2. BUILD THE PLOTTING DATAFRAME
# ==========================================
res_LFC_06h <- res_06h_exposure
res_LFC_24h <- res_24h_exposure

df_model_lfc <- data.frame(
  gene_id = clustering_universe,
  model_LFC_06h = res_LFC_06h[clustering_universe, "log2FoldChange"],
  model_LFC_24h = res_LFC_24h[clustering_universe, "log2FoldChange"]
)

# Merge with the dynamic Mfuzz clusters
df_final_plot <- inner_join(df_model_lfc, gene_clusters_dynamic, by = "gene_id")

# ==========================================
# 3. EXTRACT THE PALS COLOR VECTOR
# ==========================================
color_vector <- switch(chosen_pals_style,
  "glasbey" = as.vector(pals::glasbey(k_val)),
  "kelly"   = as.vector(pals::kelly(k_val + 1)[-1]) # Drops the baseline white/black
)

# ==========================================
# 4. BUILD THE STATIC GGPLOT GRAPHIC
# ==========================================
p_static <- ggplot(df_final_plot, aes(
    x = model_LFC_06h, 
    y = model_LFC_24h, 
    color = cluster,
    text = paste0(
      "<b>Gene:</b> ", gene_id, "<br>",
      "<b>Cluster:</b> ", cluster, "<br>",
      "<b>LFC 06h:</b> ", round(model_LFC_06h, 3), "<br>",
      "<b>LFC 24h:</b> ", round(model_LFC_24h, 3)
    )
  )) +
  geom_point(alpha = 0.7, size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted", color = "gray40") +
  # Force our custom pals high-contrast colors
  scale_color_manual(values = color_vector) + 
  theme_minimal(base_size = 12) +
  labs(
    title = paste("Interactive Mfuzz 3D Clusters Projected onto 2D LFC Space (K =", k_val, ")"),
    subtitle = paste("Color Profile:", chosen_pals_style),
    x = "Log2 Fold Change at 06h",
    y = "Log2 Fold Change at 24h",
    color = "Cluster"
  )

# ==========================================
# 5. CONVERT TO INTERACTIVE HTML WIDGET
# ==========================================
p_interactive <- ggplotly(p_static, tooltip = "text") %>%
  # Ensures the coordinate aspect ratio stays perfectly square (1:1) in the browser
  layout(yaxis = list(scaleanchor = "x", scaleratio = 1))

# Run this to render the plot
p_interactive
```


filter out genes only significant wrt time

```{r}
library(pals)
library(plotly)

# ==========================================
# 0. CONFIGURATION: SET YOUR K VALUE HERE
# ==========================================
k_val <- 7  
chosen_pals_style <- "glasbey"
p_cutoff <- 0.05

# ==========================================
# 1. SURGICAL UNIVERSE FILTERING (NO TIME-ONLY GENES)
# ==========================================
# Extract all significant genes per contrast
sig_time        <- rownames(subset(res_time,        padj < p_cutoff))
sig_exposure    <- rownames(subset(res_exposure,    padj < p_cutoff))
sig_interaction <- rownames(subset(res_interaction, padj < p_cutoff))

# Define your new universe: Significant for Exposure OR Interaction
# (This automatically drops genes that are ONLY significant for time)
treatment_universe <- unique(c(sig_exposure, sig_interaction))

cat("Original universe size with time-only genes:", length(unique(c(sig_time, sig_exposure, sig_interaction))), "\n")
cat("Refined universe size (Treatment-responsive only):", length(treatment_universe), "\n")

# ==========================================
# 2. RE-BUILD THE 3D EXPRESSION SET FOR MFUZZ
# ==========================================
# We must subset our original 3D matrix to match this new treatment_universe
mat_3d_refined <- mat_3d[rownames(mat_3d) %in% treatment_universe, ]
eset_3d_refined <- new("ExpressionSet", exprs = mat_3d_refined)

# Re-estimate the fuzzifier parameter for this cleaner dataset
m_est_refined <- mestimate(eset_3d_refined)

# ==========================================
# 3. RUN MFUZZ ON REFINED DATA
# ==========================================
set.seed(42) 
mfuzz_out <- mfuzz(eset_3d_refined, centers = k_val, m = m_est_refined)

gene_clusters_dynamic <- data.frame(
  gene_id = rownames(exprs(eset_3d_refined)),
  cluster = as.factor(mfuzz_out$cluster)
)

message(paste("--- Refined Cluster Sizes for K =", k_val, "---"))
print(table(gene_clusters_dynamic$cluster))

# ==========================================
# 4. BUILD THE PLOTTING DATAFRAME
# ==========================================
res_LFC_06h <- res_06h_exposure
res_LFC_24h <- res_24h_exposure

df_model_lfc <- data.frame(
  gene_id = treatment_universe,
  model_LFC_06h = res_LFC_06h[treatment_universe, "log2FoldChange"],
  model_LFC_24h = res_LFC_24h[treatment_universe, "log2FoldChange"]
)

df_final_plot <- inner_join(df_model_lfc, gene_clusters_dynamic, by = "gene_id")

# ==========================================
# 5. PLOT THE NEW INTERACTIVE SCATTERPLOT
# ==========================================
color_vector <- switch(chosen_pals_style,
  "glasbey" = as.vector(pals::glasbey(k_val)),
  "kelly"   = as.vector(pals::kelly(k_val + 1)[-1])
)

p_static <- ggplot(df_final_plot, aes(
    x = model_LFC_06h, 
    y = model_LFC_24h, 
    color = cluster,
    text = paste0(
      "<b>Gene:</b> ", gene_id, "<br>",
      "<b>Cluster:</b> ", cluster, "<br>",
      "<b>LFC 06h:</b> ", round(model_LFC_06h, 3), "<br>",
      "<b>LFC 24h:</b> ", round(model_LFC_24h, 3)
    )
  )) +
  geom_point(alpha = 0.7, size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted", color = "gray40") +
  scale_color_manual(values = color_vector) + 
  theme_minimal(base_size = 12) +
  labs(
    title = paste("Treatment-Only Mfuzz Clusters on 2D LFC Space (K =", k_val, ")"),
    subtitle = "Dropped baseline time-only genes",
    x = "Log2 Fold Change at 06h",
    y = "Log2 Fold Change at 24h",
    color = "Cluster"
  )

p_interactive <- ggplotly(p_static, tooltip = "text") %>%
  layout(yaxis = list(scaleanchor = "x", scaleratio = 1))

p_interactive
```

```{r}
# ==========================================
# DIAGNOSTIC: ASSESS NEW OPTIMAL K
# ==========================================
library(Mfuzz)

# Define a range of cluster numbers to test
k_range <- 3:10
distances <- numeric(length(k_range))

# Loop through and calculate the minimum distance between cluster centroids for each K
set.seed(42)
for (i in seq_along(k_range)) {
  k_test <- k_range[i]
  cl_test <- mfuzz(eset_3d_refined, centers = k_test, m = m_est_refined)
  # Calculate overlap/distance between centers
  distances[i] <- min(dist(cl_test$centers))
}

# Create a clean diagnostic summary table
k_evaluation <- data.frame(
  K = k_range,
  Min_Centroid_Distance = round(distances, 4)
)

print("--- Centroid Distance Evaluation Table ---")
print(k_evaluation)

# Plot the 'Elbow' style curve
plot(k_range, distances, type = "b", pch = 19, col = "darkblue",
     xlab = "Number of Clusters (K)", ylab = "Minimum Distance Between Centroids",
     main = "Mfuzz K-Selection Curve (Refined Data)")
```

# with lfc shrinkage

```{r}
# ==========================================
# DIAGNOSTIC: ASSESS K FOR SHRUNKEN DATA
# ==========================================
library(Mfuzz)

# Define a range of cluster numbers to test on the shrunken data
k_range <- 3:10
distances_shrunk <- numeric(length(k_range))

# Loop through and calculate the minimum distance between cluster centroids
set.seed(42)
for (i in seq_along(k_range)) {
  k_test <- k_range[i]
  cl_test <- mfuzz(eset_3d_refined, centers = k_test, m = m_est_refined)
  # Calculate the minimum Euclidean distance between any two cluster centers
  distances_shrunk[i] <- min(dist(cl_test$centers))
}

# Generate the evaluation table
k_evaluation_shrunk <- data.frame(
  K = k_range,
  Min_Centroid_Distance = round(distances_shrunk, 4)
)

print("--- Post-Shrinkage Centroid Distance Evaluation ---")
print(k_evaluation_shrunk)

# Plot the new curve
plot(k_range, distances_shrunk, type = "b", pch = 19, col = "darkred",
     xlab = "Number of Clusters (K)", ylab = "Minimum Distance Between Centroids",
     main = "Mfuzz K-Selection Curve (Shrunken + Refined Data)")
```
```{r}
library(tidyverse)
library(tidyverse)
library(plotly)
library(pals)

# ==========================================
# 0. RE-BUILD MASTER DATA WITH EXACT COLUMN NAMES
# ==========================================
df_plot_master <- data.frame(
  gene_id       = treatment_universe,
  model_LFC_06h = res_06h_shrunk[treatment_universe, "log2FoldChange"],
  model_LFC_24h = res_24h_shrunk[treatment_universe, "log2FoldChange"]
) %>%
  inner_join(gene_clusters_dynamic, by = "gene_id") %>%
  inner_join(df_membership, by = "gene_id") %>%
  mutate(is_core = if_else(membership_score > 0.25, "Core (>0.25)", "Borderline (<=0.25)"))

# ==========================================
# 1. SETUP TWO-TONE GLASBEY COLOR MAP
# ==========================================
k_val <- 7
cutoff <- 0.25
base_colors <- as.vector(pals::glasbey(k_val))

color_map <- data.frame(
  cluster = as.factor(1:k_val),
  base_color = base_colors,
  light_color = paste0(base_colors, "40") # 25% opacity for background halos
)

df_plot_master <- df_plot_master %>%
  inner_join(color_map, by = "cluster") %>%
  mutate(final_color = if_else(membership_score > cutoff, as.character(base_color), as.character(light_color)))

# ==========================================
# PLOT 1: ALL ASSIGNED GENES (SHRUNKEN)
# ==========================================
p1_static <- ggplot(df_plot_master, aes(
    x = model_LFC_06h, y = model_LFC_24h, color = cluster,
    text = paste0("<b>Gene:</b> ", gene_id, "<br><b>Cluster:</b> ", cluster, "<br><b>LFC 06h:</b> ", round(model_LFC_06h, 3), "<br><b>Score:</b> ", round(membership_score, 3))
  )) +
  geom_point(alpha = 0.7, size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted", color = "gray40") +
  scale_color_manual(values = base_colors) + 
  theme_minimal(base_size = 11) +
  labs(title = "Plot 1: All Assigned Genes (Shrunken LFC Space)",
       x = "Log2 Fold Change at 06h (model_LFC_06h)", y = "Log2 Fold Change at 24h (model_LFC_24h)")

p1_interactive <- ggplotly(p1_static, tooltip = "text") %>% layout(yaxis = list(scaleanchor = "x", scaleratio = 1))

# ==========================================
# PLOT 2: THE 0.25 CUT ONLY (SHRUNKEN CORE)
# ==========================================
df_plot_filtered <- df_plot_master %>% filter(membership_score > cutoff)

p2_static <- ggplot(df_plot_filtered, aes(
    x = model_LFC_06h, y = model_LFC_24h, color = cluster,
    text = paste0("<b>Gene:</b> ", gene_id, "<br><b>Cluster:</b> ", cluster, "<br><b>LFC 06h:</b> ", round(model_LFC_06h, 3), "<br><b>Score:</b> ", round(membership_score, 3))
  )) +
  geom_point(alpha = 0.8, size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted", color = "gray40") +
  scale_color_manual(values = base_colors) + 
  theme_minimal(base_size = 11) +
  labs(title = paste("Plot 2: Core Genes Only (Membership >", cutoff, ")"),
       x = "Log2 Fold Change at 06h (model_LFC_06h)", y = "Log2 Fold Change at 24h (model_LFC_24h)")

p2_interactive <- ggplotly(p2_static, tooltip = "text") %>% layout(yaxis = list(scaleanchor = "x", scaleratio = 1))

# ==========================================
# PLOT 3: TWO-TONE HIGHLIGHTING (SHRUNKEN)
# ==========================================
p3_static <- ggplot(df_plot_master, aes(
    x = model_LFC_06h, y = model_LFC_24h, 
    color = final_color,
    shape = is_core,
    text = paste0("<b>Gene:</b> ", gene_id, "<br><b>Cluster:</b> ", cluster, "<br><b>Score:</b> ", round(membership_score, 3), "<br><b>Status:</b> ", is_core)
  )) +
  geom_point(aes(size = is_core)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted", color = "gray40") +
  
  scale_color_identity() + 
  scale_shape_manual(values = c("Core (>0.25)" = 19, "Borderline (<=0.25)" = 16)) +
  scale_size_manual(values = c("Core (>0.25)" = 1.6, "Borderline (<=0.25)" = 1.0)) +
  
  theme_minimal(base_size = 11) +
  labs(title = "Plot 3: Two-Tone Membership Highlight (Shrunken Space)",
       x = "Log2 Fold Change at 06h (model_LFC_06h)", y = "Log2 Fold Change at 24h (model_LFC_24h)",
       shape = "Membership Status")

p3_interactive <- ggplotly(p3_static, tooltip = "text") %>% layout(yaxis = list(scaleanchor = "x", scaleratio = 1))
```
highlight core genes

```{r}
# ==========================================
# PLOT WITH A MORE LENIENT MEMBERSHIP THRESHOLD
# ==========================================

# Lowering the cutoff to 0.30 to recover valid trajectories
df_lenient_profiles <- df_profiles_mem %>% 
  filter(membership_score > 0.30) 

ggplot(df_lenient_profiles, aes(x = Timepoint, y = LFC, group = gene_id, color = cluster)) +
  # Alpha mapping automatically handles the fading based on individual score
  geom_line(aes(alpha = membership_score)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  facet_wrap(~cluster, labeller = label_both, ncol = 4) +
  scale_color_manual(values = color_vector) +
  
  # Fine-tune the fade: lower values are fainter, higher values are more solid
  scale_alpha_continuous(range = c(0.08, 0.5)) + 
  
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    strip.background = element_rect(fill = "gray95", color = NA),
    panel.spacing = unit(1, "lines")
  ) +
  labs(
    title = "Exposure Trajectories (Membership > 0.30)",
    subtitle = "Lenient threshold: Shaded by mathematical fit to archetype",
    x = "Post-Exposure Timepoint",
    y = "Log2 Fold Change"
  )
```

```{r}
library(tidyverse)

# ==========================================
# COMPUTE GENE COUNTS AT DIFFERENT THRESHOLDS
# ==========================================

# 1. Total genes assigned to each cluster (No threshold)
count_all <- df_final_plot_with_mem %>%
  count(cluster, name = "Total_Assigned")

# 2. Count genes with membership > 0.25
count_25 <- df_final_plot_with_mem %>%
  filter(membership_score > 0.25) %>%
  count(cluster, name = "Cut_0.25")

# 3. Count genes with membership > 0.30
count_30 <- df_final_plot_with_mem %>%
  filter(membership_score > 0.30) %>%
  count(cluster, name = "Cut_0.30")

# 4. Count genes with membership > 0.35
count_35 <- df_final_plot_with_mem %>%
  filter(membership_score > 0.35) %>%
  count(cluster, name = "Cut_0.35")

# 5. Count genes with membership > 0.40
count_40 <- df_final_plot_with_mem %>%
  filter(membership_score > 0.40) %>%
  count(cluster, name = "Cut_0.40")

# ==========================================
# MERGE INTO A DIAGNOSTIC AUDIT TABLE
# ==========================================
threshold_audit <- count_all %>%
  left_join(count_25, by = "cluster") %>%
  left_join(count_30, by = "cluster") %>%
  left_join(count_35, by = "cluster") %>%
  left_join(count_40, by = "cluster") %>%
  # Replace any NA values with 0 (in case a tight cluster drops completely at high cuts)
  mutate(across(everything(), ~replace_na(., 0)))

print("--- Cluster Stability Audit Table ---")
print(threshold_audit)
```

```{r}
# Lowering the cutoff to 0.30 to recover valid trajectories
df_lenient_profiles <- df_profiles_mem %>% 
  filter(membership_score > 0.25) 

ggplot(df_lenient_profiles, aes(x = Timepoint, y = LFC, group = gene_id, color = cluster)) +
  # Alpha mapping automatically handles the fading based on individual score
  geom_line(aes(alpha = membership_score)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  facet_wrap(~cluster, labeller = label_both, ncol = 4) +
  scale_color_manual(values = color_vector) +
  
  # Fine-tune the fade: lower values are fainter, higher values are more solid
  scale_alpha_continuous(range = c(0.08, 0.5)) + 
  
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    strip.background = element_rect(fill = "gray95", color = NA),
    panel.spacing = unit(1, "lines")
  ) +
  labs(
    title = "Exposure Trajectories (Membership > 0.25)",
    subtitle = "Lenient threshold: Shaded by mathematical fit to archetype",
    x = "Post-Exposure Timepoint",
    y = "Log2 Fold Change"
  )
```




```{r}
# ==========================================
# EXTRACT FINAL HIGH-CONFIDENCE GENE LISTS (CUT = 0.25)
# ==========================================

final_cluster_genes <- df_final_plot_with_mem %>%
  filter(membership_score > 0.25) %>%
  select(gene_id, cluster, membership_score, model_LFC_06h, model_LFC_24h)

# Split into an R list for easy downstream pathway analysis
final_gene_lists <- split(final_cluster_genes$gene_id, final_cluster_genes$cluster)

# Quick check to ensure the counts match our Cut_0.25 column exactly
print(sapply(final_gene_lists, length))
```


```{r}
library(tidyverse)
library(plotly)
library(pals)

# ==========================================
# BRIDGE ENVIRONMENT VARIABLES
# ==========================================
res_06h_shrunk <- res_LFC_06h
res_24h_shrunk <- res_LFC_24h

# ==========================================
# RE-RUN MASTER DATA GENERATION
# ==========================================
df_plot_master <- data.frame(
  gene_id       = treatment_universe,
  model_LFC_06h = res_06h_shrunk[treatment_universe, "log2FoldChange"],
  model_LFC_24h = res_24h_shrunk[treatment_universe, "log2FoldChange"]
) %>%
  inner_join(gene_clusters_dynamic, by = "gene_id") %>%
  inner_join(df_membership, by = "gene_id") %>%
  mutate(is_core = if_else(membership_score > 0.25, "Core (>0.25)", "Borderline (<=0.25)"))

# ==========================================
# SETUP TWO-TONE GLASBEY COLOR MAP
# ==========================================
k_val <- 7
cutoff <- 0.25
base_colors <- as.vector(pals::glasbey(k_val))

color_map <- data.frame(
  cluster = as.factor(1:k_val),
  base_color = base_colors,
  light_color = paste0(base_colors, "40") # 25% opacity for background halos
)

df_plot_master <- df_plot_master %>%
  inner_join(color_map, by = "cluster") %>%
  mutate(final_color = if_else(membership_score > cutoff, as.character(base_color), as.character(light_color)))

# ==========================================
# GENERATE PLOT 3: TWO-TONE HIGHLIGHTING
# ==========================================
p3_static <- ggplot(df_plot_master, aes(
    x = model_LFC_06h, y = model_LFC_24h, 
    color = final_color,
    shape = is_core,
    text = paste0("<b>Gene:</b> ", gene_id, "<br><b>Cluster:</b> ", cluster, "<br><b>Score:</b> ", round(membership_score, 3), "<br><b>Status:</b> ", is_core)
  )) +
  geom_point(aes(size = is_core)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_abline(intercept = 0, slope = 1, linetype = "dotted", color = "gray40") +
  
  scale_color_identity() + 
  scale_shape_manual(values = c("Core (>0.25)" = 19, "Borderline (<=0.25)" = 16)) +
  scale_size_manual(values = c("Core (>0.25)" = 1.6, "Borderline (<=0.25)" = 1.0)) +
  
  theme_minimal(base_size = 11) +
  labs(title = "Plot 3: Two-Tone Membership Highlight (Shrunken Space)",
       x = "Log2 Fold Change at 06h (model_LFC_06h)", y = "Log2 Fold Change at 24h (model_LFC_24h)",
       shape = "Membership Status")

p3_interactive <- ggplotly(p3_static, tooltip = "text") %>% layout(yaxis = list(scaleanchor = "x", scaleratio = 1))

# View the final two-tone map
p3_interactive
```

add a logfc filter to keep gene membership in a cluster if the log2fc is high 

```{r}
# ==========================================
# HYBRID CAPTURE: CORE SHAPES + BIOLOGICAL POWERHOUSES
# ==========================================

df_HYBRID_CLEAN <- df_SHRUNK_FINAL %>%
  mutate(
    # A gene is a powerhouse if its absolute LFC is greater than 1.5 at either timepoint
    is_powerhouse = (abs(model_LFC_06h) > 1.5 | abs(model_LFC_24h) > 1.5),
    
    # Final inclusion status: Keep if it passes the shape test OR the magnitude test
    keep_for_downstream = (membership_score > 0.25 | is_powerhouse)
  )

# Let's see how many powerhouse genes we just rescued
rescued_count <- df_HYBRID_CLEAN %>% 
  filter(membership_score <= 0.25 & is_powerhouse == TRUE) %>% 
  nrow()

message("Rescued ", rescued_count, " high-magnitude genes that fuzzy clustering penalized!")
```

```{r}
library(tidyverse)
library(plotly)
library(pals)

# ==========================================
# 1. CLASSIFY GENES BY HYBRID CRITERIA
# ==========================================
df_hybrid_plot <- df_SHRUNK_FINAL %>%
  mutate(
    is_powerhouse = (abs(model_LFC_06h) > 1.5 | abs(model_LFC_24h) > 1.5),
    is_shape_core = (membership_score > 0.25),
    
    # Establish the four visual categories
    hybrid_status = case_when(
      is_shape_core & is_powerhouse   ~ "Dual Champion (Shape & Magnitude)",
      !is_shape_core & is_powerhouse  ~ "Powerhouse Rescued (Magnitude Only)",
      is_shape_core & !is_powerhouse  ~ "Core Trajectory (Shape Only)",
      TRUE                            ~ "Filtered Out (Fails Both)"
    )
  )

# ==========================================
# 2. DEFINE HIGH-CONTRAST SEPARATION COLORS
# ==========================================
# Base colors for the 7 clusters
k_val <- 7
base_colors <- as.vector(pals::glasbey(k_val))

# Generate the custom coloring matrix directly in the dataframe
df_hybrid_plot <- df_hybrid_plot %>%
  mutate(final_color = case_when(
    # Filtered out genes turn a uniform, faint light gray
    hybrid_status == "Filtered Out (Fails Both)" ~ "#E0E0E080",
    
    # Rescued powerhouses get the bright, solid cluster color so they stand out
    hybrid_status == "Powerhouse Rescued (Magnitude Only)" ~ as.character(base_color),
    
    # Dual champions get the bright, solid cluster color
    hybrid_status == "Dual Champion (Shape & Magnitude)" ~ as.character(base_color),
    
    # Standard trajectory shape-followers get a slightly softer, cleaner tint
    hybrid_status == "Core Trajectory (Shape Only)" ~ paste0(base_color, "99")
  ))

# ==========================================
# 3. GENERATE THE INTERACTIVE HYBRID PLOT
# ==========================================
p4_static <- ggplot(df_hybrid_plot, aes(
    x = model_LFC_06h, 
    y = model_LFC_24h, 
    color = final_color,
    shape = hybrid_status,
    text = paste0("<b>Gene:</b> ", gene_id, 
                  "<br><b>Cluster:</b> ", cluster, 
                  "<br><b>Score:</b> ", round(membership_score, 3), 
                  "<br><b>Type:</b> ", hybrid_status)
  )) +
  # Draw background filtered points first, then overlay active groups
  geom_point(aes(size = hybrid_status)) +
  
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  # Reference gates for the LFC 1.5 thresholds
  geom_hline(yintercept = c(-1.5, 1.5), linetype = "dotted", color = "gray70") +
  geom_vline(xintercept = c(-1.5, 1.5), linetype = "dotted", color = "gray70") +
  
  scale_color_identity() +
  scale_shape_manual(values = c(
    "Dual Champion (Shape & Magnitude)"   = 19, # Solid Circle
    "Powerhouse Rescued (Magnitude Only)" = 17, # Solid Triangle
    "Core Trajectory (Shape Only)"        = 16, # Softer Circle
    "Filtered Out (Fails Both)"           = 4   # Small Cross
  )) +
  scale_size_manual(values = c(
    "Dual Champion (Shape & Magnitude)"   = 1.8,
    "Powerhouse Rescued (Magnitude Only)" = 2.0, # Slightly bigger to pull focus
    "Core Trajectory (Shape Only)"        = 1.2,
    "Filtered Out (Fails Both)"           = 0.8
  )) +
  
  theme_minimal(base_size = 12) +
  labs(
    title = "Plot 4: Hybrid Strategy Grouping (K = 7)",
    subtitle = "Triangles highlight the high-magnitude genes rescued by our LFC gate",
    x = "Log2 Fold Change at 06h (MAP Shrunken)", 
    y = "Log2 Fold Change at 24h (MAP Shrunken)",
    shape = "Inclusion Status"
  )

p4_interactive <- ggplotly(p4_static, tooltip = "text") %>% 
  layout(yaxis = list(scaleanchor = "x", scaleratio = 1))

# Render the hybrid landscape
p4_interactive
```


```{r}
library(tidyverse)

# ==========================================
# 1. RESTRUCTURE HYBRID DATA FOR TIME PROFILE
# ==========================================
df_profiles_hybrid <- df_hybrid_plot %>%
  select(gene_id, cluster, hybrid_status, model_LFC_06h, model_LFC_24h) %>%
  pivot_longer(
    cols = starts_with("model_LFC"),
    names_to = "Timepoint",
    values_to = "LFC"
  ) %>%
  mutate(Timepoint = factor(Timepoint, 
                            levels = c("model_LFC_06h", "model_LFC_24h"),
                            labels = c("06h", "24h")))

# Filter out the genes that failed both thresholds so we only plot active selections
df_profiles_active <- df_profiles_hybrid %>%
  filter(hybrid_status != "Filtered Out (Fails Both)") %>%
  # Create a simpler grouping for line-styling
  mutate(line_type = if_else(str_detect(hybrid_status, "Powerhouse"), "Rescued Driver", "Core Shape"))

# ==========================================
# 2. GENERATE UPDATED PROFILE PLOTS
# ==========================================
library(tidyverse)

# ==========================================
# GENERATE HIGH-CONTRAST SELECTIVE PROFILE PLOT
# ==========================================
ggplot(df_profiles_active, aes(x = Timepoint, y = LFC, group = gene_id)) +
  # Baseline 0-change reference line
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  # Magnitude gates (dotted references)
  geom_hline(yintercept = c(-1.5, 1.5), linetype = "dotted", color = "gray70") +
  
  # 1. DRAW BACKGROUND LINES IN SOFT GRAY
  # This keeps the overall cluster "wave" visible without competing for color
  geom_line(data = filter(df_profiles_active, line_type == "Core Shape"), 
            color = "gray80", alpha = 0.2, size = 0.4) +
  
  # 2. OVERLAY THE RESCUED POWERHOUSES IN SOLID CLUSTER COLORS
  # These will slice cleanly through the gray background waves
  geom_line(data = filter(df_profiles_active, line_type == "Rescued Driver"), 
            aes(color = cluster), alpha = 0.9, size = 1.2) +
  
  # Split cleanly by your 7 biological waves
  facet_wrap(~cluster, labeller = label_both, ncol = 4) +
  
  # Map the solid colors exclusively to the rescued lines
  scale_color_manual(values = base_colors) +
  
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none", # Cluster names are already on the facet headers
    strip.background = element_rect(fill = "gray95", color = NA),
    panel.spacing = unit(1.1, "lines"),
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Rescued Driver Kinetics Across Clusters",
    subtitle = "Vibrant lines show high-magnitude drivers (|LFC| > 1.5) slicing through gray cluster backgrounds",
    x = "Post-Exposure Timepoint",
    y = "Log2 Fold Change vs Matched Control"
  )
```

split out the rescue genes with steeper angles vs the inverse/counter regulation

```{r}
library(tidyverse)

# ==========================================
# 1. CALCULATE EXPECTED CLUSTER TRAJECTORIES
# ==========================================
# We find the average direction of each cluster at 24h to serve as our compass
cluster_directions <- df_SHRUNK_FINAL %>%
  group_by(cluster) %>%
  summarize(
    cluster_mean_24h = mean(model_LFC_24h, na.rm = TRUE),
    .groups = 'drop'
  )

# ==========================================
# 2. BUILD THE FINAL ANNOTATED MASTER DATASET
# ==========================================
df_FINAL_ANNOTATED <- df_SHRUNK_FINAL %>%
  inner_join(cluster_directions, by = "cluster") %>%
  mutate(
    # Recalculate basic flags directly to ensure no missing object errors
    is_powerhouse = (abs(model_LFC_06h) > 1.5 | abs(model_LFC_24h) > 1.5),
    is_shape_core = (membership_score > 0.25),
    
    # Classify the exact behavior of every gene
    driver_behavior = case_when(
      # If it passes the fuzzy math threshold, it's a standard core member
      is_shape_core ~ "Standard Core",
      
      # If it failed the fuzzy math but passed our magnitude gate, check its direction:
      # If its 24h LFC has a different sign than the cluster average, it's an inversion!
      is_powerhouse & (sign(model_LFC_24h) != sign(cluster_mean_24h)) ~ "Inverse Driver",
      
      # If it has the same sign but failed fuzzy math, it's just a hyper-amplified outlier
      is_powerhouse ~ "Amplified Driver",
      
      # Everything else falls below both bars
      TRUE ~ "Filtered Out"
    )
  )

# ==========================================
# 3. PRINT DIAGNOSTIC BREAKDOWN REPORT & LEDGER
# ==========================================
# A. Active Breakdown (Excluding Filtered Out)
df_active_summary <- df_FINAL_ANNOTATED %>% 
  filter(driver_behavior != "Filtered Out")

message("--- Final Cleaned Breakdown of Your 7 Waves ---")
print(table(df_active_summary$cluster, df_active_summary$driver_behavior))

# B. Complete Membership Balance Sheet (Including Filtered Out Columns)
initial_assignments <- df_FINAL_ANNOTATED %>%
  count(cluster, name = "Initially_Assigned")

behavior_counts <- df_FINAL_ANNOTATED %>%
  count(cluster, driver_behavior) %>%
  pivot_wider(
    names_from = driver_behavior, 
    values_from = n, 
    values_fill = 0
  )

final_balance_ledger <- initial_assignments %>%
  inner_join(behavior_counts, by = "cluster") %>%
  # Explicitly order the columns to trace from Initial -> Core/Drivers -> Filtered Out
  select(
    cluster, 
    Initially_Assigned, 
    `Standard Core`, 
    `Amplified Driver`, 
    `Inverse Driver`, 
    `Filtered Out`
  )

message("\n--- Ultimate Cluster Membership Balance Sheet ---")
print(final_balance_ledger)
```


```{r}
library(tidyverse)
library(clusterProfiler)

# ==========================================
# 1. SETUP KO-SPECIFIC LOOKUPS AND UNIVERSE
# ==========================================
# Explicitly loading KO definitions
ko_codes <- read_tsv("/mnt/work-drive/databases/COG-KEGG-data/KO-definitions.tsv", show_col_types = FALSE)

if ("locus_tag" %in% colnames(anno_operon) && !("Gene" %in% colnames(anno_operon))) {
  anno_operon_clean <- anno_operon %>% rename(Gene = locus_tag)
} else {
  anno_operon_clean <- anno_operon
}

# Building strict KO term-to-gene mapping table
ko_TERM2GENE <- anno_operon_clean %>%
  select(Gene, KO) %>%
  filter(!is.na(KO), KO != "") %>%
  separate_rows(KO, sep = ",") %>%
  mutate(KO = trimws(KO)) %>%
  select(term = KO, gene = Gene)

ko_TERM2NAME <- ko_codes %>% 
  transmute(term = KO, name = kDefinition)

ko_enrichment_universe <- as.character(treatment_universe)

# ==========================================
# 2. RUN STATISTICAL KO PATHWAY ENRICHMENT (POOLED)
# ==========================================
ko_combined_enrichment_results <- lapply(1:7, function(i) {
  pooled_genes <- df_FINAL_ANNOTATED %>%
    filter(cluster == i, driver_behavior %in% c("Standard Core", "Amplified Driver")) %>%
    pull(gene_id)
  
  res <- enricher(
    gene          = pooled_genes,
    universe      = ko_enrichment_universe,
    pvalueCutoff  = 0.05,
    pAdjustMethod = "BH",
    TERM2GENE     = ko_TERM2GENE,
    TERM2NAME     = ko_TERM2NAME
  )
  
  if (!is.null(res) && nrow(res) > 0) {
    df <- as.data.frame(res)
    df$Cluster <- paste0("Cluster_", i)
    return(df)
  }
  return(NULL)
})

# Final consolidated KO statistical dataframe
all_ko_enriched_pathways <- bind_rows(ko_combined_enrichment_results)

# ==========================================
# 3. DIRECT KO FUNCTIONAL TRIAGE FOR OUTLIERS
# ==========================================
message("\n=== DIRECT KO ANNOTATION TRIAGE: ALL HIGH-MAGNITUDE OUTLIERS ===")

ko_outlier_manifest <- df_FINAL_ANNOTATED %>%
  filter(driver_behavior %in% c("Amplified Driver", "Inverse Driver")) %>%
  select(gene_id, cluster, driver_behavior, model_LFC_06h, model_LFC_24h) %>%
  left_join(ko_TERM2GENE, by = c("gene_id" = "gene")) %>%
  left_join(ko_TERM2NAME, by = "term") %>%
  arrange(cluster, driver_behavior)

# View the curated KO outlier list
print(ko_outlier_manifest, n = Inf)
```

```{r}
library(tidyverse)
library(clusterProfiler)

# ==========================================
# 1. PREPARE THE KEGG/KO LOOKUP TABLES
# ==========================================
ko_codes <- read_tsv("/mnt/work-drive/databases/COG-KEGG-data/KO-definitions.tsv", show_col_types = FALSE)

# Safety check: Rename locus_tag to Gene if present
if ("locus_tag" %in% colnames(anno_operon) && !("Gene" %in% colnames(anno_operon))) {
  anno_operon_clean <- anno_operon %>% rename(Gene = locus_tag)
} else {
  anno_operon_clean <- anno_operon
}

# Split comma separated lists using the standardized column name
ko_TERM2GENE <- anno_operon_clean %>%
  select(Gene, KO) %>%
  filter(!is.na(KO), KO != "") %>%
  separate_rows(KO, sep = ",") %>%
  mutate(KO = trimws(KO)) %>%
  select(term = KO, gene = Gene)

TERM2NAME <- ko_codes %>%
  transmute(term = KO, name = kDefinition)

# ==========================================
# 2. DEFINE THE STRICT STRUCTURAL UNIVERSE
# ==========================================
enrichment_universe <- as.character(treatment_universe)

# ==========================================
# 3. GENERATE ACTIVE RE-DIRECTIONED LISTS
# ==========================================
hybrid_cluster_lists <- list()

for (i in 1:7) {
  # Wave i: Positive Responders (Core + Amplified)
  pos_genes <- df_FINAL_ANNOTATED %>%
    filter(cluster == i, driver_behavior %in% c("Standard Core", "Amplified Driver")) %>%
    pull(gene_id)
  
  if(length(pos_genes) > 0) {
    hybrid_cluster_lists[[paste0("Cluster_", i, "_Positive")]] <- pos_genes
  }
  
  # Wave i: Inverse Responders (Counter-regulators)
  inv_genes <- df_FINAL_ANNOTATED %>%
    filter(cluster == i, driver_behavior == "Inverse Driver") %>%
    pull(gene_id)
  
  if(length(inv_genes) > 0) {
    hybrid_cluster_lists[[paste0("Cluster_", i, "_Inverse")]] <- inv_genes
  }
}

# ==========================================
# 4. RUN HYBRID PATHWAY ENRICHMENT
# ==========================================
ko_hybrid_enrichment_results <- lapply(names(hybrid_cluster_lists), function(list_name) {
  genes_to_test <- hybrid_cluster_lists[[list_name]]
  
  res <- enricher(
    gene          = genes_to_test,
    universe      = enrichment_universe,
    pvalueCutoff  = 0.05,
    pAdjustMethod = "BH",
    TERM2GENE     = ko_TERM2GENE,
    TERM2NAME     = TERM2NAME
  )
  
  if (!is.null(res) && nrow(res) > 0) {
    df <- as.data.frame(res)
    df$Analysis_Group <- list_name
    return(df)
  }
  return(NULL)
})

# ==========================================
# 5. CONSOLIDATE AND INSPECT
# ==========================================
all_ko_enriched_hybrid <- bind_rows(ko_hybrid_enrichment_results)

head(all_ko_enriched_hybrid)
```







```{r}
library(tidyverse)
library(clusterProfiler)

# ==========================================
# RUN PARALLEL ENRICHMENT (CORES VS. POWERHOUSES)
# ==========================================

# Initialize a clean list to hold all test groups
ko_test_lists <- list()

for (i in 1:7) {
  # Track 1: Clean Wave (Core + Amplified) -> High statistical power
  wave_genes <- df_FINAL_ANNOTATED %>%
    filter(cluster == i, driver_behavior %in% c("Standard Core", "Amplified Driver")) %>%
    pull(gene_id)
  
  if(length(wave_genes) >= 15) {
    ko_test_lists[[paste0("Cluster_", i, "_Main_Wave")]] <- wave_genes
  }
  
  # Track 2: Absolute Outliers (Amplified + Inverse) -> Captures extreme biology
  outlier_genes <- df_FINAL_ANNOTATED %>%
    filter(cluster == i, driver_behavior %in% c("Amplified Driver", "Inverse Driver")) %>%
    pull(gene_id)
  
  # Give it a lower, safer size threshold since outliers are naturally rare
  if(length(outlier_genes) >= 3) {
    ko_test_lists[[paste0("Cluster_", i, "_Extreme_Outliers")]] <- outlier_genes
  }
}

# Run the enrichment loop across all generated groups
ko_parallel_results <- lapply(names(ko_test_lists), function(group_name) {
  res <- enricher(
    gene          = ko_test_lists[[group_name]],
    universe      = ko_enrichment_universe,
    pvalueCutoff  = 0.05,
    pAdjustMethod = "BH",
    TERM2GENE     = ko_TERM2GENE,
    TERM2NAME     = ko_TERM2NAME
  )
  
  if (!is.null(res) && nrow(res) > 0) {
    df <- as.data.frame(res)
    df$Analysis_Group <- group_name
    return(df)
  }
  return(NULL)
})

# Combine all parallel tracks into your final results table
all_ko_enriched_pathways <- bind_rows(ko_parallel_results)

# Inspect the results to confirm your missing term has returned
filter(all_ko_enriched_pathways, str_detect(Analysis_Group, "Cluster_3")) %>% head()
```

```{r}
library(tidyverse)

# ==========================================
# GENERATE STANDALONE OUTLIER FUNCTIONAL MANIFEST
# ==========================================

df_outlier_functional_manifest <- df_FINAL_ANNOTATED %>%
  # Filter strictly for the standalone, high-magnitude categories
  filter(driver_behavior %in% c("Amplified Driver", "Inverse Driver")) %>%
  
  # Select key identification and kinetic columns
  select(
    gene_id, 
    cluster, 
    driver_behavior, 
    model_LFC_06h, 
    model_LFC_24h
  ) %>%
  
  # Map to every associated KO term (explodes comma-separated genes into rows)
  left_join(ko_TERM2GENE, by = c("gene_id" = "gene")) %>%
  
  # Map KO terms to their human-readable definitions
  left_join(ko_TERM2NAME, by = "term") %>%
  
  # Organize logically by cluster and behavior
  arrange(cluster, driver_behavior, gene_id)

# ==========================================
# EXPORT AND PREVIEW
# ==========================================
# Save as a clean CSV for Excel/supplementary data sharing
write_csv(df_outlier_functional_manifest, "/mnt/work-drive/kira/deseq2New/final-split/standalone_outliers_ko_manifest.csv")

message("SUCCESS: Standalone manifest with expanded KO terms exported to '/mnt/work-drive/kira/deseq2New/final-split/standalone_outliers_ko_manifest.csv'")

# Preview the top of your standalone manifest
head(df_outlier_functional_manifest, n = 15)
```

