---
title: "03a-rnaseq-cluster-enrichment"
author: "Kira Goff"
date: "2026-06-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
# Core Data Science & Wrangling (tidyverse covers dplyr and tidyr)
library(tidyverse)

# Genomics & Bioinformatics
library(DESeq2)
library(Mfuzz)
library(clusterProfiler)

# Clustering & Math/Stats
library(cluster)
library(e1071)

# Visualization & Color Palettes
library(pals)
library(plotly)
library(ggplot2)
library(forcats)
library(enrichplot)

# Web & API Tools
library(httr)
library(jsonlite)

library(pheatmap)
library(dplyr)
library(tidyr)
```


```{r pre-cleanup}
base_dir <- "./"
base_results <- "./results2"
results_dir <- "./results2/04-cluster-enrichment-check-non-auto"
lookup_dir <- "./lookups"

ifelse(!dir.exists(file.path(results_dir)),
  dir.create(file.path(results_dir)),
  "Directory Exists"
)

if ("locus_tag" %in% colnames(anno_operon) && !("Gene" %in% colnames(anno_operon))) {
  anno_operon_clean <- anno_operon %>%
    rename(Gene = locus_tag)
} else {
  anno_operon_clean <- anno_operon
}
```

# Define annotation data structure

This is modular, so select your annotation and lookup of interest
My annotations include KO, GO, COG, COG categories, as well as KEGG modules/pathways/reactions
For KEGG and COG I made lookup tsv files of the categories and their definitions
Update as needed for the information in your anno_operon_clean table


Define databases based on the information in your annotations and structure of your lookup tables
```{r define-databases}
db_configs <- list(
  KO = list(
    db_label = "KO",
    database_tsv_path = file.path(lookup_dir, "KO-definitions.tsv"),
    db_term_col = "KO",
    db_definition_col = "kDefinition"
  ),
  kPathway = list(
    db_label = "KEGG_Pathway",
    database_tsv_path = file.path(lookup_dir, "kegg-pathwaymaps.tsv"),
    db_term_col = "pathway",
    db_definition_col = "kPathDef"
  ),
  kModule = list(
    db_label = "KEGG_Module",
    database_tsv_path = file.path(lookup_dir, "kegg-modules.tsv"),
    db_term_col = "module",
    db_definition_col = "kModDef"
  ),
  kRxn = list(
    db_label = "KEGG_Reaction",
    database_tsv_path = file.path(lookup_dir, "kegg-reactions.tsv"),
    db_term_col = "reaction",
    db_definition_col = "kReactionDef"
  ),
  COG = list(
    db_label = "COG",
    database_tsv_path = file.path(lookup_dir, "cog-def.tsv"),
    db_term_col = "COG",
    db_definition_col = "cDef"
  ),
  COGcat = list(
    db_label = "COG_cat",
    database_tsv_path = file.path(lookup_dir, "cog_codes.tsv"),
    db_term_col = "COG_code",
    db_definition_col = "cCodeDef"
  )
)
```


# Enrichment analysis
Select an annotation/database defined above, and your output thresholds.
This saves the data in three ways: by enrichment term, by gene, and as a +/- matrix.

```{r cluster-enrichment}
# ----- Definitions -----
target_db <- "COGcat" # Options: "KO", "kPathway", "kModule", "kRxn", "COG", "COGcat"

# Threshold variables for output table
qtable <- 0.2 # q threshold for out table

clust_thresh <- 0.25 # Baseline cluster membership threshold
lfc_rescue <- 1 # Log2 Fold Change fallback rescue threshold

# subfolder for reshaped outputs
subfolder_dir <- file.path(results_dir, "reshaped_outputs")
if (!dir.exists(subfolder_dir)) dir.create(subfolder_dir, recursive = TRUE)

# keywords to filter out
housekeeping_terms <- c(
  "\\b5s\\b", "\\b16s\\b", "\\b23s\\b", # Structural RNAs (exact word match)
  "rrna", # Pure structural/transport RNAs
  "ribosomal protein", # Catch the structural proteins only
  "^ribosomal large subunit$", # Strips if it's JUST the structural label
  "^ribosomal small subunit$",
  "sigma factor", "rna polymerase" # Transcription core
)

# Dynamically assign the four variables based on your target selection
db_label <- db_configs[[target_db]]$db_label
database_tsv_path <- db_configs[[target_db]]$database_tsv_path
db_term_col <- db_configs[[target_db]]$db_term_col
db_definition_col <- db_configs[[target_db]]$db_definition_col

# Load term definitions
clean_defs <- read.delim(database_tsv_path, sep = "\t", stringsAsFactors = FALSE) %>%
  select(ID = !!sym(db_term_col), Description = !!sym(db_definition_col)) %>%
  distinct()

# ----- Define universe -----
# Get detectable background from DESeq2
raw_universe_genes <- rownames(dds)

# Split multi-term genes to one per row
t2g_base <- anno_operon_clean %>%
  filter(Gene %in% raw_universe_genes) %>%
  select(!!sym(db_label), Gene) %>%
  drop_na(!!sym(db_label))

if (db_label == "COG_cat") {
  t2g_split <- t2g_base %>%
    mutate(!!sym(db_label) := gsub("[ ,]", "", !!sym(db_label))) %>%
    separate_rows(!!sym(db_label), sep = "")
} else {
  t2g_split <- t2g_base %>%
    separate_rows(!!sym(db_label), sep = ",\\s*")

  # For pathways, keep ONLY rows starting exactly with 'map' followed by digits
  if (target_db == "kPathway") {
    t2g_split <- t2g_split %>%
      filter(grepl("^map\\d+", !!sym(db_label)))
  }
}

# Filter out split gene-term pairs that match our housekeeping terms
t2g <- t2g_split %>%
  rename(ID = !!sym(db_label)) %>%
  left_join(clean_defs, by = "ID") %>%
  # Remove only the specific rows that match housekeeping terms
  filter(!grepl(paste(housekeeping_terms, collapse = "|"),
    Description,
    ignore.case = TRUE
  )) %>%
  select(term_id = ID, gene_id = Gene) %>%
  filter(term_id != "" & term_id != "NA") %>%
  distinct()

# Our cosmos
universe_vector <- unique(t2g$gene_id)

# ----- Cluster enrichment -----

# Import clusters
cluster_input <- df_final_clusters %>%
  filter(is_driver == "High-Confidence") %>% # Dynamically honors score + LFC rescues
  filter(Gene %in% universe_vector) %>%
  select(Gene, cluster)

n_clusters <- length(unique(cluster_input$cluster))

# Run cluster enrichment analysis
all_clusters <- sort(unique(cluster_input$cluster))
enrichment_results_list <- list()

for (clust_id in all_clusters) {
  current_cluster_genes <- cluster_input %>%
    filter(cluster == clust_id) %>%
    pull(Gene)

  if (length(intersect(current_cluster_genes, universe_vector)) > 0) {
    ego <- enricher(
      gene         = current_cluster_genes,
      universe     = universe_vector,
      TERM2GENE    = t2g,
      pvalueCutoff = 1,
      qvalueCutoff = qtable,
      minGSSize    = 2,
      maxGSSize    = 500
    )

    if (!is.null(ego) && nrow(ego@result) > 0) {
      res_df <- as.data.frame(ego) %>%
        mutate(Analysis_Group = paste0("Cluster_", clust_id))
      enrichment_results_list[[as.character(clust_id)]] <- res_df
    }
  }
}

# ----- Clean and save data-----

if (length(enrichment_results_list) > 0) {
  # generate the master base table with correct column order
  final_enrichment_df <- bind_rows(enrichment_results_list) %>%
    select(-any_of("Description")) %>%
    left_join(clean_defs, by = "ID") %>%
    select(Analysis_Group, ID, Description, Count, everything())

  # --- Updated Filename Pattern (Sorted by Database Prefix) ---
  # Starts with db_label so file structures group naturally by annotation source
  base_name <- sprintf("%s_enrichment_m%.2f-resc%.1f_c%d", db_label, clust_thresh, lfc_rescue, n_clusters)

  # Organized by enrichment term
  original_out_path <- file.path(results_dir, paste0(base_name, ".tsv"))
  write.table(final_enrichment_df, original_out_path, sep = "\t", row.names = FALSE, quote = FALSE)

  # Organized by cluster, then gene
  tidy_long_df <- final_enrichment_df %>%
    separate_rows(geneID, sep = "/") %>%
    rename(Gene_ID = geneID) %>%
    select(Analysis_Group, Gene_ID, ID, Description, everything()) %>%
    arrange(Analysis_Group, Gene_ID, qvalue)

  long_out_path <- file.path(subfolder_dir, paste0(base_name, "_long_by_gene.tsv"))
  write.table(tidy_long_df, long_out_path, sep = "\t", row.names = FALSE, quote = FALSE)

  # binary presence/absence matrix
  matrix_wide_df <- final_enrichment_df %>%
    mutate(Enriched = 1) %>%
    select(ID, Description, Analysis_Group, Enriched) %>%
    distinct() %>%
    pivot_wider(
      names_from = Analysis_Group,
      values_from = Enriched,
      values_fill = 0
    ) %>%
    arrange(ID)

  wide_out_path <- file.path(subfolder_dir, paste0(base_name, "_cluster_profile_matrix.tsv"))
  write.table(matrix_wide_df, wide_out_path, sep = "\t", row.names = FALSE, quote = FALSE)

  # Console feedback
  message("Success! Exported three file variations grouped by database name:")
  message(sprintf("  -> Original Table: %s", basename(original_out_path)))
  message(sprintf("  -> Long Tidy (By Gene): %s", file.path("reshaped_outputs", basename(long_out_path))))
  message(sprintf("  -> Wide Profile Matrix: %s", file.path("reshaped_outputs", basename(wide_out_path))))
} else {
  message(sprintf("No enrichment found for %s, %d clusters with current parameters.", db_label, n_clusters))
}
```


# Enrichment visualization
This is optional, but useful
These enrichment chunks pull from whatever the last database run was
And dynamically names the output

```{r basic-dotplot, fig.width=8, fig.height=10}
if (exists("final_enrichment_df") && nrow(final_enrichment_df) > 0) {
  # to keep things readable, let's use the 50 most significant terms
  # adjust 'max_terms' as needed
  max_terms <- 50

  # ----- filter and prepare data -----
  plot_data <- final_enrichment_df %>%
    group_by(Analysis_Group) %>%
    slice_min(order_by = qvalue, n = 15, with_ties = FALSE) %>% # Top 15 per cluster
    ungroup() %>%
    distinct(ID, Description, Analysis_Group, qvalue, Count) %>%
    # create a clean label combining ID and Description
    mutate(Term_Label = sprintf("%s: %s", ID, substr(Description, 1, 50))) %>%
    # order terms by their significance so the plot looks structured
    mutate(Term_Label = fct_reorder(Term_Label, qvalue, .desc = TRUE))

  if (nrow(plot_data) > max_terms) {
    # If still too many terms globally, truncate to the top overall
    top_terms <- plot_data %>%
      group_by(Term_Label) %>%
      summarise(min_p = min(qvalue)) %>%
      slice_min(order_by = min_p, n = max_terms, with_ties = FALSE) %>%
      pull(Term_Label)

    plot_data <- plot_data %>% filter(Term_Label %in% top_terms)
  }

  # ----- build matrix for dotplot -----

  # shows which clusters share pathways, scaled by gene count and colored by p-value
  enrich_plot <- ggplot(plot_data, aes(x = Analysis_Group, y = Term_Label)) +
    geom_point(aes(size = Count, color = -log10(qvalue))) +
    scale_color_viridis_c(name = "-log10(qvalue)", option = "magma", direction = 1) +
    scale_size_continuous(name = "Gene Count", range = c(3, 8)) +
    theme_bw(base_size = 11) +
    labs(
      title = sprintf("Functional Enrichment Profile (%s)", target_db),
      subtitle = "Top enriched terms mapped across analysis groups",
      x = "Analysis Cluster",
      y = "Enriched Functional Term"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
      axis.text.y = element_text(size = 9),
      panel.grid.major = element_line(color = "gray95"),
      plot.title = element_text(face = "bold", size = 14),
      legend.position = "right"
    )

  # ----- dynamic save (Updated for Database Sorting & Rescue Tracking) -----
  # 1. Corrected 's' to '%s' so the database string evaluates
  # 2. Included '-resc%.1f' to match your main enrichment table outputs
  plot_out_name <- sprintf("%s_enrichment_m%.2f-resc%.1f_c%d_dotplot.png", db_label, clust_thresh, lfc_rescue, n_clusters)
  plot_out_path <- file.path(results_dir, plot_out_name)

  # adjust height based on the number of unique terms to prevent crowding
  dynamic_height <- max(5, length(unique(plot_data$Term_Label)) * 0.25)

  ggsave(
    filename = plot_out_path,
    plot = enrich_plot,
    width = 9,
    height = dynamic_height,
    dpi = 300,
    bg = "white"
  )

  message(sprintf("Visualization Success! Plot saved to: %s", file.path("reshaped_outputs", plot_out_name)))

  # Render the plot in RStudio console/notebook
  print(enrich_plot)
} else {
  message("Skipping visualization: No enrichment data available to plot.") # oopsie daisie message
}
```



```{r hierarchical-dotplot, fig.width=9, fig.height=11}
if (exists("final_enrichment_df") && nrow(final_enrichment_df) > 0) {
  # ----- filter and prepare data -----
  # Generate matrix
  matrix_prep <- final_enrichment_df %>%
    mutate(Signal = -log10(qvalue)) %>%
    mutate(Row_Label = sprintf("%s: %s", ID, substr(Description, 1, 40))) %>%
    select(Row_Label, Analysis_Group, Signal) %>%
    distinct() %>%
    pivot_wider(names_from = Analysis_Group, values_from = Signal, values_fill = 0)

  # Filter down to top 50 rows based on variance to keep the dot plot clean
  if (nrow(matrix_prep) > 50) {
    row_vars <- apply(as.matrix(matrix_prep[, -1]), 1, var)
    matrix_prep <- matrix_prep[order(-row_vars)[1:50], ]
  }

  # ----- generate clustering matrix -----
  # convert to pure matrix format for clustering calculations
  clust_matrix <- as.matrix(matrix_prep[, -1])
  rownames(clust_matrix) <- matrix_prep$Row_Label

  # extract hierarchical order for pathways (rows)
  row_hc <- hclust(dist(clust_matrix, method = "euclidean"), method = "complete")
  ordered_pathways <- rownames(clust_matrix)[row_hc$order]

  # extract hierarchical order for clusters (columns/analysis groups)
  col_hc <- hclust(dist(t(clust_matrix), method = "euclidean"), method = "complete")
  ordered_clusters <- colnames(clust_matrix)[col_hc$order]

  # ------ reorder based on hierarchical clustering -----
  # Anchor this to final_enrichment_df and save it to plot_data_clustered
  plot_data_clustered <- final_enrichment_df %>%
    mutate(Row_Label = sprintf("%s: %s", ID, substr(Description, 1, 40))) %>%
    # Filter for only the top 50 rows selected by our variance step
    filter(Row_Label %in% ordered_pathways) %>%
    # Convert axes to factors and explicitly enforce the hclust dendrogram levels
    mutate(
      Analysis_Group = factor(Analysis_Group, levels = ordered_clusters),
      Row_Label      = factor(Row_Label, levels = ordered_pathways)
    )

  # ----- generate ordered dotplot -----
  hier_dotplot <- ggplot(plot_data_clustered, aes(x = Analysis_Group, y = Row_Label)) +
    geom_point(aes(size = Count, color = -log10(qvalue))) +
    scale_color_viridis_c(name = "-log10(qvalue)", option = "magma", direction = 1) +
    scale_size_continuous(name = "Gene Count", range = c(3, 8)) +
    theme_bw(base_size = 11) +
    labs(
      title = sprintf("Hierarchical Dot Plot Profile (%s)", target_db),
      subtitle = "Axes arranged dynamically by profile pattern similarity",
      x = "Analysis Clusters (Clustered)",
      y = "Enriched Functional Terms (Clustered)"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
      axis.text.y = element_text(size = 9),
      panel.grid.major = element_line(color = "gray95"),
      plot.title = element_text(face = "bold", size = 14)
    )

  # ----- dynamic save (Updated for Database Sorting & Rescue Tracking) -----
  # Formatted with db_label as prefix and includes rescue metrics for file stability
  dot_out_name <- sprintf("%s_enrichment_m%.2f-resc%.1f_c%d_hierarchical_dotplot.png", db_label, clust_thresh, lfc_rescue, n_clusters)
  dot_out_path <- file.path(results_dir, dot_out_name)

  dynamic_height <- max(5, length(ordered_pathways) * 0.22)

  ggsave(
    filename = dot_out_path,
    plot = hier_dotplot,
    width = 9,
    height = dynamic_height,
    dpi = 300,
    bg = "white"
  )

  message(sprintf("Hierarchical Dot Plot Success! Saved to: %s", file.path(dot_out_name)))
  print(hier_dotplot)
} else {
  message("Skipping Visualization: No matching enrichment data available.") # oopsie daisie message
}
```

```{r hierarchical-heatmap-presence-absense, fig.width=7, fig.height=9}
# ----- hierarchically clustered presence/absence heatmap -----

if (exists("matrix_wide_df") && nrow(matrix_wide_df) > 0) {
  # ----- filter and prepare data -----
  # prepare a numeric matrix for clustering using the wide matrix
  heatmap_matrix <- matrix_wide_df %>%
    # Use the pathway description as the row name so it shows up on the plot
    mutate(Row_Label = sprintf("%s: %s", ID, substr(Description, 1, 40))) %>%
    select(-ID, -Description) %>%
    # Ensure there are no duplicate row names
    distinct(Row_Label, .keep_all = TRUE) %>%
    tibble::column_to_rownames("Row_Label") %>%
    as.matrix()

  # Only plot terms that are active in at least one cluster
  heatmap_matrix <- heatmap_matrix[rowSums(heatmap_matrix) > 0, , drop = FALSE]

  # Cap the number of rows if it's too high (usually 50-60)
  if (nrow(heatmap_matrix) > 60) {
    message("Notice: Truncating heatmap to the top 60 terms with the most cluster overlaps for readability.")
    term_variance <- rowSums(heatmap_matrix)
    heatmap_matrix <- heatmap_matrix[order(-term_variance)[1:60], , drop = FALSE]
  }

  # Define output path
  heatmap_out_name <- sprintf("enrichment_heatmap_%s_m%.2f.pdf", db_label, clust_thresh)
  heatmap_out_path <- file.path(results_dir, "reshaped_outputs", heatmap_out_name)


  # 4. Draw and Save the Clustered Heatmap
  # We use a PDF output here because text labels scale perfectly without pixelating
  # can also use the pass to ggsave method
  pdf(file = heatmap_out_path, width = 8, height = max(6, nrow(heatmap_matrix) * 0.22))

  pheat <- pheatmap(
    mat = heatmap_matrix,
    color = c("#f0f0f0", "#2b8cbe"), # Light gray for 0 (Absence), Deep blue for 1 (Enriched)
    legend_breaks = c(0, 1),
    legend_labels = c("Not Enriched", "Enriched"),
    cluster_rows = TRUE, # Dynamically cluster pathways by profile similarity
    cluster_cols = TRUE, # Dynamically cluster analysis groups by pathway profile
    clustering_distance_rows = "binary", # 'binary' distance is mathematically best for 1/0 matrices
    clustering_distance_cols = "binary",
    clustering_method = "complete",
    border_color = "white",
    fontsize_row = 8,
    fontsize_col = 10,
    main = sprintf("Hierarchical Profiles: %s Enrichment", target_db)
  )

  dev.off() # Close the file stream

  message(sprintf("Heatmap Success! Clustered architecture saved to: %s", file.path("reshaped_outputs", heatmap_out_name)))
} else {
  message("Skipping Heatmap: Wide profile matrix data is missing or empty.") # oopsie daisie message
}
print(pheat)
```

```{r clustered-heatmap-continuous, fig.width=8, fig.height=10}
# ----- hierarchically clustered heatmap with continuous dynamic range -----

if (exists("final_enrichment_df") && nrow(final_enrichment_df) > 0) {
  # 1. Transform the dynamic bubble-plot data into a continuous numeric matrix
  continuous_matrix <- final_enrichment_df %>%
    # Calculate the strength of enrichment (-log10 of adjusted p-value)
    mutate(Signal_Strength = -log10(qvalue)) %>%
    # Truncate long descriptions so the plot labels look clean
    mutate(Row_Label = sprintf("%s: %s", ID, substr(Description, 1, 40))) %>%
    select(Row_Label, Analysis_Group, Signal_Strength) %>%
    distinct() %>%
    # Pivot out clusters into columns, filling missing combinations with 0 (no signal)
    pivot_wider(
      names_from = Analysis_Group,
      values_from = Signal_Strength,
      values_fill = 0
    ) %>%
    tibble::column_to_rownames("Row_Label") %>%
    as.matrix()

  # 2. Filter down to the most informative terms to avoid visual crowding
  # We select the top 50 terms that show the highest variance across your clusters
  if (nrow(continuous_matrix) > 50) {
    message("Notice: Subsetting heatmap to the top 50 terms with the highest profile variance.")
    row_variances <- apply(continuous_matrix, 1, var)
    continuous_matrix <- continuous_matrix[order(-row_variances)[1:50], , drop = FALSE]
  }

  # 3. Define output path
  heatmap_continuous_name <- sprintf("enrichment_continuous_heatmap_%s_m%.2f.pdf", db_label, clust_thresh)
  heatmap_continuous_path <- file.path(results_dir, "reshaped_outputs", heatmap_continuous_name)

  # 4. Generate the plot
  pdf(file = heatmap_continuous_path, width = 8, height = max(6, nrow(continuous_matrix) * 0.22))

  p_cont <- pheatmap(
    mat = continuous_matrix,
    # Using a continuous color gradient (Light gray -> Orange -> Deep Purple)
    color = colorRampPalette(c("#f0f0f0", "#fe9929", "#800026"))(50),
    cluster_rows = TRUE,
    cluster_cols = TRUE,
    clustering_distance_rows = "euclidean", # Euclidean is ideal for continuous gradients!
    clustering_distance_cols = "euclidean",
    clustering_method = "complete",
    border_color = "white",
    fontsize_row = 8,
    fontsize_col = 10,
    main = sprintf("Clustered Enrichment Intensity [-log10(p.adj)]: %s", target_db)
  )

  dev.off()

  message(sprintf("Continuous Heatmap Success! Saved to: %s", file.path("reshaped_outputs", heatmap_continuous_name)))
} else {
  message("Skipping Continuous Heatmap: No data available.") # oopsie daisie message
}
print(p_cont)
```

Let's try the heatmaps in complexheatmap with seriation
# TO DO: fix clustering implimentaton 

```{r}
#install.packages("seriation")

library(ComplexHeatmap)
library(seriation)
library(vegan)
```


```{r olo}
# ----- hierarchically clustered heatmap with continuous dynamic range -----

if (exists("final_enrichment_df") && nrow(final_enrichment_df) > 0) {
  
  # 1. Transform the dynamic bubble-plot data into a continuous numeric matrix
  continuous_matrix <- final_enrichment_df %>%
    mutate(Signal_Strength = -log10(qvalue)) %>%
    mutate(Row_Label = sprintf("%s: %s", ID, substr(Description, 1, 40))) %>%
    select(Row_Label, Analysis_Group, Signal_Strength) %>%
    distinct() %>%
    pivot_wider(
      names_from = Analysis_Group,
      values_from = Signal_Strength,
      values_fill = 0
    ) %>%
    tibble::column_to_rownames("Row_Label") %>%
    as.matrix()
  
  # 1. Transform the dynamic bubble-plot data into a continuous numeric matrix
  continuous_matrix <- final_enrichment_df %>%
    mutate(Signal_Strength = -log10(qvalue)) %>%
    mutate(Row_Label = sprintf("%s: %s", ID, substr(Description, 1, 40))) %>%
    select(Row_Label, Analysis_Group, Signal_Strength) %>%
    distinct() %>%
    pivot_wider(
      names_from = Analysis_Group,
      values_from = Signal_Strength,
      values_fill = 0
    ) %>%
    tibble::column_to_rownames("Row_Label") %>%
    as.matrix()

  # --- NEW: Cap extreme outliers to protect the color scale ---
  # Find the 95th (or 99th) percentile value of the non-zero data
  cap_threshold <- quantile(continuous_matrix[continuous_matrix > 0], 0.95, na.rm = TRUE)
  
  # Apply the cap: anything higher than the threshold gets set to the threshold
  continuous_matrix[continuous_matrix > cap_threshold] <- cap_threshold
  message(sprintf("Notice: Capping outlier values at a maximum -log10(p.adj) of %.2f to preserve color dynamic range.", cap_threshold))
  # -------------------------------------------------------------

  # 2. Filter down to the most informative terms to avoid visual crowding
  if (nrow(continuous_matrix) > 50) {
    message("Notice: Subsetting heatmap to the top 50 terms with the highest profile variance.")
    row_variances <- apply(continuous_matrix, 1, var)
    continuous_matrix <- continuous_matrix[order(-row_variances)[1:50], , drop = FALSE]
  }

  # 3. Define output path (PNG extension)
  heatmap_continuous_name <- sprintf("enrichment_continuous_heatmap_%s_m%.2f-bc-olo.png", db_label, clust_thresh)
  heatmap_continuous_path <- file.path(results_dir, "reshaped_outputs", heatmap_continuous_name)

  # 4. Pre-calculate Clustering using Bray-Curtis + OLO
  # Row clustering (vegdist handles the zero-heavy rows beautifully)
  row_dist <- vegan::vegdist(continuous_matrix, method = "bray")
  row_hc   <- hclust(row_dist, method = "complete")
  row_olo  <- reorder(row_hc, row_dist, method = "OLO")
  row_dend <- as.dendrogram(row_olo)

  # Column clustering 
  col_dist <- vegan::vegdist(t(continuous_matrix), method = "bray")
  col_hc   <- hclust(col_dist, method = "complete")
  col_olo  <- reorder(col_hc, col_dist, method = "OLO")
  col_dend <- as.dendrogram(col_olo)

  # Define color mapping explicitly
  col_fun <- circlize::colorRamp2(
    seq(0, max(continuous_matrix, na.rm = TRUE), length.out = 3), 
    c("#f0f0f0", "#fe9929", "#800026")
  )

  # 5. Construct the Heatmap object
  p_cont <- Heatmap(
    matrix = continuous_matrix,
    col = col_fun,
    name = "-log10(p.adj)",
    column_title = sprintf("Clustered Enrichment Intensity [-log10(p.adj)]: %s", target_db),
    cluster_rows = row_dend,
    cluster_columns = col_dend,
    rect_gp = gpar(col = "white", lwd = 1), 
    row_names_gp = gpar(fontsize = 8),      
    column_names_gp = gpar(fontsize = 10),  
    row_dend_width = unit(15, "mm"),
    column_dend_height = unit(15, "mm")
  )

  # 6. Open a high-resolution PNG device and write the plot
  calc_height <- max(6, nrow(continuous_matrix) * 0.22)
  
  png(filename = heatmap_continuous_path, 
      width = 8, 
      height = calc_height, 
      units = "in", 
      res = 300) # Crisp, publication-quality 300 DPI
  
  ComplexHeatmap::draw(p_cont)
  
  dev.off() # Close the PNG file device safely

  message(sprintf("Continuous Heatmap Success! Saved PNG to: %s", file.path("reshaped_outputs", heatmap_continuous_name)))

} else {
  message("Skipping Continuous Heatmap: No data available.") 
}

# This will draw it cleanly inline inside your R Markdown notebook console pane
if (exists("p_cont")) {
  ComplexHeatmap::draw(p_cont)
}
```


```{r}
# ----- hierarchically clustered heatmap with continuous dynamic range -----

if (exists("final_enrichment_df") && nrow(final_enrichment_df) > 0) {
  
  # 1. Transform the dynamic bubble-plot data into a continuous numeric matrix
  continuous_matrix <- final_enrichment_df %>%
    mutate(Signal_Strength = -log10(qvalue)) %>%
    mutate(Row_Label = sprintf("%s: %s", ID, substr(Description, 1, 40))) %>%
    select(Row_Label, Analysis_Group, Signal_Strength) %>%
    distinct() %>%
    pivot_wider(
      names_from = Analysis_Group,
      values_from = Signal_Strength,
      values_fill = 0
    ) %>%
    tibble::column_to_rownames("Row_Label") %>%
    as.matrix()

  # --- NEW: Cap extreme outliers to protect the color scale ---
  cap_threshold <- quantile(continuous_matrix[continuous_matrix > 0], 0.95, na.rm = TRUE)
  continuous_matrix[continuous_matrix > cap_threshold] <- cap_threshold
  message(sprintf("Notice: Capping outlier values at a maximum -log10(p.adj) of %.2f to preserve color dynamic range.", cap_threshold))
  # -------------------------------------------------------------

  # 2. Filter down to the most informative terms to avoid visual crowding
  if (nrow(continuous_matrix) > 50) {
    message("Notice: Subsetting heatmap to the top 50 terms with the highest profile variance.")
    row_variances <- apply(continuous_matrix, 1, var)
    continuous_matrix <- continuous_matrix[order(-row_variances)[1:50], , drop = FALSE]
  }

  # 3. Define output path (PNG extension)
  heatmap_continuous_name <- sprintf("enrichment_continuous_heatmap_%s_m%.2f-bc-mds.png", db_label, clust_thresh)
  heatmap_continuous_path <- file.path(results_dir, "reshaped_outputs", heatmap_continuous_name)

# ----- Advanced Ordination: Bray-Curtis Distance + PCoA (MDS) -----
  
  # 1. Clean the matrix: Remove any rows or columns that are entirely zero
  #    (which cause division-by-zero NaNs in Bray-Curtis calculations)
  valid_rows <- rowSums(continuous_matrix, na.rm = TRUE) > 0
  valid_cols <- colSums(continuous_matrix, na.rm = TRUE) > 0
  
  clean_matrix <- continuous_matrix[valid_rows, valid_cols, drop = FALSE]
  
  # Fail-safe check to ensure we still have data left to plot
  if (nrow(clean_matrix) < 2 || ncol(clean_matrix) < 2) {
    stop("Error: Not enough non-zero data remaining to calculate distances after filtering empty rows/columns.")
  }

  # 2. Calculate Bray-Curtis distance on the clean data
  row_dist <- vegan::vegdist(clean_matrix, method = "bray")
  col_dist <- vegan::vegdist(t(clean_matrix), method = "bray")

  # 3. Run Multidimensional Scaling (PCoA)
  row_mds  <- seriation::seriate(row_dist, method = "MDS")
  col_mds  <- seriation::seriate(col_dist, method = "MDS")

  # 4. Extract the clean, linear trajectory index order
  ordered_pathways <- rownames(clean_matrix)[seriation::get_order(row_mds)]
  ordered_clusters <- colnames(clean_matrix)[seriation::get_order(col_mds)]
  
  # 5. Manually reorder the cleaned matrix using the MDS paths
  continuous_matrix_ordered <- clean_matrix[ordered_pathways, ordered_clusters, drop = FALSE]
  # Define color mapping explicitly
  col_fun <- circlize::colorRamp2(
    seq(0, max(continuous_matrix_ordered, na.rm = TRUE), length.out = 3), 
    c("#f0f0f0", "#fe9929", "#800026")
  )

  # 5. Construct the Heatmap object
  # FIX: Set cluster_rows and cluster_columns to FALSE since we pre-ordered the matrix
  p_cont <- ComplexHeatmap::Heatmap(
    matrix = continuous_matrix_ordered,
    col = col_fun,
    name = "-log10(p.adj)",
    column_title = sprintf("Clustered Enrichment Intensity [-log10(p.adj)]: %s", target_db),
    cluster_rows = FALSE,      
    cluster_columns = FALSE,   
    rect_gp = gpar(col = "white", lwd = 1), 
    row_names_gp = gpar(fontsize = 8),      
    column_names_gp = gpar(fontsize = 10)
  )

  # 6. Open a high-resolution PNG device and write the plot
  calc_height <- max(6, nrow(continuous_matrix_ordered) * 0.22)
  
  png(filename = heatmap_continuous_path, 
      width = 8, 
      height = calc_height, 
      units = "in", 
      res = 300) # Crisp, publication-quality 300 DPI
  
  ComplexHeatmap::draw(p_cont)
  
  dev.off() # Close the PNG file device safely

  message(sprintf("Continuous Heatmap Success! Saved PNG to: %s", file.path("reshaped_outputs", heatmap_continuous_name)))

} else {
  message("Skipping Continuous Heatmap: No data available.") 
}

# Draw cleanly inline inside your R Markdown notebook console pane
if (exists("p_cont")) {
  ComplexHeatmap::draw(p_cont)
}
```


```{r olo}
# ----- hierarchically clustered heatmap with continuous dynamic range -----

if (exists("final_enrichment_df") && nrow(final_enrichment_df) > 0) {
  
  # 1. Transform the dynamic bubble-plot data into a continuous numeric matrix
  continuous_matrix <- final_enrichment_df %>%
    mutate(Signal_Strength = -log10(qvalue)) %>%
    mutate(Row_Label = sprintf("%s: %s", ID, substr(Description, 1, 40))) %>%
    select(Row_Label, Analysis_Group, Signal_Strength) %>%
    distinct() %>%
    pivot_wider(
      names_from = Analysis_Group,
      values_from = Signal_Strength,
      values_fill = 0
    ) %>%
    tibble::column_to_rownames("Row_Label") %>%
    as.matrix()
  
  # 1. Transform the dynamic bubble-plot data into a continuous numeric matrix
  continuous_matrix <- final_enrichment_df %>%
    mutate(Signal_Strength = -log10(qvalue)) %>%
    mutate(Row_Label = sprintf("%s: %s", ID, substr(Description, 1, 40))) %>%
    select(Row_Label, Analysis_Group, Signal_Strength) %>%
    distinct() %>%
    pivot_wider(
      names_from = Analysis_Group,
      values_from = Signal_Strength,
      values_fill = 0
    ) %>%
    tibble::column_to_rownames("Row_Label") %>%
    as.matrix()

  # --- NEW: Cap extreme outliers to protect the color scale ---
  # Find the 95th (or 99th) percentile value of the non-zero data
  cap_threshold <- quantile(continuous_matrix[continuous_matrix > 0], 0.95, na.rm = TRUE)
  
  # Apply the cap: anything higher than the threshold gets set to the threshold
  continuous_matrix[continuous_matrix > cap_threshold] <- cap_threshold
  message(sprintf("Notice: Capping outlier values at a maximum -log10(p.adj) of %.2f to preserve color dynamic range.", cap_threshold))
  # -------------------------------------------------------------

  # 2. Filter down to the most informative terms to avoid visual crowding
  if (nrow(continuous_matrix) > 50) {
    message("Notice: Subsetting heatmap to the top 50 terms with the highest profile variance.")
    row_variances <- apply(continuous_matrix, 1, var)
    continuous_matrix <- continuous_matrix[order(-row_variances)[1:50], , drop = FALSE]
  }

  # 3. Define output path (PNG extension)
  heatmap_continuous_name <- sprintf("enrichment_continuous_heatmap_%s_m%.2f-bc-olo.png", db_label, clust_thresh)
  heatmap_continuous_path <- file.path(results_dir, "reshaped_outputs", heatmap_continuous_name)

  # 4. Pre-calculate Clustering using Bray-Curtis + OLO
  # Row clustering (vegdist handles the zero-heavy rows beautifully)
  row_dist <- vegan::vegdist(continuous_matrix, method = "bray")
  row_hc   <- hclust(row_dist, method = "complete")
  row_olo  <- reorder(row_hc, row_dist, method = "OLO")
  row_dend <- as.dendrogram(row_olo)

  # Column clustering 
  col_dist <- vegan::vegdist(t(continuous_matrix), method = "bray")
  col_hc   <- hclust(col_dist, method = "complete")
  col_olo  <- reorder(col_hc, col_dist, method = "OLO")
  col_dend <- as.dendrogram(col_olo)

  # Define color mapping explicitly
  col_fun <- circlize::colorRamp2(
    seq(0, max(continuous_matrix, na.rm = TRUE), length.out = 3), 
    c("#f0f0f0", "#fe9929", "#800026")
  )

  # 5. Construct the Heatmap object
  p_cont <- Heatmap(
    matrix = continuous_matrix,
    col = col_fun,
    name = "-log10(p.adj)",
    column_title = sprintf("Clustered Enrichment Intensity [-log10(p.adj)]: %s", target_db),
    cluster_rows = row_dend,
    cluster_columns = col_dend,
    rect_gp = gpar(col = "white", lwd = 1), 
    row_names_gp = gpar(fontsize = 8),      
    column_names_gp = gpar(fontsize = 10),  
    row_dend_width = unit(15, "mm"),
    column_dend_height = unit(15, "mm")
  )

  # 6. Open a high-resolution PNG device and write the plot
  calc_height <- max(6, nrow(continuous_matrix) * 0.22)
  
  png(filename = heatmap_continuous_path, 
      width = 8, 
      height = calc_height, 
      units = "in", 
      res = 300) # Crisp, publication-quality 300 DPI
  
  ComplexHeatmap::draw(p_cont)
  
  dev.off() # Close the PNG file device safely

  message(sprintf("Continuous Heatmap Success! Saved PNG to: %s", file.path("reshaped_outputs", heatmap_continuous_name)))

} else {
  message("Skipping Continuous Heatmap: No data available.") 
}

# This will draw it cleanly inline inside your R Markdown notebook console pane
if (exists("p_cont")) {
  ComplexHeatmap::draw(p_cont)
}
```

```{r}
if (exists("final_enrichment_df") && nrow(final_enrichment_df) > 0) {
  

  # ----- filter and prepare data -----
  # Generate matrix
  matrix_prep <- final_enrichment_df %>%
    mutate(Signal = -log10(qvalue)) %>%
    mutate(Row_Label = sprintf("%s: %s", ID, substr(Description, 1, 40))) %>%
    select(Row_Label, Analysis_Group, Signal) %>%
    distinct() %>%
    pivot_wider(names_from = Analysis_Group, values_from = Signal, values_fill = 0)

  # Filter down to top 50 rows based on variance to keep the dot plot clean
  if (nrow(matrix_prep) > 50) {
    row_vars <- apply(as.matrix(matrix_prep[, -1]), 1, var)
    matrix_prep <- matrix_prep[order(-row_vars)[1:50], ]
  }

  # ----- generate clustering matrix & handle outliers -----
  # convert to pure matrix format for clustering calculations
  clust_matrix <- as.matrix(matrix_prep[, -1])
  rownames(clust_matrix) <- matrix_prep$Row_Label

  # NEW: Dynamic 95th percentile outlier cap to preserve the visual color gradient
  cap_val <- quantile(clust_matrix[clust_matrix > 0], 0.95, na.rm = TRUE)
  clust_matrix[clust_matrix > cap_val] <- cap_val
  message(sprintf("Notice: Capping outlier values at a maximum -log10(qvalue) of %.2f for sorting.", cap_val))

# ----- Advanced Ordination: Bray-Curtis Distance + PCoA (MDS) -----

  

  # 1. Calculate Bray-Curtis distance matrices
  row_dist <- vegan::vegdist(clust_matrix, method = "bray")
  col_dist <- vegan::vegdist(t(clust_matrix), method = "bray")

  # 2. Run Multidimensional Scaling (PCoA) to find the primary continuous gradient
  row_mds  <- seriation::seriate(row_dist, method = "MDS")
  col_mds  <- seriation::seriate(col_dist, method = "MDS")

  # 3. Extract the clean, linear trajectory index order
  ordered_pathways <- rownames(clust_matrix)[seriation::get_order(row_mds)]
  ordered_clusters <- colnames(clust_matrix)[seriation::get_order(col_mds)]
  # ----- generate ordered dotplot -----
  hier_dotplot <- ggplot(plot_data_clustered, aes(x = Analysis_Group, y = Row_Label)) +
    # Updated color aesthetic to use our capped 'Plot_Signal' column
    geom_point(aes(size = Count, color = Plot_Signal)) +
    scale_color_viridis_c(
      name = paste0("-log10(qvalue)\n(capped at ", round(cap_val, 1), ")"), 
      option = "magma", 
      direction = 1
    ) +
    scale_size_continuous(name = "Gene Count", range = c(3, 8)) +
    theme_bw(base_size = 11) +
    labs(
      title = sprintf("Hierarchical Dot Plot Profile (%s)", target_db),
      subtitle = "Axes arranged via Bray-Curtis Dissimilarity & OLO Seriation",
      x = "Analysis Clusters (Clustered)",
      y = "Enriched Functional Terms (Clustered)"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
      axis.text.y = element_text(size = 9),
      panel.grid.major = element_line(color = "gray95"),
      plot.title = element_text(face = "bold", size = 14)
    )

  # ----- dynamic save -----
  dot_out_name <- sprintf("%s_enrichment_m%.2f-resc%.1f_c%d_hierarchical_dotplot-bc-OLO.png", db_label, clust_thresh, lfc_rescue, n_clusters)
  dot_out_path <- file.path(results_dir, dot_out_name)

  dynamic_height <- max(5, length(ordered_pathways) * 0.22)

  ggsave(
    filename = dot_out_path,
    plot = hier_dotplot,
    width = 9,
    height = dynamic_height,
    dpi = 300,
    bg = "white"
  )

  message(sprintf("Hierarchical Dot Plot Success! Saved to: %s", file.path(dot_out_name)))
  print(hier_dotplot)
} else {
  message("Skipping Visualization: No matching enrichment data available.") 
}
```


```{r}
if (exists("final_enrichment_df") && nrow(final_enrichment_df) > 0) {
  

  # ----- filter and prepare data -----
  # Generate matrix
  matrix_prep <- final_enrichment_df %>%
    mutate(Signal = -log10(qvalue)) %>%
    mutate(Row_Label = sprintf("%s: %s", ID, substr(Description, 1, 40))) %>%
    select(Row_Label, Analysis_Group, Signal) %>%
    distinct() %>%
    pivot_wider(names_from = Analysis_Group, values_from = Signal, values_fill = 0)

  # Filter down to top 50 rows based on variance to keep the dot plot clean
  if (nrow(matrix_prep) > 50) {
    row_vars <- apply(as.matrix(matrix_prep[, -1]), 1, var)
    matrix_prep <- matrix_prep[order(-row_vars)[1:50], ]
  }

  # ----- generate clustering matrix & handle outliers -----
  # convert to pure matrix format for clustering calculations
  clust_matrix <- as.matrix(matrix_prep[, -1])
  rownames(clust_matrix) <- matrix_prep$Row_Label

  # NEW: Dynamic 95th percentile outlier cap to preserve the visual color gradient
  cap_val <- quantile(clust_matrix[clust_matrix > 0], 0.95, na.rm = TRUE)
  clust_matrix[clust_matrix > cap_val] <- cap_val
  message(sprintf("Notice: Capping outlier values at a maximum -log10(qvalue) of %.2f for sorting.", cap_val))

# ----- Advanced Ordination: Bray-Curtis Distance + PCoA (MDS) -----

  

  # 1. Calculate Bray-Curtis distance matrices
  row_dist <- vegan::vegdist(clust_matrix, method = "bray")
  col_dist <- vegan::vegdist(t(clust_matrix), method = "bray")

  # 2. Run Multidimensional Scaling (PCoA) to find the primary continuous gradient
  row_mds  <- seriation::seriate(row_dist, method = "MDS")
  col_mds  <- seriation::seriate(col_dist, method = "MDS")

  # 3. Extract the clean, linear trajectory index order
  ordered_pathways <- rownames(clust_matrix)[seriation::get_order(row_mds)]
  ordered_clusters <- colnames(clust_matrix)[seriation::get_order(col_mds)]
  # ----- generate ordered dotplot -----
  hier_dotplot <- ggplot(plot_data_clustered, aes(x = Analysis_Group, y = Row_Label)) +
    # Updated color aesthetic to use our capped 'Plot_Signal' column
    geom_point(aes(size = Count, color = Plot_Signal)) +
    scale_color_viridis_c(
      name = paste0("-log10(qvalue)\n(capped at ", round(cap_val, 1), ")"), 
      option = "magma", 
      direction = 1
    ) +
    scale_size_continuous(name = "Gene Count", range = c(3, 8)) +
    theme_bw(base_size = 11) +
    labs(
      title = sprintf("Hierarchical Dot Plot Profile (%s)", target_db),
      subtitle = "Axes arranged via Bray-Curtis Dissimilarity & OLO Seriation",
      x = "Analysis Clusters (Clustered)",
      y = "Enriched Functional Terms (Clustered)"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
      axis.text.y = element_text(size = 9),
      panel.grid.major = element_line(color = "gray95"),
      plot.title = element_text(face = "bold", size = 14)
    )

  # ----- dynamic save -----
  dot_out_name <- sprintf("%s_enrichment_m%.2f-resc%.1f_c%d_hierarchical_dotplot-bc-mds.png", db_label, clust_thresh, lfc_rescue, n_clusters)
  dot_out_path <- file.path(results_dir, dot_out_name)

  dynamic_height <- max(5, length(ordered_pathways) * 0.22)

  ggsave(
    filename = dot_out_path,
    plot = hier_dotplot,
    width = 9,
    height = dynamic_height,
    dpi = 300,
    bg = "white"
  )

  message(sprintf("Hierarchical Dot Plot Success! Saved to: %s", file.path(dot_out_name)))
  print(hier_dotplot)
} else {
  message("Skipping Visualization: No matching enrichment data available.") 
}
```





TO DO:
- add the additional annotation information formats used in the mahal distance