---
title: "deseq2-split3b-GO-clustering"
author: "Kira Goff"
date: "2026-06-05"
output: html_document
---

# R setup

```{r setup, message=FALSE, warning=FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(dplyr)          # Grammar of data manipulation
library(tidyr)          # Data reshaping and pivoting (pivot_wider/expand_grid)
library(stringr)        # Robust regex and text filtering (str_detect)
library(purrr)          # Functional programming and list iteration
library(tidytext)       # Text mining and word tokenization for word clouds


library(topGO)          # Parent enrichment engine (elim/weight algorithms)
library(clusterProfiler) # Functional profile comparison & enrichment workflows
library(AnnotationDbi)  # Core database interface for gene annotations
library(GO.db)          # Metadata mapping for Gene Ontology structural trees


library(ComplexHeatmap) # Bi-clustered matrix heatmaps (Figure 3)
library(circlize)       # Required for ComplexHeatmap precise color mapping (colorRamp2)


library(ggplot2)        # Core plotting framework
library(enrichplot)     # Visualization tools for clusterProfiler objects (cnetplot layout)
library(ggwordcloud)    # Spatial word cloud engine with text anti-collision
library(ggdendro)       # Extracts data trees into standard ggplot segments
library(tidygraph)      # Tidy framework for structural network data (tbl_graph)
library(ggraph)         # Network graph layouts (stress/force-directed structures)
library(patchwork)      # Combines multiple plots into grids (Tree-Heatmap layouts)
```


```{r}
base_dir <- "./"
base_results <- "./results2"
results_dir <- "./results2/04-cluster-enrichment/go-qcheck"


if (!dir.exists(results_dir)) {
  dir.create(results_dir, recursive = TRUE)
}


```


#GO enrichment, specifically, because we have to fetch terms instead of reading from tsv
```{r}
# ----- define variables -----
qtable <- 0.3          # q threshold for out table (Updated to match reference)
clust_thresh <- 0.25   # cluster membership threshold
lfc_rescue <- 1        # Log2 Fold Change fallback rescue threshold

target_db <- "GO_bak"

# Keywords to filter out
housekeeping_terms <- c(
  "\\b5s\\b", "\\b16s\\b", "\\b23s\\b", 
  "rrna", 
  "ribosomal protein", 
  "^ribosomal large subunit$", 
  "^ribosomal small subunit$",
  "sigma factor", "rna polymerase"
)

# Define background from detectable genes
raw_universe_genes <- rownames(dds)

# ----- mapping and universe definitions -----

# 1. Split multi-term genes to one per row
t2g_base <- anno_operon_clean %>%
  filter(Gene %in% raw_universe_genes) %>%
  dplyr::select(!!sym(target_db), Gene) %>%
  drop_na(!!sym(target_db)) %>%
  separate_rows(!!sym(target_db), sep = ",\\s*")

# 2. Extract unique GO terms to fetch descriptions from GO.db for housekeeping filters
unique_raw_gos <- unique(t2g_base[[target_db]])

library(GO.db)
clean_defs <- AnnotationDbi::select(
  x       = GO.db,
  keys    = unique_raw_gos,
  columns = c("GOID", "TERM"),
  keytype = "GOID"
) %>%
  dplyr::rename(ID = GOID, Description = TERM) %>%
  drop_na(Description)

# 3. Filter out gene-term pairs matching housekeeping terms
t2g <- t2g_base %>%
  rename(ID = !!sym(target_db)) %>%
  left_join(clean_defs, by = "ID") %>%
  filter(!grepl(paste(housekeeping_terms, collapse = "|"), Description, ignore.case = TRUE)) %>%
  select(term_id = ID, gene_id = Gene) %>%
  filter(term_id != "" & term_id != "NA") %>%
  distinct()

# Final clean background universe
universe_vector <- unique(t2g$gene_id)

# ----- Cluster processing -----

cluster_input <- df_final_clusters %>%
  filter(is_driver == "High-Confidence") %>% # Honors upstream dual-filtering rules
  filter(Gene %in% universe_vector) %>%
  select(Gene, cluster)
  
n_clusters <- length(unique(cluster_input$cluster))
all_clusters <- sort(unique(cluster_input$cluster))
enrich_list <- list()

message("Starting GO Enrichment for ", length(all_clusters), " clusters...")

for (clust_id in all_clusters) {
  current_genes <- cluster_input %>%
    filter(cluster == clust_id) %>%
    pull(Gene)

  if (length(intersect(current_genes, universe_vector)) > 0) {
    ego <- enricher(
      gene         = current_genes,
      universe     = universe_vector,
      TERM2GENE    = t2g,
      pvalueCutoff = 1,
      qvalueCutoff = qtable,
      minGSSize    = 2,
      maxGSSize    = 500
    )

    if (!is.null(ego) && nrow(ego@result) > 0) {
      enrich_list[[as.character(clust_id)]] <- as.data.frame(ego) %>%
        mutate(Analysis_Group = paste0("Cluster_", clust_id))
    }
  }
}

# ----- Compile Outputs -----

if (length(enrich_list) > 0) {
  all_go_res <- bind_rows(enrich_list)
  unique_go_ids <- unique(all_go_res$ID)

  message("Mapping full metadata for metadata file generation...")

  # Detailed definition lookup for the final output table
  go_metadata_df <- AnnotationDbi::select(
    x       = GO.db,
    keys    = unique_go_ids,
    columns = c("GOID", "TERM", "DEFINITION"),
    keytype = "GOID"
  ) %>%
    dplyr::rename(ID = GOID, GO_Name = TERM, GO_Definition = DEFINITION)

  final_go_df <- all_go_res %>%
    dplyr::left_join(go_metadata_df, by = "ID")

  if (!"geneID" %in% colnames(final_go_df) & "gene_id" %in% colnames(final_go_df)) {
    final_go_df <- final_go_df %>% dplyr::rename(geneID = gene_id)
  }

  final_go_df <- final_go_df %>%
    dplyr::select(
      dplyr::any_of(c(
        "Analysis_Group", "GO_Name", "ID", "GO_Definition",
        "Count", "GeneRatio", "BgRatio", "pvalue",
        "p.adjust", "qvalue", "geneID"
      ))
    )

  # Setup paths matching target database directory
  base_name <- sprintf("%s_enrichment_m%.2f-resc%.1f_c%d", target_db, clust_thresh, lfc_rescue, n_clusters)
  
  if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)
  out_path <- file.path(results_dir, paste0(base_name, "_go_results.csv"))
  
  write.csv(final_go_df, out_path, row.names = FALSE)
  message("GO Analysis complete. File saved to: ", out_path)
  
} else {
  message(sprintf("No enrichment found for %s, %d clusters with current parameters.", target_db, n_clusters))
}

# Namespace safety resetting
select <- dplyr::select
filter <- dplyr::filter
```


```{#r bubble-plot-all-clusters}
# ----- define variables -----
qtable <- 0.3          # q threshold for output filtering

clust_thresh <- 0.25   # cluster membership threshold
target_db <- "GO_bak"

# Define background universe from detectable genes
raw_universe_genes <- rownames(dds)

# ----- mapping -----

# Map background genes to custom GO terms
t2g <- anno_operon_clean %>%
  filter(Gene %in% raw_universe_genes) %>%
  dplyr::select(!!sym(target_db), Gene) %>%
  drop_na(!!sym(target_db)) %>%
  separate_rows(!!sym(target_db), sep = ", ") %>%
  filter(!!sym(target_db) != "" & !!sym(target_db) != "NA") %>%
  rename(term_id = !!sym(target_db), gene_id = Gene) %>%
  distinct()

universe_vector <- unique(t2g$gene_id)

# Build a comprehensive TERM2NAME mapping table from GO.db
message("Building global GO term-to-name lookup table...")
all_go_ids_in_universe <- unique(t2g$term_id)

t2n <- AnnotationDbi::select(
  x       = GO.db,
  keys    = all_go_ids_in_universe,
  columns = c("GOID", "TERM"),
  keytype = "GOID"
) %>%
  dplyr::rename(term_id = GOID, term_name = TERM) %>%
  distinct()

# Filter target driver genes present in our background universe
cluster_input <- df_final_clusters %>%
  filter(is_driver == "High-Confidence" & Gene %in% universe_vector)

# ----- simultaneous cluster enrichment -----
message("Splitting genes into cluster list and initiating compareCluster...")

# Format clusters into a clean, named list of gene vectors to avoid splitting errors
gene_list_by_cluster <- split(cluster_input$Gene, cluster_input$cluster)

# Run comparative enrichment across all groups concurrently
comp_ego <- compareCluster(
  geneClusters = gene_list_by_cluster, 
  fun          = "enricher", 
  universe     = universe_vector,
  TERM2GENE    = t2g,
  TERM2NAME    = t2n,
  pvalueCutoff = 1,
  qvalueCutoff = qtable
)

# ----- CORRECTED ONTOLOGY PATCH ---
# Use attr() since 'ontology' is a core metadata attribute for compareClusterResult
attr(comp_ego, "ontology") <- "BP"

# ----- similarity matrix ----- 
message("Calculating global semantic similarity matrix via Jaccard Coefficient...")

# Calculate overlapping gene similarities across the whole uncollapsed dataset
comp_ego_sim <- pairwise_termsim(comp_ego, method = "JC", semData = NULL)

# ----- Visualization -----
message("Collapsing redundant terms and plotting master comparative dotplot...")

suppressWarnings({
  # showCategory = 5 groups hundreds of terms into modules, displaying the top 5 per cluster
  p_compare <- dotplot(comp_ego_sim, x = "Cluster", showCategory = 5) +
    labs(
      title = "Functional Mapping Across Analysis Groups",
      subtitle = "Redundant terms collapsed dynamically; top modules compared side-by-side",
      x = "Analysis Group (Cluster)",
      y = "Enriched Functional Module"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
      axis.text.y = element_text(fontsize = 9),
      plot.title = element_text(face = "bold", size = 14)
    )

  # Ensure the results folder exists and write the comparative plot
  if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)
  
  out_plot_path <- file.path(results_dir, "all_clusters_compared_dotplot.png")
  ggsave(
    filename = out_plot_path,
    plot = p_compare, 
    width = 15, 
    height = 12, 
    dpi = 300
  )
})


message("Success! Master comparative bubble plot saved to: ", out_plot_path)

print(p_compare)

# ------ reset pointers -----
select <- dplyr::select
filter <- dplyr::filter
message("Namespace reset: 'select' and 'filter' point back to dplyr.")
```


```{r}
# ----- define variables -----
qtable <- 0.3          # q threshold for output filtering
clust_thresh <- 0.25   # cluster membership threshold
lfc_rescue <- 1        # ADDED: explicitly declare rescue metadata token
target_db <- "GO_bak"

# Define background universe from detectable genes
raw_universe_genes <- rownames(dds)

# ----- mapping -----

# Map background genes to custom GO terms (UPDATED REGEX TO MATCH CHUNK 2)
t2g <- anno_operon_clean %>%
  filter(Gene %in% raw_universe_genes) %>%
  dplyr::select(!!sym(target_db), Gene) %>%
  drop_na(!!sym(target_db)) %>%
  separate_rows(!!sym(target_db), sep = ",\\s*") %>% # Fix: space-flexible regex splitting
  filter(!!sym(target_db) != "" & !!sym(target_db) != "NA") %>%
  rename(term_id = !!sym(target_db), gene_id = Gene) %>%
  distinct()

universe_vector <- unique(t2g$gene_id)

# Build a comprehensive TERM2NAME mapping table from GO.db
message("Building global GO term-to-name lookup table...")
all_go_ids_in_universe <- unique(t2g$term_id)

t2n <- AnnotationDbi::select(
  x       = GO.db,
  keys    = all_go_ids_in_universe,
  columns = c("GOID", "TERM"),
  keytype = "GOID"
) %>%
  dplyr::rename(term_id = GOID, term_name = TERM) %>%
  distinct()

# ----- FIXED CLUSTER FILTERING -----
# Ensuring the rescue threshold logic matches your accurate data profile
cluster_input <- df_final_clusters %>%
  filter(is_driver == "High-Confidence" & Gene %in% universe_vector) %>%
  select(Gene, cluster) # Keep it lightweight

# ----- simultaneous cluster enrichment -----
message("Splitting genes into cluster list and initiating compareCluster...")

# Format clusters into a clean, named list of gene vectors to avoid splitting errors
gene_list_by_cluster <- split(cluster_input$Gene, cluster_input$cluster)

# Run comparative enrichment across all groups concurrently
comp_ego <- compareCluster(
  geneClusters = gene_list_by_cluster, 
  fun          = "enricher", 
  universe     = universe_vector,
  TERM2GENE    = t2g,
  TERM2NAME    = t2n,
  pvalueCutoff = 1,
  qvalueCutoff = qtable
)

# ----- CORRECTED ONTOLOGY PATCH ---
# Use attr() since 'ontology' is a core metadata attribute for compareClusterResult
attr(comp_ego, "ontology") <- "BP"

# ----- similarity matrix ----- 
message("Calculating global semantic similarity matrix via Jaccard Coefficient...")

# Calculate overlapping gene similarities across the whole uncollapsed dataset
comp_ego_sim <- pairwise_termsim(comp_ego, method = "JC", semData = NULL)

# ----- Visualization -----
message("Collapsing redundant terms and plotting master comparative dotplot...")

suppressWarnings({
  # showCategory = 5 groups hundreds of terms into modules, displaying the top 5 per cluster
  p_compare <- dotplot(comp_ego_sim, x = "Cluster", showCategory = 5) +
    labs(
      title = "Functional Mapping Across Analysis Groups",
      subtitle = "Redundant terms collapsed dynamically; top modules compared side-by-side",
      x = "Analysis Group (Cluster)",
      y = "Enriched Functional Module"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
      axis.text.y = element_text(fontsize = 9),
      plot.title = element_text(face = "bold", size = 14)
    )

  # Ensure the results folder exists and write the comparative plot
  if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)
  
  out_plot_path <- file.path(results_dir, "all_clusters_compared_dotplot.png")
  ggsave(
    filename = out_plot_path,
    plot = p_compare, 
    width = 15, 
    height = 12, 
    dpi = 300
  )
})


message("Success! Master comparative bubble plot saved to: ", out_plot_path)

print(p_compare)

# ------ reset pointers -----
select <- dplyr::select
filter <- dplyr::filter
message("Namespace reset: 'select' and 'filter' point back to dplyr.")
```



.
Let's look at the drivers per cluster

```{r !!!go-enrichment-semantic-terminology-sep-trees}
# ----- define variables -----
qtable <- 0.3          # q threshold for out table

clust_thresh <- 0.25   # cluster membership threshold
target_db <- "GO_bak"

raw_universe_genes <- rownames(dds)

# ----- mapping -----

t2g <- anno_operon_clean %>%
  filter(Gene %in% raw_universe_genes) %>%
  dplyr::select(!!sym(target_db), Gene) %>%
  drop_na(!!sym(target_db)) %>%
  separate_rows(!!sym(target_db), sep = ", ") %>%
  filter(!!sym(target_db) != "" & !!sym(target_db) != "NA") %>%
  rename(term_id = !!sym(target_db), gene_id = Gene) %>%
  distinct()

universe_vector <- unique(t2g$gene_id)

message("Building GO ID to Term Name mapping table...")
all_go_ids_in_universe <- unique(t2g$term_id)

t2n <- AnnotationDbi::select(
  x       = GO.db,
  keys    = all_go_ids_in_universe,
  columns = c("GOID", "TERM"),
  keytype = "GOID"
) %>%
  dplyr::rename(term_id = GOID, term_name = TERM) %>%
  distinct()

# Prepare cluster inputs
cluster_input <- df_final_clusters %>%
  filter(is_driver == "High-Confidence" & Gene %in% universe_vector)

all_clusters <- sort(unique(cluster_input$cluster))
enrich_list <- list()

message("Starting GO Enrichment for ", length(all_clusters), " clusters...")

# ----- enrichment and visualization ---
for (clust_id in all_clusters) {
  current_genes <- cluster_input %>%
    filter(cluster == clust_id) %>%
    pull(Gene)

  if (length(intersect(current_genes, universe_vector)) > 0) {
    
    # Run natively with standard stable alphanumeric tables
    ego <- enricher(
      gene         = current_genes,
      universe     = universe_vector, 
      TERM2GENE    = t2g,
      pvalueCutoff = 1,
      qvalueCutoff = qtable,
      minGSSize    = 2,
      maxGSSize    = 500
    )

    if (!is.null(ego) && nrow(as.data.frame(ego)) > 0) {
      
      ego@ontology <- "BP" 
      ego_sim <- pairwise_termsim(ego, method = "JC", semData = NULL)
      
      n_terms <- nrow(as.data.frame(ego_sim))
      
      if (n_terms >= 3) {
        target_groups <- if (n_terms < 5) (n_terms - 1) else 5
        
        suppressWarnings({
          # 1. Generate the standard base treeplot
          p_tree <- treeplot(
            ego_sim, 
            showCategory = 40, 
            nCluster     = target_groups
          )
          
          # ====================================================================
          # DYNAMIC DICTIONARY INTERCEPT (The Driver Filtering Fix)
          # ====================================================================
          # Directly query GO.db for whatever specific keys are present in this object
          active_go_ids <- ego_sim@result$ID
          
          local_t2n <- AnnotationDbi::select(
            x       = GO.db,
            keys    = active_go_ids,
            columns = c("GOID", "TERM"),
            keytype = "GOID"
          )
          
          # Build a pristine named lookup map unique to this specific runtime result
          namemap <- setNames(local_t2n$TERM, local_t2n$GOID)
          
          # Swap the y-axis labels using the freshly queried terms
          p_tree <- p_tree + 
            scale_y_discrete(labels = function(x) {
              ifelse(x %in% names(namemap), namemap[x], x)
            }) +
            labs(
              title = paste0("Cluster ", clust_id, " GO Tree (Top 40 Terms)"),
              subtitle = sprintf("Terms grouped by semantic similarity (%d modules colored)", target_groups)
            ) +
            theme(
              axis.text.y = element_text(size = 9, face = "bold") 
            )
          # ====================================================================
          
          if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)
          
          base_plot_name <- sprintf("cluster_%s_%s_m%.2f-resc%.1f_treeplot.png", clust_id, target_db, clust_thresh, lfc_rescue)
          ggsave(
            filename = file.path(results_dir, base_plot_name), 
            plot = p_tree, 
            width = 18,     
            height = 11, 
            dpi = 300
          )
        })
        
        print(p_tree)
      } else {
        message(sprintf("Skipping treeplot for Cluster %s: Insufficient terms (%d found) for semantic clustering.", clust_id, n_terms))
      }
      
      enrich_list[[as.character(clust_id)]] <- as.data.frame(ego) %>%
        mutate(Analysis_Group = paste0("Cluster_", clust_id))
        
    } else {
      message("No significant GO enrichment found passing thresholds for Cluster ", clust_id)
    }
  }
}

# ----- metadata, merge, and output ---
if (length(enrich_list) > 0) {
  all_go_res <- bind_rows(enrich_list)
  unique_go_ids <- unique(all_go_res$ID)

  message("Mapping extra metadata for ", length(unique_go_ids), " unique GO terms...")

  go_metadata_df <- AnnotationDbi::select(
    x       = GO.db,
    keys    = unique_go_ids,
    columns = c("GOID", "TERM", "DEFINITION"),
    keytype = "GOID"
  ) %>%
    dplyr::rename(ID = GOID, GO_Name = TERM, GO_Definition = DEFINITION)

  # Merge metadata with results (keeping columns consistent)
  final_go_df <- all_go_res %>%
    dplyr::left_join(go_metadata_df, by = "ID")

  # Standardize gene column naming conventions across variations of clusterProfiler
  if (!"geneID" %in% colnames(final_go_df) & "gene_id" %in% colnames(final_go_df)) {
    final_go_df <- final_go_df %>% dplyr::rename(geneID = gene_id)
  }

  final_go_df <- final_go_df %>%
    dplyr::select(
      dplyr::any_of(c(
        "Analysis_Group", "GO_Name", "ID", "GO_Definition",
        "Count", "GeneRatio", "BgRatio", "pvalue",
        "p.adjust", "qvalue", "geneID"
      ))
    )

  # Save local CSV
  if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)
  out_path <- file.path(results_dir, "go_anno_cluster_results.csv")
  write.csv(final_go_df, out_path, row.names = FALSE)

  message("GO Analysis complete. File and plots successfully saved with descriptive text names.")
  
  # ----- ADDED PRINT STATEMENT ---
  # Spits out a clean text preview of the dataframe right to the console
  cat("\n--- FINAL RESULTS DATAFRAME PREVIEW ---\n")
  print(tibble::as_tibble(final_go_df), n = 20) # Prints the top 20 rows nicely
  
} else {
  message("Process finished: No significant GO enrichment across any clusters.")
}

# ----- fix pointers ---

select <- dplyr::select
filter <- dplyr::filter

message("Namespace reset: 'select' and 'filter' point back to dplyr.")
```

```{r !!!}
# ----- define variables -----
qtable <- 0.3          # q threshold for out table
clust_thresh <- 0.25   # cluster membership threshold
lfc_rescue <- 1        # ADDED: explicitly declare rescue metadata token
target_db <- "GO_bak"

# Keywords to filter out
housekeeping_terms <- c(
  "\\b5s\\b", "\\b16s\\b", "\\b23s\\b", 
  "rrna", 
  "ribosomal protein", 
  "^ribosomal large subunit$", 
  "^ribosomal small subunit$",
  "sigma factor", "rna polymerase"
)

raw_universe_genes <- rownames(dds)

# ----- mapping and universe definitions -----

# 1. Split multi-term genes using the flexible space regex
t2g_base <- anno_operon_clean %>%
  filter(Gene %in% raw_universe_genes) %>%
  dplyr::select(!!sym(target_db), Gene) %>%
  drop_na(!!sym(target_db)) %>%
  separate_rows(!!sym(target_db), sep = ",\\s*") # Fix: space-flexible splitting

# 2. Extract unique GO terms to fetch descriptions for housekeeping filters
unique_raw_gos <- unique(t2g_base[[target_db]])

library(GO.db)
clean_defs <- AnnotationDbi::select(
  x       = GO.db,
  keys    = unique_raw_gos,
  columns = c("GOID", "TERM"),
  keytype = "GOID"
) %>%
  dplyr::rename(ID = GOID, Description = TERM) %>%
  drop_na(Description)

# 3. Filter out gene-term pairs matching housekeeping terms
t2g <- t2g_base %>%
  rename(ID = !!sym(target_db)) %>%
  left_join(clean_defs, by = "ID") %>%
  filter(!grepl(paste(housekeeping_terms, collapse = "|"), Description, ignore.case = TRUE)) %>%
  select(term_id = ID, gene_id = Gene) %>%
  filter(term_id != "" & term_id != "NA") %>%
  distinct()

universe_vector <- unique(t2g$gene_id)

message("Building GO ID to Term Name mapping table...")
t2n <- clean_defs %>% 
  rename(term_id = ID, term_name = Description) %>%
  filter(term_id %in% universe_vector)

# Prepare cluster inputs matching your correct upstream data profile
cluster_input <- df_final_clusters %>%
  filter(is_driver == "High-Confidence" & Gene %in% universe_vector) %>%
  select(Gene, cluster)

all_clusters <- sort(unique(cluster_input$cluster))
n_clusters <- length(all_clusters)
enrich_list <- list()

message("Starting GO Enrichment for ", n_clusters, " clusters...")

# ----- enrichment and visualization ---
for (clust_id in all_clusters) {
  current_genes <- cluster_input %>%
    filter(cluster == clust_id) %>%
    pull(Gene)

  if (length(intersect(current_genes, universe_vector)) > 0) {
    
    # Run enricher natively with robust alphanumeric IDs
    ego <- enricher(
      gene         = current_genes,
      universe     = universe_vector, 
      TERM2GENE    = t2g,
      TERM2NAME    = t2n,     
      pvalueCutoff = 1,
      qvalueCutoff = qtable,
      minGSSize    = 2,
      maxGSSize    = 500
    )

    # Check if there are valid results before continuing
    if (!is.null(ego) && nrow(as.data.frame(ego)) > 0) {
      
      ego@ontology <- "BP" 
      
      # Step 1: Calculate semantic matrix while slots match perfectly
      ego_sim <- pairwise_termsim(ego, method = "JC", semData = NULL)
      
      # ========================================================================
      # THE POST-STATISTICAL TREEPLOT SLOT INTERCEPT PATCH
      # ========================================================================
      # Build a direct mapping dictionary from the internal alphanumeric keys
      raw_ids <- ego_sim@result$ID
      matched_words <- t2n$term_name[match(raw_ids, t2n$term_id)]
      final_clean_labels <- ifelse(!is.na(matched_words) & matched_words != "", matched_words, raw_ids)
      
      # Overwrite the visual description slots treeplot uses for text generation
      ego_sim@result$Description <- final_clean_labels
      ego_sim@result$ID          <- final_clean_labels
      rownames(ego_sim@result)   <- final_clean_labels
      
      # Update the underlying similarity matrix labels so clustering coordinates align
      if (!is.null(ego_sim@termsim)) {
        rownames(ego_sim@termsim) <- final_clean_labels[match(rownames(ego_sim@termsim), raw_ids)]
        colnames(ego_sim@termsim) <- final_clean_labels[match(colnames(ego_sim@termsim), raw_ids)]
      }
      # ========================================================================
      
      n_terms <- nrow(as.data.frame(ego_sim))
      
      if (n_terms >= 3) {
        target_groups <- if (n_terms < 5) (n_terms - 1) else 5
        
        suppressWarnings({
          p_tree <- treeplot(
            ego_sim, 
            showCategory = 40, 
            nCluster     = target_groups
          ) + 
            labs(
              title = paste0("Cluster ", clust_id, " GO Tree (Top 40 Terms)"),
              subtitle = sprintf("Terms grouped by semantic similarity (%d modules colored)", target_groups)
            ) +
            theme(
              axis.text.y = element_text(size = 9, face = "bold") 
            )
          
          if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)
          
          base_plot_name <- sprintf("cluster_%s_%s_m%.2f-resc%.1f_treeplot.png", clust_id, target_db, clust_thresh, lfc_rescue)
          ggsave(
            filename = file.path(results_dir, base_plot_name), 
            plot = p_tree, 
            width = 18,     # Extra room for descriptive words
            height = 11, 
            dpi = 300
          )
        })
        
        print(p_tree)
      } else {
        message(sprintf("Skipping treeplot for Cluster %s: Insufficient terms (%d found) for semantic clustering.", clust_id, n_terms))
      }
      
      # Append results using the stable original object data
      enrich_list[[as.character(clust_id)]] <- as.data.frame(ego) %>%
        mutate(Analysis_Group = paste0("Cluster_", clust_id))
        
    } else {
      message("No significant GO enrichment found passing thresholds for Cluster ", clust_id)
    }
  }
}
# ----- metadata, merge, and output ---
if (length(enrich_list) > 0) {
  all_go_res <- bind_rows(enrich_list)
  unique_go_ids <- unique(all_go_res$ID)

  message("Mapping extra metadata for ", length(unique_go_ids), " unique GO terms...")

  go_metadata_df <- AnnotationDbi::select(
    x       = GO.db,
    keys    = unique_go_ids,
    columns = c("GOID", "TERM", "DEFINITION"),
    keytype = "GOID"
  ) %>%
    dplyr::rename(ID = GOID, GO_Name = TERM, GO_Definition = DEFINITION)

  final_go_df <- all_go_res %>%
    dplyr::left_join(go_metadata_df, by = "ID")

  if (!"geneID" %in% colnames(final_go_df) & "gene_id" %in% colnames(final_go_df)) {
    final_go_df <- final_go_df %>% dplyr::rename(geneID = gene_id)
  }

  final_go_df <- final_go_df %>%
    dplyr::select(
      dplyr::any_of(c(
        "Analysis_Group", "GO_Name", "ID", "GO_Definition",
        "Count", "GeneRatio", "BgRatio", "pvalue",
        "p.adjust", "qvalue", "geneID"
      ))
    )

  if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)
  base_csv_name <- sprintf("%s_enrichment_m%.2f-resc%.1f_c%d_go_results.csv", target_db, clust_thresh, lfc_rescue, n_clusters)
  out_path <- file.path(results_dir, base_csv_name)
  write.csv(final_go_df, out_path, row.names = FALSE)

  message("GO Analysis complete. File and plots successfully saved with descriptive text names.")
  
  cat("\n--- FINAL RESULTS DATAFRAME PREVIEW ---\n")
  print(tibble::as_tibble(final_go_df), n = 20)
  
} else {
  message("Process finished: No significant GO enrichment across any clusters.")
}

select <- dplyr::select
filter <- dplyr::filter
```

```{r}
# ----- define variables -----
qtable <- 0.3          # q threshold for out table
clust_thresh <- 0.25   # cluster membership threshold
lfc_rescue <- 1        # ADDED: explicitly declare rescue metadata token
target_db <- "GO_bak"

# Keywords to filter out
housekeeping_terms <- c(
  "\\b5s\\b", "\\b16s\\b", "\\b23s\\b", 
  "rrna", 
  "ribosomal protein", 
  "^ribosomal large subunit$", 
  "^ribosomal small subunit$",
  "sigma factor", "rna polymerase"
)

raw_universe_genes <- rownames(dds)

# ----- mapping and universe definitions -----

# 1. Split multi-term genes using the flexible space regex
t2g_base <- anno_operon_clean %>%
  filter(Gene %in% raw_universe_genes) %>%
  dplyr::select(!!sym(target_db), Gene) %>%
  drop_na(!!sym(target_db)) %>%
  separate_rows(!!sym(target_db), sep = ",\\s*") 

# 2. Extract unique GO terms to fetch descriptions for housekeeping filters
unique_raw_gos <- unique(t2g_base[[target_db]])

library(GO.db)
clean_defs <- AnnotationDbi::select(
  x       = GO.db,
  keys    = unique_raw_gos,
  columns = c("GOID", "TERM"),
  keytype = "GOID"
) %>%
  dplyr::rename(ID = GOID, Description = TERM) %>%
  drop_na(Description)

# 3. Filter out gene-term pairs matching housekeeping terms
t2g <- t2g_base %>%
  rename(ID = !!sym(target_db)) %>%
  left_join(clean_defs, by = "ID") %>%
  filter(!grepl(paste(housekeeping_terms, collapse = "|"), Description, ignore.case = TRUE)) %>%
  select(term_id = ID, gene_id = Gene) %>%
  filter(term_id != "" & term_id != "NA") %>%
  distinct()

universe_vector <- unique(t2g$gene_id)

message("Building GO ID to Term Name mapping table...")
# CHANGE 1: Maintain structural names expected by enricher but bake in the combined layout
t2n <- clean_defs %>% 
  mutate(Combined_Label = sprintf("%s: %s", ID, Description)) %>%
  select(ID, Description = Combined_Label) %>%
  filter(ID %in% unique(t2g$term_id))

# Prepare cluster inputs matching your correct upstream data profile
cluster_input <- df_final_clusters %>%
  filter(is_driver == "High-Confidence" & Gene %in% universe_vector) %>%
  select(Gene, cluster)

all_clusters <- sort(unique(cluster_input$cluster))
n_clusters <- length(all_clusters)
enrich_list <- list()

message("Starting GO Enrichment for ", n_clusters, " clusters...")

# ----- enrichment and visualization ---
for (clust_id in all_clusters) {
  current_genes <- cluster_input %>%
    filter(cluster == clust_id) %>%
    pull(Gene)

  if (length(intersect(current_genes, universe_vector)) > 0) {
    
    # Run enricher natively with robust alphanumeric IDs
    ego <- enricher(
      gene         = current_genes,
      universe     = universe_vector, 
      TERM2GENE    = t2g,
      TERM2NAME    = t2n,     
      pvalueCutoff = 1,
      qvalueCutoff = qtable,
      minGSSize    = 2,
      maxGSSize    = 500
    )

    # Check if there are valid results before continuing
    if (!is.null(ego) && nrow(as.data.frame(ego)) > 0) {
      
      ego@ontology <- "BP" 
      
      # Step 1: Calculate semantic matrix while slots match perfectly
      ego_sim <- pairwise_termsim(ego, method = "JC", semData = NULL)
      
# ========================================================================
      # THE POST-STATISTICAL TREEPLOT SLOT INTERCEPT PATCH (STREAMLINED)
      # ========================================================================
      # Build a direct mapping dictionary from the internal alphanumeric keys
      raw_ids <- ego_sim@result$ID
      matched_words <- t2n$Description[match(raw_ids, t2n$ID)] # Pulls "GOID: Description"
      final_clean_labels <- ifelse(!is.na(matched_words) & matched_words != "", matched_words, raw_ids)
      
      # ONLY patch the Description slot. Treeplot reads this for text rendering,
      # while everything else (ID, matrix dimensions) stays safely untouched.
      ego_sim@result$Description <- final_clean_labels
      # ========================================================================
      n_terms <- nrow(as.data.frame(ego_sim))
      
      if (n_terms >= 3) {
        target_groups <- if (n_terms < 5) (n_terms - 1) else 5
        
        suppressWarnings({
          p_tree <- treeplot(
            ego_sim, 
            showCategory = 40, 
            nCluster     = target_groups
          ) + 
            labs(
              title = paste0("Cluster ", clust_id, " GO Tree (Top 40 Terms)"),
              subtitle = sprintf("Terms grouped by semantic similarity (%d modules colored)", target_groups)
            ) +
            theme(
              axis.text.y = element_text(size = 9, face = "bold") 
            )
          
          if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)
          
          base_plot_name <- sprintf("cluster_%s_%s_m%.2f-resc%.1f_treeplot.png", clust_id, target_db, clust_thresh, lfc_rescue)
          ggsave(
            filename = file.path(results_dir, base_plot_name), 
            plot = p_tree, 
            width = 18,     # Extra room for descriptive words
            height = 11, 
            dpi = 300
          )
        })
        
        print(p_tree)
      } else {
        message(sprintf("Skipping treeplot for Cluster %s: Insufficient terms (%d found) for semantic clustering.", clust_id, n_terms))
      }
      
      # Append results using the stable original object data
      enrich_list[[as.character(clust_id)]] <- as.data.frame(ego) %>%
        mutate(Analysis_Group = paste0("Cluster_", clust_id))
        
    } else {
      message("No significant GO enrichment found passing thresholds for Cluster ", clust_id)
    }
  }
}
# ----- metadata, merge, and output ---
if (length(enrich_list) > 0) {
  all_go_res <- bind_rows(enrich_list)
  unique_go_ids <- unique(all_go_res$ID)

  message("Mapping extra metadata for ", length(unique_go_ids), " unique GO terms...")

  go_metadata_df <- AnnotationDbi::select(
    x       = GO.db,
    keys    = unique_go_ids,
    columns = c("GOID", "TERM", "DEFINITION"),
    keytype = "GOID"
  ) %>%
    dplyr::rename(ID = GOID, GO_Name = TERM, GO_Definition = DEFINITION)

  final_go_df <- all_go_res %>%
    dplyr::left_join(go_metadata_df, by = "ID")

  if (!"geneID" %in% colnames(final_go_df) & "gene_id" %in% colnames(final_go_df)) {
    final_go_df <- final_go_df %>% dplyr::rename(geneID = gene_id)
  }

  final_go_df <- final_go_df %>%
    dplyr::select(
      dplyr::any_of(c(
        "Analysis_Group", "GO_Name", "ID", "GO_Definition",
        "Count", "GeneRatio", "BgRatio", "pvalue",
        "p.adjust", "qvalue", "geneID"
      ))
    )

  if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)
  base_csv_name <- sprintf("%s_enrichment_m%.2f-resc%.1f_c%d_go_results.csv", target_db, clust_thresh, lfc_rescue, n_clusters)
  out_path <- file.path(results_dir, base_csv_name)
  write.csv(final_go_df, out_path, row.names = FALSE)

  message("GO Analysis complete. File and plots successfully saved with descriptive text names.")
  
  cat("\n--- FINAL RESULTS DATAFRAME PREVIEW ---\n")
  print(tibble::as_tibble(final_go_df), n = 20)
  
} else {
  message("Process finished: No significant GO enrichment across any clusters.")
}

select <- dplyr::select
filter <- dplyr::filter
```



```{r}
# ==============================================================================
# 0. GLOBAL DEFINITIONS & INPUT DATA SETTINGS
# ==============================================================================
qtable       <- 0.3      # q threshold for output filtering

clust_thresh <- 0.25     # cluster membership threshold
lfc_rescue   <- 1        # Log2 Fold Change fallback rescue threshold
target_db    <- "GO_bak"

housekeeping_terms <- c(
  "\\b5s\\b", "\\b16s\\b", "\\b23s\\b", "rrna", "ribosomal protein", 
  "^ribosomal large subunit$", "^ribosomal small subunit$", "sigma factor", "rna polymerase"
)

raw_universe_genes <- rownames(dds)

# Map background genes to custom GO terms
t2g_base <- anno_operon_clean %>%
  filter(Gene %in% raw_universe_genes) %>%
  dplyr::select(!!sym(target_db), Gene) %>%
  drop_na(!!sym(target_db)) %>%
  separate_rows(!!sym(target_db), sep = ",\\s*")

unique_raw_gos <- unique(t2g_base[[target_db]])

library(GO.db)
clean_defs <- AnnotationDbi::select(
  x       = GO.db,
  keys    = unique_raw_gos,
  columns = c("GOID", "TERM"),
  keytype = "GOID"
) %>%
  dplyr::rename(ID = GOID, Description = TERM) %>%
  drop_na(Description)

t2g <- t2g_base %>%
  rename(ID = !!sym(target_db)) %>%
  left_join(clean_defs, by = "ID") %>%
  filter(!grepl(paste(housekeeping_terms, collapse = "|"), Description, ignore.case = TRUE)) %>%
  select(term_id = ID, gene_id = Gene) %>%
  filter(term_id != "" & term_id != "NA") %>%
  distinct()

universe_vector <- unique(t2g$gene_id)

# ==============================================================================
# CORRECTED MAPPING & ENRICHMENT FOR NON-MODEL ORGANISMS
# ==============================================================================

# ==============================================================================
# COMBINED GO ID AND DESCRIPTION MAPPING
# ==============================================================================

message("Building combined GO ID and Term Name mapping table...")
t2n <- clean_defs %>% 
  # Combine ID and Description into the expected Description slot
  mutate(Combined_Label = sprintf("%s: %s", ID, Description)) %>%
  select(ID, Description = Combined_Label) %>%
  filter(ID %in% unique(t2g$term_id))

cluster_input <- df_final_clusters %>%
  filter(is_driver == "High-Confidence" & Gene %in% universe_vector) %>%
  select(Gene, cluster)

gene_list_by_cluster <- split(cluster_input$Gene, cluster_input$cluster)

# Run standard comparative enrichment base tool
comp_ego <- compareCluster(
  geneClusters = gene_list_by_cluster, 
  fun          = "enricher", 
  universe     = universe_vector,
  TERM2GENE    = t2g,
  TERM2NAME    = t2n,  # Now contains "GOXXXXX: Functional Description"
  pvalueCutoff = 1,
  qvalueCutoff = qtable
)
attr(comp_ego, "ontology") <- "BP"

# Establish target base directories
if (!dir.exists(results_dir)) dir.create(results_dir, recursive = TRUE)
subfolder_dir <- file.path(results_dir, "reshaped_outputs")
if (!dir.exists(subfolder_dir)) dir.create(subfolder_dir, recursive = TRUE)

n_clusters <- length(unique(cluster_input$cluster))
base_name  <- sprintf("%s_enrichment_m%.2f-resc%.1f_c%d", target_db, clust_thresh, lfc_rescue, n_clusters)

# ==============================================================================
# PIPELINE STREAM 1: COLLAPSED / SEMANTIC FUNCTIONAL MODULE OUTPUTS
# ==============================================================================
message("\n--- Executing Stream 1: Generating Semantically Collapsed Visualizations ---")

comp_ego_sim <- pairwise_termsim(comp_ego, method = "JC", semData = NULL)

suppressWarnings({
  p_compare_collapsed <- dotplot(comp_ego_sim, x = "Cluster", showCategory = 5) +
    labs(
      title    = "Functional Mapping Across Analysis Groups",
      subtitle = "Redundant terms collapsed dynamically into structural summary modules",
      x        = "Analysis Group (Cluster)", y = "Enriched Functional Module"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
      axis.text.y = element_text(fontsize = 9),
      plot.title  = element_text(face = "bold", size = 14)
    )

  collapsed_plot_path <- file.path(results_dir, paste0(base_name, "_collapsed_modules_dotplot.png"))
  ggsave(filename = collapsed_plot_path, plot = p_compare_collapsed, width = 15, height = 12, dpi = 300)
})
print(p_compare_collapsed)


# ==============================================================================
# PIPELINE STREAM 2: UNCOLLAPSED GRANULAR SPECIFIC GO TERM OUTPUTS
# ==============================================================================
message("\n--- Executing Stream 2: Generating Granular Uncollapsed Term Outputs ---")

# Step A: Generate Granular Dotplot using the base uncollapsed object
suppressWarnings({
  p_compare_granular <- dotplot(comp_ego, x = "Cluster", showCategory = 5) +
    labs(
      title    = "Granular Functional Mapping Across Analysis Groups",
      subtitle = "Redundant terms remain split; rendering explicit top terms per cluster",
      x        = "Analysis Group (Cluster)", y = "Enriched Specific Functional Term"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
      axis.text.y = element_text(fontsize = 9),
      plot.title  = element_text(face = "bold", size = 14)
    )

  granular_plot_path <- file.path(results_dir, paste0(base_name, "_granular_specific_terms_dotplot.png"))
  ggsave(filename = granular_plot_path, plot = p_compare_granular, width = 15, height = 12, dpi = 300)
})
print(p_compare_granular)

# Step B: Export full uncollapsed text variations to matching data frames
if (!is.null(comp_ego) && nrow(as.data.frame(comp_ego)) > 0) {
  
  final_enrichment_df <- as.data.frame(comp_ego) %>%
    select(-any_of("Description")) %>%
    left_join(clean_defs, by = "ID") %>%
    select(Cluster, ID, Description, Count, everything())

  # 1. Master long table
  original_out_path <- file.path(results_dir, paste0(base_name, "_granular_results.tsv"))
  write.table(final_enrichment_df, original_out_path, sep = "\t", row.names = FALSE, quote = FALSE)

  # 2. Organized by cluster, then gene long format
  tidy_long_df <- final_enrichment_df %>%
    separate_rows(geneID, sep = "/") %>%
    rename(Gene_ID = geneID) %>%
    select(Cluster, Gene_ID, ID, Description, everything()) %>%
    arrange(Cluster, Gene_ID, qvalue)

  long_out_path <- file.path(subfolder_dir, paste0(base_name, "_granular_long_by_gene.tsv"))
  write.table(tidy_long_df, long_out_path, sep = "\t", row.names = FALSE, quote = FALSE)

  # 3. Binary presence/absence matrix tracking individual GO terms
  matrix_wide_df <- final_enrichment_df %>%
    mutate(Enriched = 1) %>%
    select(ID, Description, Cluster, Enriched) %>%
    distinct() %>%
    pivot_wider(
      names_from  = Cluster,
      values_from = Enriched,
      values_fill = 0
    ) %>%
    arrange(ID)

  wide_out_path <- file.path(subfolder_dir, paste0(base_name, "_granular_cluster_profile_matrix.tsv"))
  write.table(matrix_wide_df, wide_out_path, sep = "\t", row.names = FALSE, quote = FALSE)

  message("Success! Exported uncollapsed granular term datasets:")
  message(sprintf("  -> Original Table: %s", basename(original_out_path)))
  message(sprintf("  -> Long Tidy: %s", file.path("reshaped_outputs", basename(long_out_path))))
  message(sprintf("  -> Wide Profile Matrix: %s", file.path("reshaped_outputs", basename(wide_out_path))))
}

# ------ Reset Namespace Pointers -----
select <- dplyr::select
filter <- dplyr::filter
message("Namespace reset completed successfully.")
```


# !!! ELIM

```{r}

# Prepare a clean, named list mapping Genes to GO terms for topGO
gene_2_go_list <- split(t2g$term_id, t2g$gene_id)

# Identify all available clusters from your fuzzy clustering input
all_cluster_ids <- sort(unique(df_final_clusters$cluster))

# Custom wrapper to run Elim for a single ontology branch
run_tiered_ontology <- function(current_cluster, ontology_branch) {
  # Get genes belonging to the current cluster
  cluster_genes <- df_final_clusters %>%
    filter(cluster == current_cluster & is_driver == "High-Confidence") %>%
#        filter(cluster == current_cluster & membership_score >= clust_thresh) %>%
    pull(Gene)

  # Create the binary factor vector topGO expects (1 = in cluster, 0 = background)
  selector_vector <- as.integer(universe_vector %in% cluster_genes)
  names(selector_vector) <- universe_vector
  selector_vector <- factor(selector_vector)

  # Initialize the structural topGO data object
  go_data_obj <- new("topGOdata",
    ontology = ontology_branch,
    allGenes = selector_vector,
    annot = annFUN.gene2GO,
    gene2GO = gene_2_go_list
  )

  # Execute the TRUE tiered Elim algorithm
  elim_result <- runTest(go_data_obj, algorithm = "elim", statistic = "fisher")

  # Extract the top non-redundant terms
  raw_table <- GenTable(go_data_obj,
    Elim_p = elim_result,
    orderBy = "Elim_p",
    topNodes = 40,
    numChar = 1000
  )

  # Format and clean the output metadata
  raw_table %>%
    mutate(
      Analysis_Group = paste0("Cluster_", current_cluster),
      Ontology = ifelse(ontology_branch == "BP", "Biological Process", "Molecular Function"),
      # Clean character p-values (e.g., "< 1e-30") to numeric values
      qvalue = as.numeric(gsub("< ", "", Elim_p))
    ) %>%
    filter(qvalue <= 0.05) %>% # Keep terms that survived decorrelation
    dplyr::rename(ID = GO.ID, GO_Name = Term) %>%
    dplyr::select(Analysis_Group, ID, GO_Name, Ontology, qvalue, Significant)
}
```


```{r clustering-of-elim}
message("Running global structural Elim decorrelation...")

compiled_elim_results <- map_dfr(all_cluster_ids, function(clust) {
  message("  Processing Cluster ", clust, "...")

  # Safe execution block to gracefully handle empty/flat branches
  bp_res <- tryCatch(run_tiered_ontology(clust, "BP"), error = function(e) NULL)
  mf_res <- tryCatch(run_tiered_ontology(clust, "MF"), error = function(e) NULL)

  bind_rows(bp_res, mf_res)
})

# Quick validation check to confirm cluster presence
print(table(compiled_elim_results$Analysis_Group))
```


This was originally a few different scripts - combining them broke my variable naming, this fixes it. 

```{r fix-variable-structure}

if ("Ontology" %in% colnames(compiled_elim_results) && !"Ontology_Branch" %in% colnames(compiled_elim_results)) {
  compiled_elim_results <- compiled_elim_results %>%
    mutate(Ontology_Branch = case_when(
      Ontology == "Biological Process" ~ "BP",
      Ontology == "Molecular Function"  ~ "MF",
      TRUE ~ Ontology
    ))
} else if ("Ontology_Branch" %in% colnames(compiled_elim_results) && !"Ontology" %in% colnames(compiled_elim_results)) {
  compiled_elim_results <- compiled_elim_results %>%
    mutate(Ontology = case_when(
      Ontology_Branch == "BP" ~ "Biological Process",
      Ontology_Branch == "MF" ~ "Molecular Function",
      TRUE ~ Ontology_Branch
    ))
}
```

Split the biological processes terms and molecular function terms for plotting

```{r biological-processes-bubbleplots}

# ----- filter and mutate data -----
bp_base_data <- compiled_elim_results %>%
  filter(Ontology == "Biological Process") %>%
  mutate(neg_log_p = -log10(qvalue))

# theme to share across plots
shared_theme <- theme_minimal(base_size = 10) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, face = "bold"),
    axis.text.y = element_text(face = "bold", size = 8),
    panel.grid.major = element_line(color = "gray95"),
    plot.title = element_text(face = "bold", size = 13)
  )

# ----- plot 1: top 10 terms, ordered by significance ----- 
bp_standard_data <- bp_base_data %>%
  group_by(Analysis_Group) %>%
  slice_max(order_by = neg_log_p, n = 10, with_ties = FALSE) %>%
  ungroup()

plot_standard <- ggplot(bp_standard_data, aes(x = Analysis_Group, y = reorder(GO_Name, neg_log_p))) +
  geom_point(aes(size = Significant, color = neg_log_p)) +
  scale_color_gradient(low = "deepskyblue", high = "magenta4", name = "-log10(p)") +
  shared_theme +
  labs(
    title = "GO Biological Process Architecture",
    subtitle = "Decorrelated Macro-Phenotypes & Temporal Pathways via Elim",
    x = "Fuzzy Cluster Wave",
    y = "Decorrelated Biological Process",
    size = "Gene Count"
  )

# print and save
print(plot_standard)
ggsave(
  filename = file.path(results_dir, "elim_landscape_biological_process_rescue.png"),
  plot = plot_standard,
  width = 9, height = 12, units = "in", dpi = 300
)

# ----- plot 2: top 8 terms, hierarchically clustered by hclust ----- 
bp_clust_data <- bp_base_data %>%
  group_by(Analysis_Group) %>%
  slice_max(order_by = neg_log_p, n = 8, with_ties = FALSE) %>%
  ungroup()

# Calculate distance matrices and orders
wide_matrix_df <- bp_clust_data %>%
  dplyr::select(GO_Name, Analysis_Group, neg_log_p) %>%
  distinct() %>%
  pivot_wider(names_from = Analysis_Group, values_from = neg_log_p, values_fill = 0)

clust_matrix <- as.matrix(wide_matrix_df[, -1])
rownames(clust_matrix) <- wide_matrix_df$GO_Name

# Cluster rows (GO terms)
row_dist <- dist(clust_matrix, method = "euclidean")
row_hc   <- hclust(row_dist, method = "complete")
go_term_order <- rownames(clust_matrix)[row_hc$order]

# Cluster columns (analysis groups/fuzzy clusters)
if (ncol(clust_matrix) >= 3) {
  col_dist <- dist(t(clust_matrix), method = "euclidean")
  col_hc   <- hclust(col_dist, method = "complete")
  analysis_group_order <- colnames(clust_matrix)[col_hc$order]
} else {
  analysis_group_order <- sort(unique(bp_clust_data$Analysis_Group))
}

# Apply clustered factors back to data
bp_plot_clustered_data <- bp_clust_data %>%
  mutate(
    GO_Name        = factor(GO_Name, levels = go_term_order),
    Analysis_Group = factor(Analysis_Group, levels = analysis_group_order)
  )

plot_clustered <- ggplot(bp_plot_clustered_data, aes(x = Analysis_Group, y = GO_Name)) +
  geom_point(aes(size = Significant, color = neg_log_p)) +
  scale_color_gradient(low = "deepskyblue1", high = "magenta4", name = "-log10(p)") +
  shared_theme +
  labs(
    title = "Hierarchically Clustered BP Functional Landscape",
    subtitle = "Decorrelated Macro-Phenotypes & Temporal Pathways via Elim",
    x = "Hierarchically Ordered Cluster Waves",
    y = "Clustered Biological Process Terms",
    size = "Gene Count"
  )

# Print and save
print(plot_clustered)
ggsave(
  filename = file.path(results_dir, "elim_landscape_biological_process_rescue_hier_n8.png"),
  plot = plot_clustered,
  width = 9, height = 12, units = "in", dpi = 300
)

message("Both BP plots rendered to console and saved as PNG format.")
```
potential colour combinations: steelblue1-purple, deepskyblue-magenta4, gold2-cyan3


```{r molecular-function-bubbleplots}
# ----- filter and mutate -----
mf_base_data <- compiled_elim_results %>%
  filter(Ontology == "Molecular Function") %>%
  mutate(neg_log_p = -log10(qvalue))

# Common theme to share across plots
shared_theme <- theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, face = "bold"),
    axis.text.y = element_text(face = "bold", size = 9),
    panel.grid.major = element_line(color = "gray95"),
    plot.title = element_text(face = "bold", size = 13)
  )

# this set of labels gets too long for display
# trims long labels to 50 characters and append "..."
truncate_y_labels <- function(x) stringr::str_trunc(x, width = 50, side = "right")

# ----- plot 1: top 10 terms, ordered by significance ----- 
mf_standard_data <- mf_base_data %>%
  group_by(Analysis_Group) %>%
  slice_max(order_by = neg_log_p, n = 10, with_ties = FALSE) %>%
  ungroup()

plot_standard <- ggplot(mf_standard_data, aes(x = Analysis_Group, y = reorder(GO_Name, neg_log_p))) +
  geom_point(aes(size = Significant, color = neg_log_p)) +
  scale_color_gradient(low = "deepskyblue", high = "magenta4", name = "-log10(p)") +
  scale_y_discrete(labels = truncate_y_labels) + # Truncates labels visually
  shared_theme +
  labs(
    title = "GO Molecular Function Architecture",
    subtitle = "Decorrelated Enzymatic, Binding, & Regulatory Mechanisms via Elim",
    x = "Fuzzy Cluster Wave",
    y = "Decorrelated Molecular Function",
    size = "Gene Count"
  )

# Print and save
print(plot_standard)
ggsave(
  filename = file.path(results_dir, "elim_landscape_molecular_function_rescue_n10.png"),
  plot = plot_standard,
  width = 9, height = 12, units = "in", dpi = 300
)

# ----- plot 2: top 8 terms, hierarchically clustered by hclust ----- 
mf_clust_data <- mf_base_data %>%
  group_by(Analysis_Group) %>%
  slice_max(order_by = neg_log_p, n = 10, with_ties = FALSE) %>%
  ungroup()

# calculate distance matrices and orders using un-truncated text strings
wide_matrix_df <- mf_clust_data %>%
  dplyr::select(GO_Name, Analysis_Group, neg_log_p) %>%
  distinct() %>%
  pivot_wider(names_from = Analysis_Group, values_from = neg_log_p, values_fill = 0)

clust_matrix <- as.matrix(wide_matrix_df[, -1])
rownames(clust_matrix) <- wide_matrix_df$GO_Name

# Cluster rows (GO terms)
row_dist <- dist(clust_matrix, method = "euclidean")
row_hc   <- hclust(row_dist, method = "complete")
go_term_order <- rownames(clust_matrix)[row_hc$order]

# Cluster columns (analysis groups)
if (ncol(clust_matrix) >= 3) {
  col_dist <- dist(t(clust_matrix), method = "euclidean")
  col_hc   <- hclust(col_dist, method = "complete")
  analysis_group_order <- colnames(clust_matrix)[col_hc$order]
} else {
  analysis_group_order <- sort(unique(mf_clust_data$Analysis_Group))
}

# Apply clustered factors back to data
mf_plot_clustered_data <- mf_clust_data %>%
  mutate(
    GO_Name        = factor(GO_Name, levels = go_term_order),
    Analysis_Group = factor(Analysis_Group, levels = analysis_group_order)
  )

plot_clustered <- ggplot(mf_plot_clustered_data, aes(x = Analysis_Group, y = GO_Name)) +
  geom_point(aes(size = Significant, color = neg_log_p)) +
  scale_color_gradient(low = "deepskyblue", high = "magenta4", name = "-log10(p)") +
  scale_y_discrete(labels = truncate_y_labels) + # Truncates labels visually
  shared_theme +
  labs(
    title = "Hierarchically Clustered MF Functional Landscape",
    subtitle = "Decorrelated Enzymatic, Binding, & Regulatory Mechanisms via Elim",
    x = "Hierarchically Ordered Cluster Waves",
    y = "Clustered Molecular Function Terms",
    size = "Gene Count"
  )

# Print and save
print(plot_clustered)
ggsave(
  filename = file.path(results_dir, "elim_landscape_molecular_function_rescue_hier_n10.png"),
  plot = plot_clustered,
  width = 9, height = 12, units = "in", dpi = 300
)

message("Both MF plots rendered to console, saved as high-res PNG, and y-axis character limits applied successfully.")
```



```{r cluster-elim-visualization}
# ----- get target clusters ---
all_analysis_groups <- sort(unique(compiled_elim_results$Analysis_Group))

message("Initiating master visualization loop for ", length(all_analysis_groups), " groups...")

# ----- rollercoaster ---
for (group_id in all_analysis_groups) {
  
  message("------------------------------------------------------------")
  message("Processing Plots for Group: ", group_id)
  
  raw_clust_num <- stringr::str_remove(group_id, "Cluster_")
  
  cluster_dir <- file.path(results_dir, group_id)
  if (!dir.exists(cluster_dir)) dir.create(cluster_dir, recursive = TRUE)
  
  cluster_base_data <- compiled_elim_results %>%
    filter(Analysis_Group == group_id) %>%
    mutate(neg_log_p = -log10(qvalue))
  
  if (nrow(cluster_base_data) == 0) {
    message("  Skipping: No enrichment records found for ", group_id)
    next
  }

  # ----- significance-weighted word cloud -----
  tryCatch({
    word_counts <- cluster_base_data %>%
      filter(qvalue <= 0.05) %>%
      dplyr::select(GO_Name, neg_log_p) %>%
      unnest_tokens(word, GO_Name) %>%
      anti_join(stop_words, by = "word") %>%
      # FIXED SYNTAX: Changed %in= to %in%
      filter(!word %in% c("activity", "process", "regulation", "positive", "negative", "cellular", "biological") & 
             !stringr::str_detect(word, "^\\d+$"))
    
    if (nrow(word_counts) > 0) {
      word_cloud_data <- word_counts %>%
        group_by(word) %>%
        summarize(weight = max(neg_log_p), count = n()) %>%
        ungroup() %>%
        slice_max(order_by = weight, n = 50, with_ties = FALSE)
      
      p_wordcloud <- ggplot(word_cloud_data, aes(label = word, size = weight, color = weight)) +
        geom_text_wordcloud_area(
          shape = "circle", 
          eccentricity = 0.7,
          grid_size = 4,           
          grid_margin = 1.5,       
          fontface = "bold",       
          family = "sans"
        ) +
        scale_size_area(max_size = 7) + 
        scale_color_gradient(low = "deepskyblue", high = "magenta4", name = "-log10(p)") +
        theme_minimal() +
        labs(
          title = paste0(group_id, " Functional Word Cloud"),
          subtitle = "Visual emphasis scaled by maximum -log10(p-value) [Numbers Removed]"
        ) +
        theme(
          plot.title = element_text(face = "bold", size = 13, hjust = 0.5),
          plot.subtitle = element_text(size = 10, hjust = 0.5)
        )
      
      print(p_wordcloud)
      ggsave(filename = file.path(cluster_dir, paste0(group_id, "_wordcloud.png")),
             plot = p_wordcloud, width = 8, height = 6, units = "in", dpi = 300)
      message("  [✓] Non-overlapping Word Cloud saved.")
    } else {
      message("  [-] Word Cloud skipped: No qualifying terms.")
    }
  }, error = function(e) { message("  [!] Error building Word Cloud: ", e$message) })

  # ----- structural tree heatmap dendogram -----
  tryCatch({
    dendro_subset <- cluster_base_data %>%
      mutate(Short_Name = stringr::str_trunc(GO_Name, width = 50, side = "right")) %>%
      slice_max(order_by = neg_log_p, n = 15, with_ties = FALSE)
    
    if (nrow(dendro_subset) >= 2) {
      hc_matrix <- as.matrix(dendro_subset$neg_log_p)
      rownames(hc_matrix) <- dendro_subset$Short_Name
      
      hc_dist <- dist(hc_matrix, method = "euclidean")
      hc_obj  <- hclust(hc_dist, method = "complete")
      dendro_data_obj <- dendro_data(hc_obj)
      
      dendro_subset$Short_Name <- factor(dendro_subset$Short_Name, levels = hc_obj$labels[hc_obj$order])
      
      p_tree <- ggplot(segment(dendro_data_obj)) +
        geom_segment(aes(x = x, y = y, xend = xend, yend = yend), color = "gray40", width = 0.7) +
        coord_flip() + scale_y_reverse(expand = c(0.1, 0)) + theme_dendro()
      
      p_heatmap <- ggplot(dendro_subset, aes(x = "Significance Profile", y = Short_Name)) +
        geom_tile(aes(fill = neg_log_p), color = "white", width = 0.9) +
        geom_text(aes(label = round(neg_log_p, 2)), fontface = "bold", size = 3, color = "black") +
        scale_fill_gradient(low = "deepskyblue", high = "magenta4", name = "-log10(p)") +
        scale_y_discrete(position = "right") +
        theme_minimal() +
        theme(axis.text.x = element_text(face = "bold", size = 10),
              axis.text.y = element_text(face = "bold", size = 9),
              axis.title = element_blank(), panel.grid = element_blank())
      
      combined_tree_plot <- p_tree + p_heatmap + 
        plot_layout(widths = c(1, 2.5)) +
        plot_annotation(
          title = paste0(group_id, ": Structural Tree-Heatmap Dendrogram"),
          subtitle = "Functional modules grouped hierarchically by numerical variance profiles",
          theme = theme(plot.title = element_text(face = "bold", size = 13))
        )
      
      print(combined_tree_plot)
      ggsave(filename = file.path(cluster_dir, paste0(group_id, "_hierarchical_tree_heatmap.png")),
             plot = combined_tree_plot, width = 9, height = 7, units = "in", dpi = 300)
      message("  [✓] Hierarchical Tree-Heatmap saved.")
    } else {
      message("  [-] Tree-Heatmap skipped: Requires at least 2 distinct pathways.")
    }
  }, error = function(e) { message("  [!] Error building Tree-Heatmap: ", e$message) })

  # ----- pathway concept network (cnetplot) -----

  tryCatch({
    net_terms <- cluster_base_data %>%
      slice_max(order_by = neg_log_p, n = 10, with_ties = FALSE)
    
    current_drivers <- cluster_input %>% 
      filter(cluster == raw_clust_num | cluster == group_id) %>% 
      pull(Gene)
    
    network_edges_raw <- t2g %>%
      filter(gene_id %in% current_drivers & term_id %in% net_terms$ID) %>%
      left_join(net_terms, by = c("term_id" = "ID")) %>%
      mutate(Short_Name = stringr::str_trunc(GO_Name, width = 40, side = "right")) %>%
      dplyr::select(Short_Name, gene_id, neg_log_p, Significant)
    
    if (nrow(network_edges_raw) > 0) {
      edges <- network_edges_raw %>%
        dplyr::rename(from = Short_Name, to = gene_id) %>%
        dplyr::select(from, to) %>% distinct()
      
      nodes_pathway <- net_terms %>%
        mutate(Short_Name = stringr::str_trunc(GO_Name, width = 40, side = "right")) %>%
        filter(Short_Name %in% edges$from) %>%
        dplyr::select(Short_Name, neg_log_p, Significant) %>%
        dplyr::rename(name = Short_Name, size_val = Significant, color_val = neg_log_p) %>%
        mutate(node_type = "Pathway")
      
      nodes_gene <- data.frame(name = unique(edges$to)) %>%
        mutate(size_val = 1, color_val = NA, node_type = "Gene")
      
      nodes <- bind_rows(nodes_pathway, nodes_gene)
      graph_obj <- tbl_graph(nodes = nodes, edges = edges, directed = FALSE)
      
      network_plot <- ggraph(graph_obj, layout = "stress") +
        geom_edge_link(alpha = 0.15, color = "gray50", width = 0.5) +
        geom_node_point(aes(size = size_val, color = color_val, shape = node_type)) +
        scale_shape_manual(values = c("Pathway" = 16, "Gene" = 17)) +
        scale_color_gradient(low = "deepskyblue", high = "magenta4", na.value = "gold", name = "-log10(p)") +
        scale_size_continuous(name = "Node Weight", range = c(2.5, 8)) +
        geom_node_text(aes(label = ifelse(node_type == "Pathway", name, "")), 
                       repel = TRUE, fontface = "bold", size = 3, bg.color = "white", bg.r = 0.1) +
        theme_graph(base_family = "sans") +
        labs(
          title = paste0(group_id, ": Pathway Concept Network (cnetplot)"),
          subtitle = "Triangles represent component genes mapped from t2g universe definitions"
        )
      
      print(network_plot)
      ggsave(filename = file.path(cluster_dir, paste0(group_id, "_pathway_concept_network.png")),
             plot = network_plot, width = 11, height = 9, units = "in", dpi = 300)
      message("  [✓] Concept Network Map saved.")
    } else {
      message("  [-] Concept Network skipped: No gene-to-term linkages verified.")
    }
  }, error = function(e) { message("  [!] Error building Concept Network Map: ", e$message) })

}


message("Success! All pictures are pretty and present")
```

# extended cnetplots
```{r loop-all-three-cnetplot-variants}
# ----- define which GO branches to use -----
all_analysis_groups <- sort(unique(compiled_elim_results$Analysis_Group))
cnet_variants <- c("BP", "MF", "Combined")

message("Initiating master concept network engine for ", length(all_analysis_groups), " groups...")
message("Generating 3 distinct structural variants per cluster wave...")

# ----- rollercoaster ----- 
for (group_id in all_analysis_groups) {
  
  raw_clust_num <- stringr::str_remove(group_id, "Cluster_")
  cluster_dir <- file.path(results_dir, group_id)
  if (!dir.exists(cluster_dir)) dir.create(cluster_dir, recursive = TRUE)
  
  # Pull driving gene inputs for the active wave
  current_drivers <- cluster_input %>% 
    filter(cluster == raw_clust_num | cluster == group_id) %>% 
    pull(Gene)
  
  # Inner Loop: BP, MF, and Combined 
  for (variant in cnet_variants) {
    
    tryCatch({
      # Dynamic subset based on loop target
      if (variant == "Combined") {
        subtitle_text <- "Combined Ontology Profile (Biological Process & Molecular Function)"
        
        net_terms <- compiled_elim_results %>%
          filter(Analysis_Group == group_id) %>%
          mutate(neg_log_p = -log10(qvalue)) %>%
          slice_max(order_by = neg_log_p, n = 10, with_ties = FALSE)
      } else {
        full_name <- ifelse(variant == "BP", "Biological Process", "Molecular Function")
        subtitle_text <- paste0("Ontology Profile Focus: ", full_name)
        
        net_terms <- compiled_elim_results %>%
          filter(Analysis_Group == group_id & Ontology_Branch == variant) %>%
          mutate(neg_log_p = -log10(qvalue)) %>%
          slice_max(order_by = neg_log_p, n = 10, with_ties = FALSE)
      }
      
      # Safety: Skip if data slice contains no records
      if (nrow(net_terms) == 0) {
        message("  [-] Group ", group_id, " (", variant, ") skipped: No records met thresholds.")
        next
      }
      
      # Extract linkages from global annotation universe (t2g)
      network_edges_raw <- t2g %>%
        filter(gene_id %in% current_drivers & term_id %in% net_terms$ID) %>%
        left_join(net_terms, by = c("term_id" = "ID")) %>%
        mutate(Short_Name = stringr::str_trunc(GO_Name, width = 40, side = "right")) %>%
        dplyr::select(Short_Name, gene_id, neg_log_p, Significant)
      
      if (nrow(network_edges_raw) > 0) {
        
        # Isolate clean edge coordinate tables
        edges <- network_edges_raw %>%
          dplyr::rename(from = Short_Name, to = gene_id) %>%
          dplyr::select(from, to) %>% 
          distinct()
        
        # Build node metadata sets
        nodes_pathway <- net_terms %>%
          mutate(Short_Name = stringr::str_trunc(GO_Name, width = 40, side = "right")) %>%
          filter(Short_Name %in% edges$from) %>%
          dplyr::select(Short_Name, neg_log_p, Significant) %>%
          dplyr::rename(name = Short_Name, size_val = Significant, color_val = neg_log_p) %>%
          mutate(node_type = "Pathway")
        
        nodes_gene <- data.frame(name = unique(edges$to)) %>%
          mutate(size_val = 1, color_val = NA, node_type = "Gene")
        
        # Construct structural tidy graph network
        nodes <- bind_rows(nodes_pathway, nodes_gene)
        graph_obj <- tbl_graph(nodes = nodes, edges = edges, directed = FALSE)
        
        # figure
        network_plot <- ggraph(graph_obj, layout = "stress") +
          geom_edge_link(alpha = 0.15, color = "gray50", width = 0.5) +
          geom_node_point(aes(size = size_val, color = color_val, shape = node_type)) +
          scale_shape_manual(values = c("Pathway" = 16, "Gene" = 17)) +
          scale_color_gradient(low = "deepskyblue", high = "magenta4", na.value = "gold", name = "-log10(p)") +
          scale_size_continuous(name = "Node Weight", range = c(2.5, 8)) +
          geom_node_text(aes(label = ifelse(node_type == "Pathway", name, "")), 
                         repel = TRUE, fontface = "bold", size = 3, bg.color = "white", bg.r = 0.1) +
          theme_graph(base_family = "sans") +
          labs(
            title = paste0(group_id, ": Pathway Concept Network (cnetplot)"),
            subtitle = paste0(subtitle_text, " | Triangles map component gene positions")
          )
        print(network_plot)
        # dynamic save
        file_export_name <- file.path(cluster_dir, paste0(group_id, "_", variant, "_cnetplot.png"))
        
        ggsave(filename = file_export_name, plot = network_plot, width = 11, height = 9, units = "in", dpi = 300)
        message("  [✓] Saved: ", group_id, " [", variant, " Version]")
        
      } else {
        message("  [-] Linkage skip: ", group_id, " (", variant, ") has no verified gene-to-term edges.")
      }
      
    }, error = function(e) { 
      message("  [!] Error processing variant ", variant, " for ", group_id, ": ", e$message) 
    })
  }
}


message("Success! All pictures are pretty and present.")
```



# sub category elim plots

```{r}
# Ensure the subcategory directory is ready
sub_plots_dir <- file.path(results_dir, "targeted_subcategories")
if (!dir.exists(sub_plots_dir)) dir.create(sub_plots_dir, recursive = TRUE)

# Prepare universal high-resolution metric baseline
raw_high_res_df <- compiled_elim_results %>%
  mutate(
    neg_log_p = -log10(qvalue),
    Wrapped_GO_Name = str_wrap(as.character(GO_Name), width = 40)
  )

# ----- define functions -----
plot_targeted_subcategory <- function(subset_df, plot_title, filename, color_palette = "mako") {
  if (nrow(subset_df) == 0) {
    message(paste("Skipping", plot_title, "- No enriched terms matched this keyword filter."))
    return(NULL)
  }

  # Guarantee structural grid integrity across all unique clusters in your data
  complete_grid <- expand_grid(
    Analysis_Group = unique(raw_high_res_df$Analysis_Group),
    Wrapped_GO_Name = unique(subset_df$Wrapped_GO_Name)
  ) %>%
    left_join(subset_df, by = c("Analysis_Group", "Wrapped_GO_Name")) %>%
    mutate(neg_log_p = replace_na(neg_log_p, 0))

  sub_heatmap <- ggplot(complete_grid, aes(x = Analysis_Group, y = reorder(Wrapped_GO_Name, neg_log_p))) +
    geom_tile(aes(fill = neg_log_p), color = "white", lwd = 0.4) +
    scale_fill_viridis_c(option = color_palette, direction = -1, name = "-log10(p)") +
    theme_minimal(base_size = 11) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, face = "bold"),
      axis.text.y = element_text(size = 9, lineheight = 0.85, face = "bold"),
      panel.grid = element_blank(),
      plot.title = element_text(face = "bold", size = 12)
    ) +
    labs(
      title = plot_title,
      subtitle = "High-Resolution Elim Term Sub-Slices Across Waves",
      x = "Cluster Wave", y = NULL
    )

  # Save clean layout vectors
  ggsave(file.path(sub_plots_dir, paste0(filename, ".pdf")),
    plot = sub_heatmap, width = 7.5, height = max(3, nrow(subset_df) * 0.35 + 2), units = "in"
  )

  return(sub_heatmap)
}

# aa subcategory extractions 
message("Extracting targeted Amino Acid subcategories...")

# ----- A. Branched-Chain Amino Acids (BCAA) ---
bcaa_terms <- raw_high_res_df %>%
  filter(str_detect(GO_Name, "(?i)isoleucine|leucine|valine|branched-chain"))

plot_targeted_subcategory(bcaa_terms, "Branched-Chain Amino Acid (BCAA) Tracks", "sub_bp_bcaa", "mako")


# ----- B. Aromatic Amino Acids ---
aromatic_terms <- raw_high_res_df %>%
  filter(str_detect(GO_Name, "(?i)tryptophan|tyrosine|phenylalanine|aromatic amino acid"))

plot_targeted_subcategory(aromatic_terms, "Aromatic Amino Acid Modules", "sub_bp_aromatic_aa", "mako")


# ----- C. Sulfur-Containing & Nitrogen-Core Amino Acids ---
# Catches primary regulatory vectors like Glutamine/Glutamate and Methionine/Cysteine
nitrogen_sulfur_aa_terms <- raw_high_res_df %>%
  filter(str_detect(GO_Name, "(?i)glutamine|glutamate|arginine|methionine|cysteine|sulfur amino acid"))

plot_targeted_subcategory(nitrogen_sulfur_aa_terms, "Core Nitrogen & Sulfur Amino Acids", "sub_bp_nitrogen_sulfur_aa", "mako")

# ----- cyclic and aromatic amino acids 
message("Extracting targeted Cyclic & Aromatic Amino Acid subcategories...")

# captures specific cyclic branches and their core upstream precursor architecture (shikimate)
cyclic_aromatic_terms <- raw_high_res_df %>%
  filter(str_detect(GO_Name, "(?i)phenylalanine|tyrosine|tryptophan|histidine|aromatic amino acid|shikimate|chorismate"))

# make plot
plot_targeted_subcategory(
  subset_df = cyclic_aromatic_terms,
  plot_title = "Cyclic & Aromatic Amino Acid Architecture",
  filename = "sub_bp_cyclic_aromatic_aa",
  color_palette = "mako"
)


# ----- lipids and fatty acids ---
message("Extracting targeted Lipid & Fatty Acid subcategories...")

# ----- A. Fatty Acid Metabolism & Beta-Oxidation ---
# Separates raw structural breakdown chains and oxidation processing pathways
fatty_acid_oxidation_terms <- raw_high_res_df %>%
  filter(str_detect(GO_Name, "(?i)fatty acid|beta-oxidation|acyl-coa|ketone body"))

plot_targeted_subcategory(fatty_acid_oxidation_terms, "Fatty Acid Metabolism & Beta-Oxidation", "sub_lipid_fatty_acid_oxidation", "rocket")


# ----- B. Membrane Lipids & Phospholipids ---
# Isolates structural building modifications (e.g., cell wall envelope integrity)
membrane_lipid_terms <- raw_high_res_df %>%
  filter(str_detect(GO_Name, "(?i)phospholipid|glycerolipid|sphingolipid|lipid bilayer|cardiolipin"))

plot_targeted_subcategory(membrane_lipid_terms, "Membrane & Phospholipid Remodeling", "sub_lipid_membrane_structures", "rocket")

message("Targeted subcategory pipeline executed successfully. Check the 'targeted_subcategories' directory.")
```



```{r aa-branch-cyclic}
# Ensure output directory paths are ready
sub_plots_dir <- file.path(results_dir, "targeted_subcategories")
if (!dir.exists(sub_plots_dir)) dir.create(sub_plots_dir, recursive = TRUE)


message("Generating Figure 1 (BCAA + Cyclic/Aromatic)...")

fig1_data <- raw_high_res_df %>%
  mutate(
    Facet_Group = case_when(
      str_detect(GO_Name, "(?i)isoleucine|leucine|valine|branched-chain") ~ "Branched-Chain (BCAA)",
      str_detect(GO_Name, "(?i)phenylalanine|tyrosine|tryptophan|histidine|aromatic|shikimate|chorismate") ~ "Cyclic & Aromatic",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(Facet_Group))

# Matrix tracking cross-grid integrity
grid_fig1 <- expand_grid(
  Analysis_Group = unique(raw_high_res_df$Analysis_Group),
  Wrapped_GO_Name = unique(fig1_data$Wrapped_GO_Name)
) %>%
  left_join(distinct(fig1_data, Wrapped_GO_Name, Facet_Group), by = "Wrapped_GO_Name") %>%
  left_join(fig1_data, by = c("Analysis_Group", "Wrapped_GO_Name", "Facet_Group")) %>%
  mutate(
    neg_log_p = replace_na(neg_log_p, 0),
    Wrapped_Facet = str_wrap(Facet_Group, width = 15)
  )

h1 <- max(4, (length(unique(grid_fig1$Wrapped_GO_Name)) * 0.3) + 2)

plot_fig1 <- ggplot(grid_fig1, aes(x = Analysis_Group, y = reorder(Wrapped_GO_Name, neg_log_p))) +
  geom_tile(aes(fill = neg_log_p), color = "white", lwd = 0.4) +
  scale_fill_viridis_c(option = "mako", direction = -1, name = "-log10(p)") +
  facet_grid(Wrapped_Facet ~ ., scales = "free_y", space = "free_y") +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, face = "bold"),
    axis.text.y = element_text(size = 8.5, lineheight = 0.85, face = "bold"),
    panel.grid = element_blank(),
    strip.text.y = element_text(angle = 90, face = "bold", size = 9.5, lineheight = 0.9, vjust = 0.5),
    strip.background = element_rect(fill = "gray95", color = "gray85", linewidth = 0.5),
    plot.title = element_text(face = "bold", size = 12)
  ) +
  labs(title = "Figure 1: Regulatory Amino Acid Infrastructure", x = "Cluster Wave", y = NULL)

print(plot_fig1)

ggsave(file.path(sub_plots_dir, "figure_1_bcaa_and_cyclic.pdf"), plot = plot_fig1, width = 9, height = h1, units = "in")
```

```{r TODO-FIX-THIS}
# ----- fig 3 v
message("Generating Figure 3 (Figure 1 + Other Amino Acids with General Structural Terms)...")

fig3_data <- raw_high_res_df %>%
  # UPDATED FILTER: Added general structural catchphrases to the root query
  filter(str_detect(GO_Name, "(?i)amino acid|peptide|aminotransferase|isoleucine|leucine|valine|branched-chain|branched chain|phenylalanine|tyrosine|tryptophan|histidine|shikimate|chorismate|cyclic|aromatic")) %>%
  
  # Filter out explicit Nitrogen and Sulfur terms to keep this subset mathematically unmixed
  filter(!ID %in% fig2_data$ID) %>%
  
  mutate(
    Facet_Group = case_when(
      # UPDATED CONDITION: Catches both hyphenated and unhyphenated variant terms
      str_detect(GO_Name, "(?i)isoleucine|leucine|valine|branched-chain|branched chain") ~ "Branched-Chain (BCAA)",
      
      # UPDATED CONDITION: Added specific explicit catches for general structural terms
      str_detect(GO_Name, "(?i)phenylalanine|tyrosine|tryptophan|histidine|aromatic|cyclic|shikimate|chorismate") ~ "Cyclic & Aromatic",

      # NEW HIGH-RESOLUTION SPLITS TO ELIMINATE HOUSEKEEPING BLOCKS:
      str_detect(GO_Name, "(?i)polyamine|putrescine|spermidine|cadaverine") ~ "Polyamines & Stress Diamines",
      str_detect(GO_Name, "(?i)aspartate|asparagine|lysine|diaminopimelate") ~ "Aspartate-Family Super-Cluster",
      str_detect(GO_Name, "(?i)glycine|serine|threonine|alanine|pyruvate-derived") ~ "Small Aliphatics (Precursors)",
      str_detect(GO_Name, "(?i)proline|ectoine|osmoprotect|osmotic") ~ "Osmoprotection & Proline",
      TRUE ~ "General Peptide & AA Transport"
    )
  )

grid_fig3 <- expand_grid(
  Analysis_Group = unique(raw_high_res_df$Analysis_Group),
  Wrapped_GO_Name = unique(fig3_data$Wrapped_GO_Name)
) %>%
  left_join(distinct(fig3_data, Wrapped_GO_Name, Facet_Group), by = "Wrapped_GO_Name") %>%
  left_join(fig3_data, by = c("Analysis_Group", "Wrapped_GO_Name", "Facet_Group")) %>%
  mutate(
    neg_log_p = replace_na(neg_log_p, 0),
    Wrapped_Facet = str_wrap(Facet_Group, width = 15)
  )

h3 <- max(4.5, (length(unique(grid_fig3$Wrapped_GO_Name)) * 0.3) + 2)

plot_fig3 <- ggplot(grid_fig3, aes(x = Analysis_Group, y = reorder(Wrapped_GO_Name, neg_log_p))) +
  geom_tile(aes(fill = neg_log_p), color = "white", lwd = 0.4) +
  scale_fill_viridis_c(option = "mako", direction = -1, name = "-log10(p)") +
  facet_grid(Wrapped_Facet ~ ., scales = "free_y", space = "free_y") +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, face = "bold"),
    axis.text.y = element_text(size = 8.5, lineheight = 0.85, face = "bold"),
    panel.grid = element_blank(),
    strip.text.y = element_text(angle = 90, face = "bold", size = 9.5, lineheight = 0.9, vjust = 0.5),
    strip.background = element_rect(fill = "gray95", color = "gray85", linewidth = 0.5),
    plot.title = element_text(face = "bold", size = 12)
  ) +
  labs(title = "Figure 3: Comprehensive Profile & Structural Modulators", x = "Cluster Wave", y = NULL)

print(plot_fig3)

ggsave(file.path(sub_plots_dir, "figure_3_expanded_amino_landscape.png"), plot = plot_fig3, width = 9.5, height = h3, units = "in")

message("All three distinct figures have been processed and saved successfully.")
```

```{r TODO-FIX_THIS-clustered-axes-aa}


# Ensure results directory exists
if (!dir.exists(sub_plots_dir)) dir.create(sub_plots_dir, recursive = TRUE)

message("Executing complete Figure 3 Pipeline: Expanded Landscape with Bi-Clustered Axes (Mako Palette)...")

# ----- extract and facet data ----- 
fig3_data <- raw_high_res_df %>%
  # Expanded query to include generic branched chain, cyclic, and aromatic terms
  filter(str_detect(GO_Name, "(?i)amino acid|peptide|aminotransferase|isoleucine|leucine|valine|branched-chain|branched chain|phenylalanine|tyrosine|tryptophan|histidine|shikimate|chorismate|cyclic|aromatic")) %>%
  
  # Keep this pool mathematically isolated from the explicit Nitrogen/Sulfur subsets
  filter(!ID %in% fig2_data$ID) %>%
  
  mutate(
    Facet_Group = case_when(
      str_detect(GO_Name, "(?i)isoleucine|leucine|valine|branched-chain|branched chain") ~ "Branched-Chain (BCAA)",
      str_detect(GO_Name, "(?i)phenylalanine|tyrosine|tryptophan|histidine|aromatic|cyclic|shikimate|chorismate") ~ "Cyclic & Aromatic",
      str_detect(GO_Name, "(?i)polyamine|putrescine|spermidine|cadaverine") ~ "Polyamines & Stress Diamines",
      str_detect(GO_Name, "(?i)aspartate|asparagine|lysine|diaminopimelate") ~ "Aspartate-Family Super-Cluster",
      str_detect(GO_Name, "(?i)glycine|serine|threonine|alanine|pyruvate-derived") ~ "Small Aliphatics (Precursors)",
      str_detect(GO_Name, "(?i)proline|ectoine|osmoprotect|osmotic") ~ "Osmoprotection & Proline",
      TRUE ~ "General Peptide & AA Transport"
    )
  )

# Build comprehensive observation matrix network 
grid_fig3 <- expand_grid(
  Analysis_Group = unique(raw_high_res_df$Analysis_Group),
  Wrapped_GO_Name = unique(fig3_data$Wrapped_GO_Name)
) %>%
  left_join(distinct(fig3_data, Wrapped_GO_Name, Facet_Group), by = "Wrapped_GO_Name") %>%
  left_join(fig3_data, by = c("Analysis_Group", "Wrapped_GO_Name", "Facet_Group")) %>%
  mutate(neg_log_p = replace_na(neg_log_p, 0))

# scale output height based on final row counts
h3 <- max(4.5, (length(unique(grid_fig3$Wrapped_GO_Name)) * 0.28) + 2)

# pivot to wide format
matrix_data <- grid_fig3 %>%
  dplyr::select(Wrapped_GO_Name, Analysis_Group, neg_log_p) %>%
  distinct() %>%
  pivot_wider(names_from = Analysis_Group, values_from = neg_log_p, values_fill = 0) %>%
  as.data.frame()

# Relocate row labels to formal row identifier metadata slots
rownames(matrix_data) <- matrix_data$Wrapped_GO_Name
matrix_data$Wrapped_GO_Name <- NULL
heatmap_matrix <- as.matrix(matrix_data)

# ----- colour faceting ---
row_metadata <- data.frame(Wrapped_GO_Name = rownames(heatmap_matrix)) %>%
  left_join(distinct(grid_fig3, Wrapped_GO_Name, Facet_Group), by = "Wrapped_GO_Name")

row_anno <- rowAnnotation(
  `Structural Group` = row_metadata$Facet_Group,
  col = list(`Structural Group` = c(
    "Branched-Chain (BCAA)"          = "darkred",
    "Cyclic & Aromatic"              = "darkblue",
    "General Peptide & AA Transport" = "grey"
  )),
  show_annotation_name = FALSE
)

# ----- define mako so it looks like our other figures
max_val <- max(heatmap_matrix, na.rm = TRUE)
if(max_val == 0) max_val <- 1 

color_fun <- colorRamp2(
  seq(0, max_val, length.out = 5), 
  rev(viridis::mako(5, direction = 1))
)

# ----- hierarchical heatmap, baby! ---
clustered_heatmap <- Heatmap(
  heatmap_matrix,
  name = "-log10(p)",
  col = color_fun,
  
  # cluster your axes
  cluster_rows = TRUE,          
  cluster_columns = TRUE,       
  clustering_distance_rows = "euclidean",
  clustering_method_rows = "complete",
  clustering_distance_columns = "euclidean",
  clustering_method_columns = "complete",
  
  # layout
  row_names_gp = gpar(fontface = "bold", fontsize = 8.5),
  column_names_gp = gpar(fontface = "bold", fontsize = 10, rotation = 45),
  rect_gp = gpar(col = "white", lwd = 0.4), 
  
  # structural groups as a left color block
  left_annotation = row_anno,
  
  column_title = "Figure 3: Bi-Clustered Structural Profile Matrix (Mako Palette)",
  column_title_gp = gpar(fontface = "bold", fontsize = 12)
)

# Save 
png(
  filename = file.path(sub_plots_dir, "figure_3_bi_clustered_mako_landscape.png"),
  width = 11, height = h3, units = "in", res = 300
)
draw(clustered_heatmap)
dev.off()

# Render to active IDE/Markdown notebook panel
print(clustered_heatmap)

message("Figure 3 fully generated and saved successfully.")
```



```{r}
# Ensure subfolder exists
lipid_plots_dir <- file.path(results_dir, "targeted_subcategories")
if (!dir.exists(lipid_plots_dir)) dir.create(lipid_plots_dir, recursive = TRUE)

# Prepare base dataset metrics
raw_lipid_df <- compiled_elim_results %>%
  mutate(
    neg_log_p = -log10(qvalue),
    Wrapped_GO_Name = str_wrap(as.character(GO_Name), width = 40)
  )

# ----- membrane synthesis and adaptation ----- 
message("Compiling Figure 4: Bacterial Membrane Core Synthesis...")

fig4_data <- raw_lipid_df %>%
  mutate(
    Facet_Group = case_when(
      str_detect(GO_Name, "(?i)unsaturated|cyclopropane|desaturase|fluidity") ~ "Membrane Fluidity & Stress Adaptation",
      str_detect(GO_Name, "(?i)phosphatidyl|cardiolipin|glycerolipid|phospholipid biosyn") ~ "Phospholipid Head Groups (Charge Control)",
      str_detect(GO_Name, "(?i)fatty acid biosyn|acyl-carrier|[^catabolic] biosynthetic") ~ "FAS II Core Elongation",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(Facet_Group))

grid_fig4 <- expand_grid(
  Analysis_Group = unique(raw_lipid_df$Analysis_Group),
  Wrapped_GO_Name = unique(fig4_data$Wrapped_GO_Name)
) %>%
  left_join(distinct(fig4_data, Wrapped_GO_Name, Facet_Group), by = "Wrapped_GO_Name") %>%
  left_join(fig4_data, by = c("Analysis_Group", "Wrapped_GO_Name", "Facet_Group")) %>%
  mutate(
    neg_log_p = replace_na(neg_log_p, 0),
    Wrapped_Facet = str_wrap(Facet_Group, width = 15)
  )

h4 <- max(4, (length(unique(grid_fig4$Wrapped_GO_Name)) * 0.3) + 2)

plot_fig4 <- ggplot(grid_fig4, aes(x = Analysis_Group, y = reorder(Wrapped_GO_Name, neg_log_p))) +
  geom_tile(aes(fill = neg_log_p), color = "white", lwd = 0.4) +
  scale_fill_viridis_c(option = "rocket", direction = -1, name = "-log10(p)") +
  facet_grid(Wrapped_Facet ~ ., scales = "free_y", space = "free_y") +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, face = "bold"),
    axis.text.y = element_text(size = 8.5, lineheight = 0.85, face = "bold"),
    panel.grid = element_blank(),
    strip.text.y = element_text(angle = 90, face = "bold", size = 9.5, lineheight = 0.9, vjust = 0.5),
    strip.background = element_rect(fill = "gray95", color = "gray85", linewidth = 0.5),
    plot.title = element_text(face = "bold", size = 12)
  ) +
  labs(title = "Figure 4: Bacterial Membrane Biogenesis & Fluidity Control", x = "Cluster Wave", y = NULL)

ggsave(file.path(lipid_plots_dir, "figure_4_bacterial_membrane_core.pdf"), plot = plot_fig4, width = 9.5, height = h4, units = "in")
print(plot_fig4)

# ----- lipid catabolism and envelope ----- 
message("Compiling Figure 5: Lipid Degradation & Outer Envelope Components...")

fig5_data <- raw_lipid_df %>%
  mutate(
    Facet_Group = case_when(
      str_detect(GO_Name, "(?i)beta-oxidation|fatty acid catabolic|acyl-coa dehydro") ~ "Beta-Oxidation (Carbon Sourcing)",
      str_detect(GO_Name, "(?i)lipid A|lipopolysaccharide|teichoic|outer membrane lipid") ~ "Envelope Armor (LPS / Anchors)",
      str_detect(GO_Name, "(?i)lipid transport|lipid translocation|flippase|msbA") ~ "Lipid Translocation & Export",
      TRUE ~ NA_character_
    )
  ) %>%
  # Filter for general lipid terms that weren't captured by Figure 4 or the explicit filters above
  bind_rows(
    raw_lipid_df %>%
      filter(str_detect(GO_Name, "(?i)lipid|fatty acid")) %>%
      filter(!ID %in% fig4_data$ID) %>%
      mutate(Facet_Group = "General Lipid Housekeeping")
  ) %>%
  # Filter down to eliminate missing data duplicates
  filter(!is.na(Facet_Group)) %>%
  distinct(Analysis_Group, ID, .keep_all = TRUE)

grid_fig5 <- expand_grid(
  Analysis_Group = unique(raw_lipid_df$Analysis_Group),
  Wrapped_GO_Name = unique(fig5_data$Wrapped_GO_Name)
) %>%
  left_join(distinct(fig5_data, Wrapped_GO_Name, Facet_Group), by = "Wrapped_GO_Name") %>%
  left_join(fig5_data, by = c("Analysis_Group", "Wrapped_GO_Name", "Facet_Group")) %>%
  mutate(
    neg_log_p = replace_na(neg_log_p, 0),
    Wrapped_Facet = str_wrap(Facet_Group, width = 15)
  )

h5 <- max(4, (length(unique(grid_fig5$Wrapped_GO_Name)) * 0.3) + 2)

plot_fig5 <- ggplot(grid_fig5, aes(x = Analysis_Group, y = reorder(Wrapped_GO_Name, neg_log_p))) +
  geom_tile(aes(fill = neg_log_p), color = "white", lwd = 0.4) +
  scale_fill_viridis_c(option = "mako", direction = -1, name = "-log10(p)") +
  facet_grid(Wrapped_Facet ~ ., scales = "free_y", space = "free_y") +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, face = "bold"),
    axis.text.y = element_text(size = 8.5, lineheight = 0.85, face = "bold"),
    panel.grid = element_blank(),
    strip.text.y = element_text(angle = 90, face = "bold", size = 9.5, lineheight = 0.9, vjust = 0.5),
    strip.background = element_rect(fill = "gray95", color = "gray85", linewidth = 0.5),
    plot.title = element_text(face = "bold", size = 12)
  ) +
  labs(title = "Figure 5: Lipid Catabolism & Outer Envelope Architecture", x = "Cluster Wave", y = NULL)

print(plot_fig5)

ggsave(file.path(lipid_plots_dir, "figure_5_lipid_catabolism_and_armor.pdf"), plot = plot_fig5, width = 9.5, height = h5, units = "in")

message("Lipid subcategory pipeline executed. Figures 4 and 5 saved.")
```

TO DO:
- add the additional annotation information formats used in the mahal distance
- fix cluster selection parameters