---
title: "Mahalanobis distances"
author: "Kira Goff"
date: "2026-06-16"
output: html_document
---

```{r}

```

```{r load-libraries}
# data manipulation
library(dplyr)          # Essential for data wrangling (mutate, select, %>% etc.)
library(tidyr)          # Used for reshaping data (pivot_longer, pivot_wider)

# data visualization
library(ggplot2)        # Core plotting engine
library(scales)         # Formatting scales on plot axes (percentages, currency, etc.)
library(ggrepel)        # Prevents text labels from overlapping on plots
library(plotly)         # Converts ggplot2 plots into interactive HTML graphics

# bioinformatics analysis
library(clusterProfiler) # Core functional enrichment analysis tool
library(KEGGREST)       # Client interface to the KEGG REST API
```

```{r define-paths}
base_dir <- "./"
base_results <- "./results"
results_dir <- "./results/05-mahalanobis-distance"

if (!dir.exists(results_dir)) {
  dir.create(results_dir, recursive = TRUE)
}

```

```{r check-mahalanobis-histogram}
mahal_dist <- df_final_clusters$mahalanobis_dist
mahal_max <- max(mahal_dist)
mahal_full <- hist(mahal_dist)
print(mahal_full)
mahal_26 <- hist(mahal_dist, breaks = 26)
print(mahal_26)
```
```{r tight-mahalanobic-hist}
mahal_tight_dist <- df_final_clusters %>%
  filter(mahalanobis_dist <= 10)
mahal_tight <- mahal_tight_dist$mahalanobis_dist
hist(mahal_tight)

```

```{r new-runs-on-existing-dataframes}

# 1. Filter for high-leverage Mahalanobis outliers that show split fuzzy membership
# Adjust thresholds dynamically based on your dataset's specific distributions
bridge_drivers <- df_final_clusters %>%
  filter(
#    is_driver == "High-Confidence" & 
    membership_score < 0.50 # Low primary assignment score = high multi-cluster uncertainty
  ) %>%
  arrange(desc(mahalanobis_dist))

# 2. Left-join with your operon and pathway annotation dataset
annotated_bridge_drivers <- bridge_drivers %>%
  left_join(anno_operon_clean, by = "Gene") %>%
  select(
    Gene, 
    cluster, 
    membership_score, 
    mahalanobis_dist, 
    operon, 
    Product, 
    KO_def, 
    KEGG_Pathway, 
    KEGG_path_def
  )

# Print a summary report of your top network-shifting bridge genes
message(sprintf("Identified %d high-leverage multi-mapping bridge drivers.", nrow(annotated_bridge_drivers)))
print(head(annotated_bridge_drivers, 10))

# 3. Optional: Export this prioritized gene list for separate pathway enrichment loops
write.csv(annotated_bridge_drivers, file.path(results_dir,"priority_multivariate_bridge_drivers.csv"), row.names = FALSE)
```


```{r graph-mahal-vs-cluster-membership}
# 1. Recreate your Glasbey color mapping table
k_final <- k_used
base_colors <- as.vector(pals::glasbey(k_final))

color_map <- data.frame(
  cluster    = 1:k_final,  # Numeric matching your df_final_clusters data layout
  base_color = base_colors,
  stringsAsFactors = FALSE
)

# 2. Merge mapping and calculate continuous opacity scales based on Mahalanobis distance
df_topology_data <- df_final_clusters %>%
  mutate(cluster = as.numeric(as.character(cluster))) %>% # Force numeric alignment for join
  inner_join(color_map, by = "cluster") %>%
  mutate(
    # Normalize Mahalanobis distance between 0.25 (faint background) and 1.0 (max driver peak)
    norm_dist = scales::rescale(mahalanobis_dist, to = c(0.25, 1.0)),
    
    # Convert normalized distance to hex alpha values (0 to 255)
    hex_alpha = sprintf("%02X", as.integer(norm_dist * 255)),
    
    # Stitch base color together with its custom continuous alpha value
    final_color = paste0(base_color, hex_alpha)
  )

# legend for custom colour scheme
legend_guide_map <- df_topology_data %>%
  select(cluster, base_color) %>%
  distinct() %>%
  arrange(cluster)


p_topology_static <- ggplot(df_topology_data, aes(
  x = membership_score,
  y = mahalanobis_dist,
  color = final_color,
  text = paste0(
    "<b>Gene:</b> ", Gene, "<br>",
    "<b>Fuzzy Cluster:</b> ", cluster, "<br>",
    "<b>Membership Score:</b> ", round(membership_score, 3), "<br>",
    "<b>Mahalanobis Dist:</b> ", round(mahalanobis_dist, 2)
  )
)) +
  geom_point(aes(size = mahalanobis_dist), alpha = 1.0) +
  
  # Set horizontal/vertical topology guides
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "gray60", size = 0.4) +
  
  scale_size_continuous(range = c(1.0, 3.5), guide = "none") +
  
  # map our unified breaks and labels vectors to keep lengths identical
  scale_color_identity(
    name   = "Cluster",
    guide  = "legend",
    breaks = legend_guide_map$base_color,
    labels = legend_guide_map$cluster
  ) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Continuous Driver Topology Across Fuzzy Clusters",
    subtitle = "Solid opacity mirrors strong Mahalanobis leverage; Background context fades away",
    x = "Primary Cluster Membership Score (Lower = Higher Multi-Mapping / Bridge Multi-Tasking)",
    y = "Mahalanobis Distance (Multivariate Feature Leverage)"
  ) +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )

# Save and export
figname_topology <- sprintf("driver-topology-%.0fclust-lg2fc%.1f-memb%.2f.png", k_final, r_thresh, score_thresh)
ggsave(file.path(results_dir, figname_topology), plot = p_topology_static, width = 8, height = 6, dpi = 300)

p_topology_static
```

```{r bridge-distance}
df_clusters_prep <- df_final_clusters %>%
  mutate(
    Gene = as.character(Gene),
    cluster = as.numeric(as.character(cluster))
  )

anno_prep <- anno_operon_clean %>%
  mutate(Gene = as.character(Gene))

# 2. Build the Comprehensive Continuous Topology Matrix
master_topology_matrix <- df_clusters_prep %>%
  left_join(anno_prep, by = "Gene") %>%
  mutate(
    # High score = High Mahalanobis Distance AND Low Primary Membership (Split variance)
    Bridge_Priority_Score = mahalanobis_dist * (1 - membership_score)
  ) %>%
  select(
    Gene,
    cluster,
    membership_score,
    mahalanobis_dist,
    Bridge_Priority_Score,
    model_LFC_06h,
    model_LFC_24h,
    operon,
    Product,
    KO,
    KO_def,
    KEGG_Pathway,
    KEGG_path_def,
    GO_bak
  ) %>%
  # Sort the genome so the most impactful network-shifters are at the top
  arrange(desc(Bridge_Priority_Score))

# 3. Verify the extraction metrics
message("--- STEP 1 VERIFICATION ---")
message(sprintf("Master Topology Matrix built successfully with %d genes.", nrow(master_topology_matrix)))
message(sprintf("Missing annotations caught: %d genes lack KEGG pathway definitions.", sum(is.na(master_topology_matrix$KEGG_Pathway))))

# Preview the top 5 highest priority continuous genes
print(head(master_topology_matrix, 5))
```


The Benjamini-Hochberg p.adj value assumes that there are no enrichments at all in a dataset (all null)
The Storey q value uses p-value histograms of the dataset to calculate the appropriate proportion of null tests
Because we have thousands of genes and terms, the q value is more accurate for our dataset

I'm splitting KEGG and COG because of different source for the additional annotation
```{r kegg-based-enrichment}
# define variables
qtable <- 0.1 
mahal_cut <- 4      # Globally flags multivariate outliers on 06h vs 24h scatterplot
fc_cut    <- 0.5    # Specifically captures upregulation at 6h OR 24h

target_databases <- c("KO", "kPathway", "kModule", "kRxn")

# Setup subfolder for reshaped outputs
subfolder_dir <- file.path(results_dir, "reshaped_outputs")
if (!dir.exists(subfolder_dir)) dir.create(subfolder_dir, recursive = TRUE)

# Keywords to filter out housekeeping/structural terms
housekeeping_terms <- c(
  "\\b5s\\b", "\\b16s\\b", "\\b23s\\b", 
  "rrna", 
  "ribosomal protein", 
  "^ribosomal large subunit$", 
  "^ribosomal small subunit$",
  "sigma factor", "rna polymerase"
)

# --- helper fn for annotation ---
get_kegg_layer_metadata <- function(ids, db_type) {
  if (length(ids) == 0) return(NULL)
  message(sprintf("--> Querying KEGG API for %d enriched %s terms...", length(unique(ids)), db_type))
  
  results <- list()
  
  for (id in unique(ids)) {
    clean_id <- gsub("ko:|path:|md:|rn:", "", id)
    
    tryCatch({
      kegg_data <- keggGet(clean_id)[[1]]
      
      pathways <- NA; modules <- NA; reactions <- NA; class_hierarchy <- NA
      
      if (!is.null(kegg_data$PATHWAY)) {
        pathways <- paste(names(kegg_data$PATHWAY), kegg_data$PATHWAY, sep=": ", collapse=" | ")
      }
      if (!is.null(kegg_data$MODULE)) {
        modules <- paste(names(kegg_data$MODULE), kegg_data$MODULE, sep=": ", collapse=" | ")
      }
      if (!is.null(kegg_data$REACTION)) {
        reactions <- paste(kegg_data$REACTION, collapse=", ")
      }
      if (!is.null(kegg_data$CLASS)) {
        class_hierarchy <- kegg_data$CLASS
      }
      
      if (db_type == "KO") {
        results[[clean_id]] <- data.frame(ID = id, Pathways = pathways, Modules = modules, Reactions = reactions, stringsAsFactors = FALSE)
      } else if (db_type == "kModule") {
        results[[clean_id]] <- data.frame(ID = id, Parent_Pathways = pathways, Functional_Class = class_hierarchy, stringsAsFactors = FALSE)
      } else if (db_type == "kPathway") {
        results[[clean_id]] <- data.frame(ID = id, Pathway_Class = class_hierarchy, stringsAsFactors = FALSE)
      } else if (db_type == "kRxn") {
        definition <- ifelse(!is.null(kegg_data$DEFINITION), kegg_data$DEFINITION, NA)
        enzymes    <- ifelse(!is.null(kegg_data$ENZYME), paste(kegg_data$ENZYME, collapse=", "), NA)
        results[[clean_id]] <- data.frame(ID = id, Equation = definition, Enzymes_EC = enzymes, Modules = modules, Pathways = pathways, stringsAsFactors = FALSE)
      }
      
    }, error = function(e) {
      results[[clean_id]] <- data.frame(ID = id, stringsAsFactors = FALSE)
    })
  }
  return(bind_rows(results))
}

# --- loop through databases ---

for (target_db in target_databases) {
  
  message(sprintf("\n=== Processing Enrichment for Database: %s ===", target_db))
  
  # 1. Dynamically assign configuration variables
  db_label          <- db_configs[[target_db]]$db_label
  database_tsv_path <- db_configs[[target_db]]$database_tsv_path
  db_term_col       <- db_configs[[target_db]]$db_term_col
  db_definition_col <- db_configs[[target_db]]$db_definition_col
  
  # Load term definitions safely
  clean_defs <- read.delim(database_tsv_path, sep = "\t", stringsAsFactors = FALSE) %>%
    select(ID = !!sym(db_term_col), Description = !!sym(db_definition_col)) %>%
    distinct()
  
  # 2. Define database-specific universe
  raw_universe_genes <- rownames(dds)
  
  t2g_base <- anno_operon_clean %>%
    filter(Gene %in% raw_universe_genes) %>%
    select(!!sym(db_label), Gene) %>%
    drop_na(!!sym(db_label))
  
  if (db_label == "COG_cat") {
    t2g_split <- t2g_base %>%
      mutate(!!sym(db_label) := gsub("[ ,]", "", !!sym(db_label))) %>%
      separate_rows(!!sym(db_label), sep = "")
  } else {
    t2g_split <- t2g_base %>%
      separate_rows(!!sym(db_label), sep = ",\\s*")
    
    if (target_db == "kPathway") {
      t2g_split <- t2g_split %>%
        filter(grepl("^map\\d+", !!sym(db_label)))
    }
  }
  
  # Filter out structural/housekeeping annotations
  t2g <- t2g_split %>%
    rename(ID = !!sym(db_label)) %>%
    left_join(clean_defs, by = "ID") %>%
    filter(!grepl(paste(housekeeping_terms, collapse = "|"),
                  Description,
                  ignore.case = TRUE
    )) %>%
    select(term_id = ID, gene_id = Gene) %>%
    filter(term_id != "" & term_id != "NA") %>%
    distinct()
  
  universe_vector <- unique(t2g$gene_id)
  
  # 3. Dynamic Driver Gene Generation (Scoped to current universe background)
  driver_genes <- master_topology_matrix %>%
    filter(
      mahalanobis_dist > mahal_cut,
      model_LFC_06h >= fc_cut | model_LFC_24h >= fc_cut
    ) %>%
    filter(Gene %in% universe_vector) %>% 
    select(Gene) %>%
    distinct()
  
  flagged_genes_vector <- driver_genes %>% pull(Gene)
  
  # Skip to next DB if no driver genes map to this database's universe
  if (length(intersect(flagged_genes_vector, universe_vector)) == 0) {
    message(sprintf("Skipping %s: No flagged driver genes exist in this functional annotation universe.", target_db))
    next
  }
  
  # 4. Enrichment Analysis
  enrichment_results_list <- list()
  
  # OPTIMIZED: pvalueCutoff = 1 lets Storey's q-value exclusively control significance thresholds
  # cluster enricher applies p to p and p.adj
  ego <- enricher(
    gene         = flagged_genes_vector,
    universe     = universe_vector,
    TERM2GENE    = t2g,
    pvalueCutoff = 1,
    qvalueCutoff = qtable,
    minGSSize    = 2,
    maxGSSize    = 500
  )
  
  if (!is.null(ego) && nrow(ego@result) > 0) {
    res_df <- as.data.frame(ego) %>%
      mutate(Analysis_Group = "Mahalanobis_Outliers")
    enrichment_results_list[["Mahalanobis_Outliers"]] <- res_df
  }
  
  # 5. Annotate and save
  if (length(enrichment_results_list) > 0) {
    final_enrichment_df <- bind_rows(enrichment_results_list) %>%
      select(-any_of("Description")) %>%
      left_join(clean_defs, by = "ID") %>%
      select(Analysis_Group, ID, Description, Count, everything())
    
    # INTERCEPT & EXTEND: Only target explicit KEGG prefixes for live metadata injection
    if (target_db %in% c("KO", "kPathway", "kModule", "kRxn")) {
      kegg_meta <- get_kegg_layer_metadata(final_enrichment_df$ID, target_db)
      
      if (!is.null(kegg_meta) && nrow(kegg_meta) > 0) {
        final_enrichment_df <- final_enrichment_df %>%
          left_join(kegg_meta, by = "ID")
      }
    }
    
    # re-order for ease of visual interpretation
    final_enrichment_df <- final_enrichment_df %>%
      relocate(qvalue, .after = ID) %>%
      select(Analysis_Group, ID, qvalue, Description, 
               any_of(c("Equation", "Enzymes_EC", "Modules", "Pathways", "Parent_Pathways", "Functional_Class", "Pathway_Class")), 
               everything())
    
    # pull file information for naming
    n_enriched_terms <- nrow(final_enrichment_df)
    n_enriched_genes <- final_enrichment_df %>%
      separate_rows(geneID, sep = "/") %>%
      pull(geneID) %>%
      n_distinct()
    
    # save names
    base_name <- sprintf("%s_enrichment_mahal%.1f_lfc%.2f_q%.1f_nterms%d_ngenes%d", 
                         db_label, mahal_cut, fc_cut, qtable, n_enriched_terms, n_enriched_genes)
    
    # File 1: Standard output 
    original_out_path <- file.path(results_dir, paste0(base_name, ".tsv"))
    write.table(final_enrichment_df %>% select(-any_of("Analysis_Group")), original_out_path, sep = "\t", row.names = FALSE, quote = FALSE)
    
    # File 2: Long format by gene 
    tidy_long_df <- final_enrichment_df %>%
      separate_rows(geneID, sep = "/") %>%
      rename(Gene_ID = geneID) %>%
      select(Analysis_Group, Gene_ID, ID, everything()) %>%
      arrange(Analysis_Group, Gene_ID, p.adjust)
    
    long_out_path <- file.path(subfolder_dir, paste0(base_name, "_long_by_gene.tsv"))
    write.table(tidy_long_df, long_out_path, sep = "\t", row.names = FALSE, quote = FALSE)
    
    # File 3: Matrix profile wide
    matrix_wide_df <- final_enrichment_df %>%
      mutate(Enriched = 1) %>%
      select(ID, Analysis_Group, Enriched) %>%
      distinct() %>%
      pivot_wider(
        names_from = Analysis_Group,
        values_from = Enriched,
        values_fill = 0
      ) %>%
      arrange(ID)
    
    wide_out_path <- file.path(subfolder_dir, paste0(base_name, "_profile_matrix.tsv"))
    write.table(matrix_wide_df, wide_out_path, sep = "\t", row.names = FALSE, quote = FALSE)
    
    # Console status logs
    message(sprintf("Successfully exported outputs for %s:", db_label))
    message(sprintf("  -> %s", basename(original_out_path)))
    
  } else {
    message(sprintf("No significant enrichment found for %s (Mahal > %.1f, LFC >= %.2f, q-value <= %.2f).", 
                    db_label, mahal_cut, fc_cut, qtable))
  }
}

message("\n*** All database enrichment loops completed! ***")
```

```{r cog-enrichment}
# define variables
target_databases <- c("COG", "COGcat")
qtable <- 0.1 
mahal_cut <- 4      # Threshold for outliers on 06h vs 24h scatterplot
fc_cut    <- 0.5    # Specifically captures upregulation at 6h OR 24h

# Setup subfolder for reshaped outputs
subfolder_dir <- file.path(results_dir, "reshaped_outputs")
if (!dir.exists(subfolder_dir)) dir.create(subfolder_dir, recursive = TRUE)

# Keywords to filter out housekeeping/structural terms
housekeeping_terms <- c(
  "\\b5s\\b", "\\b16s\\b", "\\b23s\\b", 
  "rrna", 
  "ribosomal protein", 
  "^ribosomal large subunit$", 
  "^ribosomal small subunit$",
  "sigma factor", "rna polymerase"
)

# Path to the comprehensive COG master lookup file for additional downstream annotation
cog_master_lookup_path <- "/mnt/work-drive/databases/COG-KEGG-data/cog-def.tsv"

# --- databases ---

for (target_db in target_databases) {
  
  message(sprintf("\n=== Processing Enrichment for Database: %s ===", target_db))
  
  # 1. database variables
  db_label          <- db_configs[[target_db]]$db_label
  database_tsv_path <- db_configs[[target_db]]$database_tsv_path
  db_term_col       <- db_configs[[target_db]]$db_term_col
  db_definition_col <- db_configs[[target_db]]$db_definition_col
  
  # Load term definitions safely
  clean_defs <- read.delim(database_tsv_path, sep = "\t", stringsAsFactors = FALSE) %>%
    select(ID = !!sym(db_term_col), Description = !!sym(db_definition_col)) %>%
    distinct()
  
  # 2. Define database-specific universe
  raw_universe_genes <- rownames(dds)
  
  t2g_base <- anno_operon_clean %>%
    filter(Gene %in% raw_universe_genes) %>%
    select(!!sym(db_label), Gene) %>%
    drop_na(!!sym(db_label))
  
  if (db_label == "COG_cat") {
    t2g_split <- t2g_base %>%
      mutate(!!sym(db_label) := gsub("[ ,]", "", !!sym(db_label))) %>%
      separate_rows(!!sym(db_label), sep = "")
  } else {
    t2g_split <- t2g_base %>%
      separate_rows(!!sym(db_label), sep = ",\\s*")
  }
  
  # Filter out structural/housekeeping annotations
  t2g <- t2g_split %>%
    rename(ID = !!sym(db_label)) %>%
    left_join(clean_defs, by = "ID") %>%
    filter(!grepl(paste(housekeeping_terms, collapse = "|"),
                  Description,
                  ignore.case = TRUE
    )) %>%
    select(term_id = ID, gene_id = Gene) %>%
    filter(term_id != "" & term_id != "NA") %>%
    distinct()
  
  universe_vector <- unique(t2g$gene_id)
  
  # 3. Dynamic Driver Gene Generation (Scoped to current universe background)
  driver_genes <- master_topology_matrix %>%
    filter(
      mahalanobis_dist > mahal_cut,
      model_LFC_06h >= fc_cut | model_LFC_24h >= fc_cut
    ) %>%
    filter(Gene %in% universe_vector) %>% 
    select(Gene) %>%
    distinct()
  
  flagged_genes_vector <- driver_genes %>% pull(Gene)
  
  # Skip to next DB if no driver genes map to this database's universe
  if (length(intersect(flagged_genes_vector, universe_vector)) == 0) {
    message(sprintf("Skipping %s: No flagged driver genes exist in this functional annotation universe.", target_db))
    next
  }
  
  # 4. Enrichment analysis
  enrichment_results_list <- list()
  
  ego <- enricher(
    gene         = flagged_genes_vector,
    universe     = universe_vector,
    TERM2GENE    = t2g,
    pvalueCutoff = 1,
    qvalueCutoff = qtable,
    minGSSize    = 2,
    maxGSSize    = 500
  )
  
  if (!is.null(ego) && nrow(ego@result) > 0) {
    res_df <- as.data.frame(ego) %>%
      mutate(Analysis_Group = "Mahalanobis_Outliers")
    enrichment_results_list[["Mahalanobis_Outliers"]] <- res_df
  }
  
  # 5. annotate and safe
  if (length(enrichment_results_list) > 0) {
    final_enrichment_df <- bind_rows(enrichment_results_list) %>%
      select(-any_of("Description")) %>%
      left_join(clean_defs, by = "ID")
    
    # POST-HOC EXTRA ANNOTATION: Local layer injection exclusively for COG terms
    if (target_db == "COG" && file.exists(cog_master_lookup_path)) {
      message("--> Mapping local metadata definitions from master cog-def.tsv...")
      
      # Read extra descriptive metadata from master file
      cog_metadata <- read.delim(cog_master_lookup_path, sep = "\t", stringsAsFactors = FALSE) %>%
        select(ID = COG, cFunction, Ccode) %>%
        distinct()
      
      # Read the functional category single-letter definitions file 
      # (maps single-letter Ccode to its verbose phrase ccodedef)
      cog_cat_metadata <- read.delim(db_configs[["COGcat"]]$database_tsv_path, sep = "\t", stringsAsFactors = FALSE) %>%
        select(Ccode = COG_code, ccodedef = cCodeDef) %>%
        distinct()
      
      # Join extra variables to main data frames
      final_enrichment_df <- final_enrichment_df %>%
        left_join(cog_metadata, by = "ID") %>%
        left_join(cog_cat_metadata, by = "Ccode")
    }
    
    
    
    # do the clean up and reorganize thing
    # Moves qvalue up and sets custom positioning for extra COG columns if present
    final_enrichment_reorg <- final_enrichment_df %>%
      relocate(qvalue, .after = ID) %>%
      select(Analysis_Group, ID, qvalue, Description, 
             any_of(c("cFunction", "Ccode", "ccodedef")), 
             Count, everything())
    
    # pull stats for file name
    n_enriched_terms <- nrow(final_enrichment_df)
    n_enriched_genes <- final_enrichment_df %>%
      separate_rows(geneID, sep = "/") %>%
      pull(geneID) %>%
      n_distinct()
    
    # dynamic file naming
    base_name <- sprintf("%s_enrichment_mahal%.1f_lfc%.2f_q%.1f_nterms%d_ngenes%d", 
                         db_label, mahal_cut, fc_cut, qtable, n_enriched_terms, n_enriched_genes)
    
    # Standard output
    original_out_path <- file.path(results_dir, paste0(base_name, ".tsv"))
    write.table(final_enrichment_df %>% select(-any_of("Analysis_Group")), original_out_path, sep = "\t", row.names = FALSE, quote = FALSE)
    
    # File 2: Long format by gene 
    tidy_long_df <- final_enrichment_df %>%
      separate_rows(geneID, sep = "/") %>%
      rename(Gene_ID = geneID) %>%
      select(Analysis_Group, Gene_ID, ID, everything()) %>%
      arrange(Analysis_Group, Gene_ID, p.adjust)
    
    long_out_path <- file.path(subfolder_dir, paste0(base_name, "_long_by_gene.tsv"))
    write.table(tidy_long_df, long_out_path, sep = "\t", row.names = FALSE, quote = FALSE)
    
    # File 3: Matrix wide
    matrix_wide_df <- final_enrichment_df %>%
      mutate(Enriched = 1) %>%
      select(ID, Analysis_Group, Enriched) %>%
      distinct() %>%
      pivot_wider(
        names_from = Analysis_Group,
        values_from = Enriched,
        values_fill = 0
      ) %>%
      arrange(ID)
    
    wide_out_path <- file.path(subfolder_dir, paste0(base_name, "_profile_matrix.tsv"))
    write.table(matrix_wide_df, wide_out_path, sep = "\t", row.names = FALSE, quote = FALSE)
    
    # Console status logs
    message(sprintf("Successfully exported outputs for %s:", db_label))
    message(sprintf("  --> %s", basename(original_out_path)))
    
  } else {
    message(sprintf("No significant enrichment found for %s (Mahal > %.1f, LFC >= %.2f, q-value <= %.2f).", 
                    db_label, mahal_cut, fc_cut, qtable))
  }
}

message("\n*** COG/COGcat functional enrichment runs completed! ***")
```
