---
title: "Untitled"
author: "Kira Goff"
date: "2026-08-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-library}
library(tidyverse)
```

For some reason, some combination of running the facet sorting chunks twice and the figure generation twice results in the correct ordering of the descriptions with a facet.

Just roll with it. 

# STYLE DEFINITIONS

```{r universal-formatting}
colour_scale <- "plasma"
scale_direction <- -1
xgrid_colour <- "grey93"
xgrid_linetype <- "solid"
ygrid_colour <- "grey93"
ygrid_linetype <- "dotted"
x_text <- "Hierarchical Clustering of Fuzzy Clusters"
save_dir <- "/mnt/work-drive/kira/rproj-rnaseq/results2/new-gsea-facet-plots/"
tsv_dir <- "/mnt/work-drive/kira/rproj-rnaseq/results2/04-cluster-enrichment-auto-q-naremove/"
```

```{r pathway-formatting}
y_text_p <- "Pathway Description"
figure_title_p <- "Significantly Enriched KEGG Pathways"
caption_text_p <- "Pathways grouped by functional subcatetory. Significance cutoff q < 0.05."
width_p <- 10
height_p <- 12
```

```{r module-formatting}
y_text_m <- "Module Description"
figure_title_m <- "Significantly Enriched KEGG Modules"
caption_text_m <- "Modules grouped by functional subcatetory. Significance cutoff q < 0.05."
width_m <- 10
height_m <- 12
```

```{r cog-letter-formatting}
y_text_c <- "Letter Code Description"
figure_title_c <- "Significantly Enriched COG Letter Codes"
caption_text_c <- "Letter codes grouped by functional catetory. Significance cutoff q < 0.05."
width_c <- 10
height_c <- 10
```



# KEGG PATHWAYS

## ingestion

```{r import-and-fetch-data}
# read files in
folder_path <- file.path(tsv_dir,"KEGG_Pathway")  
files <- list.files(folder_path, pattern = "\\.csv$", full.names = TRUE) #currently using tsvs named as csv
combined_df <- bind_rows(lapply(files, read.delim, stringsAsFactors = FALSE))

# extract unique pathway IDs and query KEGG REST API
unique_ids <- unique(gsub("path:", "", combined_df$ID))

get_kegg_class <- function(path_id) {
  url <- paste0("https://rest.kegg.jp/get/", path_id)
  tryCatch({
    lines <- readLines(url, warn = FALSE)
    class_line <- lines[grep("^CLASS", lines)]
    if(length(class_line) > 0) {
      sub_cat <- sub("^CLASS\\s+", "", class_line)
      return(data.frame(ID = path_id, Subcategory_Full = sub_cat, stringsAsFactors = FALSE))
    }
  }, error = function(e) NULL)
  return(data.frame(ID = path_id, Subcategory_Full = "Unmapped", stringsAsFactors = FALSE))
}

message("Fetching sub-categories directly from KEGG API...")
kegg_meta_list <- lapply(unique_ids, function(id) {
  res <- get_kegg_class(id)
  Sys.sleep(0.1) 
  return(res)
})

kegg_meta <- bind_rows(kegg_meta_list) %>%
  mutate(Subcategory = sub(".*;\\s*", "", Subcategory_Full))
```

```{r merge-and-override-clusters}
# merge data and force custom overrides when things are at a weird level
combined_df_annotated <- combined_df %>%
  mutate(Clean_ID = gsub("path:", "", ID)) %>%
  left_join(kegg_meta, by = c("Clean_ID" = "ID")) %>%
  mutate(
    Subcategory = replace_na(Subcategory, "Other Pathways"),
    Subcategory = case_when(
      Clean_ID == "map01212" ~ "Lipid metabolism",
      Clean_ID == "map01200" ~ "Carbohydrate metabolism",
      Clean_ID == "map01230" ~ "Amino acid metabolism",
      Clean_ID == "map00410" ~ "Amino acid metabolism", # <--- Beta-alanine mapped here
      TRUE ~ Subcategory
    )
  )

# convert long-format GSEA results into a wide matrix for clustering
cluster_matrix <- combined_df_annotated %>%
  select(Description, Analysis_Group, FoldEnrichment) %>%
  pivot_wider(names_from = Analysis_Group, values_from = FoldEnrichment, values_fill = 0) %>%
  column_to_rownames("Description")

# compute distance and run hierarchical clustering on the Analysis Groups (columns)
dist_matrix <- dist(t(as.matrix(cluster_matrix)))
hc <- hclust(dist_matrix, method = "complete")

# extract the new ordered levels of your Analysis Groups based on the tree
ordered_groups <- hc$labels[hc$order]

# convert Analysis_Group to a factor
combined_df_annotated$Analysis_Group <- factor(
  combined_df_annotated$Analysis_Group, 
  levels = ordered_groups
)
```

Hash out to organize facets alphabetically

```{r cluster-facets}
module_category_matrix <- combined_df_annotated %>%
  group_by(Subcategory, Analysis_Group) %>%
  summarise(FoldEnrichment = mean(FoldEnrichment, na.rm = TRUE), .groups = "drop") %>% #change mean to max etc
  pivot_wider(names_from = Analysis_Group,values_from = FoldEnrichment,values_fill = 0) %>%
  column_to_rownames("Subcategory")

module_category_dist <- dist(as.matrix(module_category_matrix))

module_category_hc <- hclust(module_category_dist,method = "complete")

ordered_module_categories <- module_category_hc$labels[module_category_hc$order]

combined_df_annotated <- combined_df_annotated %>%
  mutate(Subcategory = factor(Subcategory,levels = ordered_module_categories))
# this is not currently reordering within each facet, fix this later, idk package name, maybe tidytext?
combined_df_annotated <- combined_df_annotated %>%
  group_by(Subcategory) %>%
  mutate(
    Description = factor(Description, levels = rev(sort(unique(Description))))
  ) %>%
  ungroup()
```


## graphing

```{r graph-pathways-padj-enrich}

ggplot(combined_df_annotated, aes(x = Analysis_Group, y = Description)) +
  geom_point(aes(size = -log10(qvalue), colour = FoldEnrichment)) +
  scale_colour_viridis_c(option = colour_scale, direction = scale_direction, name = "Fold Enrichment") +
  scale_size_continuous(name = "-log10(qvalue)", breaks = scales::breaks_pretty(n = 4)) +
  
  # Wrap text at ~22 characters so long names break cleanly onto multiple lines
  facet_grid(
    rows = vars(Subcategory), 
    scales = "free_y", 
    space = "free_y",
    labeller = labeller(Subcategory = label_wrap_gen(width = 22))
  ) +
  
  theme_bw(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, face = "bold"),
    
    # Rotate strip labels to horizontal (angle = 0) and left-align them for readability
    strip.text.y = element_text(face = "bold", size = 8, angle = 0, hjust = 0),
    
    strip.background = element_rect(fill = "grey85"),
    panel.grid.major.x = element_line(colour = xgrid_colour, linetype = xgrid_linetype),
    panel.grid.major.y = element_line(colour = ygrid_colour, linetype = ygrid_linetype)
  ) +
  labs(
    x = x_text,
    y = y_text_p,
    title = figure_title_p,
    caption = caption_text_p
  )

ggsave(file = file.path(save_dir,"kegg_pathway_clustered_a.png"), 
       width = width_p, height = height_p, dpi = 300)
```

```{r graph-pathways-count-enrichment}

ggplot(combined_df_annotated, aes(x = Analysis_Group, y = Description)) +
  geom_point(aes(size = FoldEnrichment, colour = Count)) +
  scale_colour_viridis_c(option = colour_scale, direction = scale_direction, name = "Gene Count") +
  scale_size_continuous(name = "Fold Enrichment", breaks = scales::breaks_pretty(n = 4)) +
  
  facet_grid(
    rows = vars(Subcategory), 
    scales = "free_y", 
    space = "free_y",
    labeller = labeller(Subcategory = label_wrap_gen(width = 22))
  ) +
  
  theme_bw(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, face = "bold"),
    strip.text.y = element_text(face = "bold", size = 8, angle = 0, hjust = 0),
    strip.background = element_rect(fill = "grey85"),
    panel.grid.major.x = element_line(colour = xgrid_colour, linetype = xgrid_linetype),
    panel.grid.major.y = element_line(colour = ygrid_colour, linetype = ygrid_linetype)
  ) +
  labs(
    x = x_text,
    y = y_text_p,
    title = figure_title_p,
    caption = caption_text_p
  )

ggsave(file = file.path(save_dir,"kegg_pathway_clustered_b.png"), 
       width = width_p, height = height_p, dpi = 300)
```

```{r graph-pathways-count-enrichment}

ggplot(combined_df_annotated, aes(x = Analysis_Group, y = Description)) +
  geom_point(aes(size = Count, colour = -log10(qvalue))) +
  scale_colour_viridis_c(option = colour_scale, direction = scale_direction, name = "-log10(qvalue)") +
  scale_size_continuous(name = "Gene Count", breaks = scales::breaks_pretty(n = 4)) +
  
  facet_grid(
    rows = vars(Subcategory), 
    scales = "free_y", 
    space = "free_y",
    labeller = labeller(Subcategory = label_wrap_gen(width = 22))
  ) +
  
  theme_bw(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, face = "bold"),
    strip.text.y = element_text(face = "bold", size = 8, angle = 0, hjust = 0),
    strip.background = element_rect(fill = "grey85"),
    panel.grid.major.x = element_line(colour = xgrid_colour, linetype = xgrid_linetype),
    panel.grid.major.y = element_line(colour = ygrid_colour, linetype = ygrid_linetype)
  ) +
  labs(
    x = x_text,
    y = y_text_p,
    title = figure_title_p,
    caption = caption_text_p
  )

ggsave(file = file.path(save_dir,"kegg_pathway_clustered_c.png"), 
       width = width_p, height = height_p, dpi = 300)
```

# KEGG MODULES

## ingestion

```{r}
# read tsv
folder_path <- file.path(tsv_dir,"KEGG_Module") # 
files <- list.files(folder_path, pattern = "\\.csv$", full.names = TRUE) # currently using tsvs saves as csv bc librecalc
combined_mod_df <- bind_rows(lapply(files, read.delim, stringsAsFactors = FALSE))

# extract unique module IDs ("M00001")
unique_mod_ids <- unique(gsub("path:", "", combined_mod_df$ID))

# get module classification from KEGG REST API
get_kegg_module_class <- function(mod_id) {
  url <- paste0("https://rest.kegg.jp/get/", mod_id)
  tryCatch({
    lines <- readLines(url, warn = FALSE)
    class_line <- lines[grep("^CLASS", lines)]
    if(length(class_line) > 0) {
      sub_cat <- sub("^CLASS\\s+", "", class_line)
      return(data.frame(ID = mod_id, Subcategory_Full = sub_cat, stringsAsFactors = FALSE))
    }
  }, error = function(e) NULL)
  return(data.frame(ID = mod_id, Subcategory_Full = "Unmapped", stringsAsFactors = FALSE))
}

message("Fetching module sub-categories directly from KEGG API...")
mod_meta_list <- lapply(unique_mod_ids, function(id) {
  res <- get_kegg_module_class(id)
  Sys.sleep(0.1) 
  return(res)
})

mod_meta <- bind_rows(mod_meta_list) %>%
  mutate(Subcategory = sub(".*;\\s*", "", Subcategory_Full))
```

```{r clean-and-cluster}
# merge module data and clean up categories
combined_mod_annotated <- combined_mod_df %>%
  mutate(Clean_ID = gsub("path:", "", ID)) %>%
  left_join(mod_meta, by = c("Clean_ID" = "ID")) %>%
  mutate(
    Subcategory = replace_na(Subcategory, "Other Modules"),
  # before adding the cluster heirarchical sort, I reformatting the carb metab to come together alphabetically
#    Subcategory = case_when(
#      Clean_ID == "M00550" ~ "Carbohydrate metabolism, other",
#      Clean_ID == "M00009" ~ "Carbohydrate metabolism, central",
#      Clean_ID == "M00010" ~ "Carbohydrate metabolism, central",
#      TRUE ~ Subcategory
#    )
  )

# heirarchical clustering
module_matrix <- combined_mod_annotated %>%
  select(Description, Analysis_Group, FoldEnrichment) %>%
  pivot_wider(names_from = Analysis_Group, values_from = FoldEnrichment, values_fill = 0) %>%
  column_to_rownames("Description")

dist_matrix <- dist(t(as.matrix(module_matrix)))
hc <- hclust(dist_matrix, method = "complete")
ordered_groups <- hc$labels[hc$order]

combined_mod_annotated$Analysis_Group <- factor(
  combined_mod_annotated$Analysis_Group, 
  levels = ordered_groups
)
```

Hash out to organize facets alphabetically

```{r cluster-facets}
module_category_matrix <- combined_mod_annotated %>%
  group_by(Subcategory, Analysis_Group) %>%
  summarise(FoldEnrichment = mean(FoldEnrichment, na.rm = TRUE), .groups = "drop") %>% #change mean to max etc
  pivot_wider(names_from = Analysis_Group,values_from = FoldEnrichment,values_fill = 0) %>%
  column_to_rownames("Subcategory")

module_category_dist <- dist(as.matrix(module_category_matrix))

module_category_hc <- hclust(module_category_dist,method = "complete")

ordered_module_categories <- module_category_hc$labels[module_category_hc$order]

combined_mod_annotated <- combined_mod_annotated %>%
  mutate(Subcategory = factor(Subcategory,levels = ordered_module_categories))
# this is not currently reordering within each facet, fix this later, idk package name, maybe tidytext?
# for some reason, it seems running this block twice fixes it
combined_mod_annotated <- combined_mod_annotated %>%
  group_by(Subcategory) %>%
  mutate(
    Description = factor(Description, levels = rev(sort(unique(Description))))
  ) %>%
  ungroup()
```

## graphing

```{r graph-module-padj-enrich}
ggplot(combined_mod_annotated, aes(x = Analysis_Group, y = Description)) +
  geom_point(aes(size = -log10(qvalue), colour = FoldEnrichment)) +
  scale_colour_viridis_c(option = colour_scale, direction = scale_direction, name = "Fold Enrichment") +
  scale_size_continuous(name = "-log10(qvalue)", breaks = scales::breaks_pretty(n = 4)) +
  
  facet_grid(
    rows = vars(Subcategory), 
    scales = "free_y", 
    space = "free_y",
    labeller = labeller(Subcategory = label_wrap_gen(width = 22))
  ) +
  
  theme_bw(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, face = "bold"),
    strip.text.y = element_text(face = "bold", size = 8, angle = 0, hjust = 0),
    strip.background = element_rect(fill = "grey85"),
    panel.grid.major.x = element_line(colour = xgrid_colour, linetype = xgrid_linetype),
    panel.grid.major.y = element_line(colour = ygrid_colour, linetype = ygrid_linetype)
  ) +
  labs(
    x = x_text,
    y = y_text_m,
    title = figure_title_m,
    caption = caption_text_m
  )

ggsave(file = file.path(save_dir, "kegg_modules_clustered_a.png"), 
       width = width_m, height = height_m, dpi = 300)
```


```{r graph-module-enrich-count}
ggplot(combined_mod_annotated, aes(x = Analysis_Group, y = Description)) +
  geom_point(aes(size = Count, colour = FoldEnrichment)) +
  scale_colour_viridis_c(option = colour_scale, direction = scale_direction, name = "Fold Enrichment") +
  scale_size_continuous(name = "Gene Count", breaks = scales::breaks_pretty(n = 4)) +
  
  facet_grid(
    rows = vars(Subcategory), 
    scales = "free_y", 
    space = "free_y",
    labeller = labeller(Subcategory = label_wrap_gen(width = 22))
  ) +
  
  theme_bw(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, face = "bold"),
    strip.text.y = element_text(face = "bold", size = 8, angle = 0, hjust = 0),
    strip.background = element_rect(fill = "grey85"),
    panel.grid.major.x = element_line(colour = xgrid_colour, linetype = xgrid_linetype),
    panel.grid.major.y = element_line(colour = ygrid_colour, linetype = ygrid_linetype)
  ) +
  labs(
    x = x_text,
    y = y_text_m,
    title = figure_title_m,
    caption = caption_text_m
  )

ggsave(file = file.path(save_dir,"kegg_modules_clustered_b.png"), 
       width = width_m, height = height_m,  dpi = 300)
```


# COG LETTER CATEGORIES

## ingestion

```{r}
# Read COG enrichment results
cog_folder_path <- file.path(tsv_dir, "COG_cat")

cog_files <- list.files(cog_folder_path, pattern = "\\.csv$", full.names = TRUE)

combined_cog_df <- bind_rows(lapply(cog_files, read.delim, stringsAsFactors = FALSE))


# Read COG indexing file
cog_index <- read.csv("/mnt/work-drive/databases/COG-KEGG-data/cog_letter_supercategories.csv",
  stringsAsFactors = FALSE) %>%
  rename(
    COG_Letter = letter,
    COG_Description = description,
    COG_Category = category
)


combined_cog_annotated <- combined_cog_df %>%
  left_join(cog_index,by = c("ID" = "COG_Letter")) %>%
  mutate(
    Description = COG_Description,
    Subcategory = COG_Category
  )

```

```{r}
# Hierarchical clustering of COG enrichment profiles
cog_matrix <- combined_cog_annotated %>%
  select(Description, Analysis_Group, FoldEnrichment) %>%
  pivot_wider(
    names_from = Analysis_Group,
    values_from = FoldEnrichment,
    values_fill = 0
  ) %>%
  column_to_rownames("Description")


cog_dist_matrix <- dist(t(as.matrix(cog_matrix)))

cog_hc <- hclust(cog_dist_matrix,method = "complete")

ordered_cog_groups <- cog_hc$labels[cog_hc$order]


combined_cog_annotated$Analysis_Group <- factor(
  combined_cog_annotated$Analysis_Group,
  levels = ordered_cog_groups
)

```
Hash out to organize facets alphabetically

```{r cluster-facets}
module_category_matrix <- combined_cog_annotated %>%
  group_by(Subcategory, Analysis_Group) %>%
  summarise(FoldEnrichment = mean(FoldEnrichment, na.rm = TRUE), .groups = "drop") %>% #change mean to max etc
  pivot_wider(names_from = Analysis_Group,values_from = FoldEnrichment,values_fill = 0) %>%
  column_to_rownames("Subcategory")

module_category_dist <- dist(as.matrix(module_category_matrix))

module_category_hc <- hclust(module_category_dist,method = "complete")

ordered_module_categories <- module_category_hc$labels[module_category_hc$order]

combined_mod_annotated <- combined_cog_annotated %>%
  mutate(Subcategory = factor(Subcategory,levels = ordered_module_categories))
# this is not currently reordering within each facet, fix this later, idk package name, maybe tidytext?
# for some reason, it seems running this block twice fixes it
combined_cog_annotated <- combined_cog_annotated %>%
  group_by(Subcategory) %>%
  mutate(
    Description = factor(Description, levels = rev(sort(unique(Description))))
  ) %>%
  ungroup()
```

## graphing

```{r graph-cog-let-padj-enrich}
ggplot(combined_cog_annotated, aes(x = Analysis_Group, y = Description)) +
  geom_point(aes(size = -log10(qvalue), colour = FoldEnrichment)) +
  scale_colour_viridis_c(option = colour_scale, direction = scale_direction, name = "Fold Enrichment") +
  scale_size_continuous( name = "-log10(qvalue)", breaks = scales::breaks_pretty(n = 4)) +
  
  facet_grid(
    rows = vars(Subcategory), 
    scales = "free_y", 
    space = "free_y",
    labeller = labeller(Subcategory = label_wrap_gen(width = 22))
  ) +
  
  theme_bw(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, face = "bold"),
    strip.text.y = element_text(face = "bold", size = 8, angle = 0, hjust = 0),
    strip.background = element_rect(fill = "grey85"),
    panel.grid.major.x = element_line(colour = xgrid_colour, linetype = xgrid_linetype),
    panel.grid.major.y = element_line(colour = ygrid_colour, linetype = ygrid_linetype)
  ) +
  labs(
    x = x_text,
    y = y_text_c,
    title = figure_title_c,
    caption = caption_text_c
  )

ggsave(file = file.path(save_dir, "cog_letters_clustered_a.png"), 
       width = width_c, height = height_c, dpi = 300)
```


```{r graph-cog-let-count-enrich}
ggplot(combined_cog_annotated, aes(x = Analysis_Group, y = Description)) +
  geom_point(aes(size = Count, colour = FoldEnrichment)) +
  scale_colour_viridis_c(option = colour_scale, direction = scale_direction, name = "FoldEnrichment") +
  scale_size_continuous(name = "Gene Count", breaks = scales::breaks_pretty(n = 4)) +
  
  facet_grid(
    rows = vars(Subcategory), 
    scales = "free_y", 
    space = "free_y",
    labeller = labeller(Subcategory = label_wrap_gen(width = 22))
  ) +
  
  theme_bw(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, face = "bold"),
    strip.text.y = element_text(face = "bold", size = 8, angle = 0, hjust = 0),
    strip.background = element_rect(fill = "grey85"),
    panel.grid.major.x = element_line(colour = xgrid_colour, linetype = xgrid_linetype),
    panel.grid.major.y = element_line(colour = ygrid_colour, linetype = ygrid_linetype)
  ) +
  labs(
    x = x_text,
    y = y_text_c,
    title = figure_title_c,
    caption = caption_text_c
  )

ggsave(file = file.path(save_dir, "cog_letters_clustered_b.png"), 
       width = width_c, height = height_c, dpi = 300)
```


```{r graph-cog-let-count-sig}
ggplot(combined_cog_annotated, aes(x = Analysis_Group, y = Description)) +
  geom_point(aes(size = Count, colour = -log10(qvalue))) +
  scale_colour_viridis_c(option = colour_scale, direction = scale_direction, name = "-log10(qvalue)") +
  scale_size_continuous(name = "Gene Count", breaks = scales::breaks_pretty(n = 4)) +
  
  facet_grid(
    rows = vars(Subcategory), 
    scales = "free_y", 
    space = "free_y",
    labeller = labeller(Subcategory = label_wrap_gen(width = 22))
  ) +
  
  theme_bw(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, face = "bold"),
    strip.text.y = element_text(face = "bold", size = 8, angle = 0, hjust = 0),
    strip.background = element_rect(fill = "grey85"),
    panel.grid.major.x = element_line(colour = xgrid_colour, linetype = xgrid_linetype),
    panel.grid.major.y = element_line(colour = ygrid_colour, linetype = ygrid_linetype)
  ) +
  labs(
    x = x_text,
    y = y_text_c,
    title = figure_title_c,
    caption = caption_text_c
  )

ggsave(file = file.path(save_dir, "cog_letters_clustered_c.png"), 
       width = width_c, height = height_c, dpi = 300)
```